---
title: "NMD analysis: current GENCODE-only scopes (2026-07-11)"
subtitle: "Canonical scopes for Figure 3 (D/E/F) + Figure 4 (A/B at all-3-ENST + coding-CDS n=130; C/D at ref-AUG-traceable re-intersected n=819/819). Branched from 05_final_report_mashr.Rmd; that report is now legacy."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
.nmd_root <- local({ d <- getwd(); while (!file.exists(file.path(d, "config", "paths.yml"))) {
  p <- dirname(d); if (identical(p, d)) stop("repo root not found"); d <- p }; d })
source(file.path(.nmd_root, "R", "load_config.R")); .P <- nmd_paths(.nmd_root)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

# CLAIM-VALUE EMISSION (Phase 2). Each §4 claim's number is written by the chunk that computes
# it, with the population in words, to $NMD_CLAIM_VALUES (default tmp/claim_values.tsv).
# tools/check_claim_values.py then compares against what the paper printed. The population is
# required and is captured HERE because the call site is the only place it is known --
# unstated populations are the dominant defect class in this manuscript, and this report is
# where the clearest instance lives: Panel C's legend names n=130 for a panel computed over
# 1,578 pairs per arm (claim 4.4.3), a factor of 12.
source(file.path(.nmd_root, "R", "claim_emit.R"))

# REPOINTED 2026-07-27 off .P$LEGACY onto the deposit-rebuilt tree, so §4's claims can be
# rebuilt from the deposit. This is the CANONICAL gencode-scope report (the 2026-06-15 file is
# legacy-bannered "DO NOT cite its numbers as canonical" and is deliberately left alone --
# repointing a report retained to reproduce a PRIOR state would defeat its purpose).
#
# NMD_ROOT was used for TWO DIFFERENT THINGS through one variable:
#   DATA (3 uses)     -- decide every number this report computes. Repointed below.
#   FIGURES (14 uses) -- include_graphics() embeds of pre-rendered PNGs, which cannot change
#                        a computed value. They stay on the legacy figure tree.
# Naming them apart is the point: one variable covering both is how a presentation path gets
# mistaken for evidence that the data root was repointed.
DM              <- .P$ISOPAIR_OUT
ISOPAIR_WRAPPER <- file.path(.nmd_root, "analysis/isopair/isopair_wrapper")  # helper source()
                                                                             # only; no data
FASTA_PATH      <- file.path(.P$SQANTI, "nmd_lungcells_corrected.fasta")

# Presentation only -- see above. Deliberately NOT called NMD_ROOT.
#
# REPOINTED 2026-07-29 (D22) off .P$LEGACY onto the repo itself. The old default made this
# report UNRENDERABLE by anyone without ../nmd, and not obviously so: knitr's include_graphics
# defaults to error = TRUE (verified, knitr 1.51; nothing here sets knitr.graphics.error), so a
# missing PNG ABORTS THE KNIT. The first embed is at L1034 and the chunk that prints n=130 is at
# L1058 -- a reader holding only the deposit died twenty-four lines short of the number, with a
# missing-file error that says nothing about why.
#
# The setup comment above is right that these embeds "cannot change a computed value", and that
# was read for months as meaning they could not matter. They cannot change the value; they can
# and did stop anyone else from ever seeing it. Causality and reachability are different
# questions.
FIG_ROOT <- Sys.getenv("NMD_FIG_ROOT", .nmd_root)

# EMBED FIGURES WITHOUT LETTING knitr REWRITE THE PATH (2026-08-03).
#
# The clean-room run died here at chunk 35/98 with "Cannot find the file(s):
# ../../../figures/...". The files were present and FIG_ROOT was correct -- knitr had REWRITTEN
# the absolute path to be relative to the DOCUMENT (three levels down, in
# analysis/isopair/isopair_wrapper/) and then resolved it against the WORKING DIRECTORY, which
# render() sets to the repo root. Correct path, two different bases.
#
# rel_path = FALSE tells knitr to leave the absolute path alone. Measured, not assumed: passing a
# knit-root-relative path instead renders "successfully" but emits
# "[WARNING] Could not fetch resource ..." and drops the image from the output -- a silent
# degradation, which is worse than the halt it replaces. This form renders with no warning and
# the image embedded.
inc_fig <- function(x) knitr::include_graphics(x, rel_path = FALSE)

# CREATE THE DIRECTORIES THIS REPORT WRITES INTO, rather than relying on them existing.
#
# W69's clean-tree run died here at chunk 29/98: fwrite() to
# figures/multipanel/figure3_isopair_and_ptc/data/panelB_sequence_similarity.tsv with
# "Unable to create new file for writing (it does not exist already)". In the authoring tree
# these directories exist because tracked TSVs sit in them -- figure3/data has 5 tracked files
# and SF33/data has 7 -- so the report has never had to create one. That is a directory
# existing BY ACCIDENT OF WHAT GIT TRACKS, and two things make it fragile: git does not track
# empty directories, so any of these becomes absent the moment its last file is removed; and
# whether figure-data TSVs ship at all is an open reading of D1, which admits "figures" and
# excludes "intermediate data" without saying which of the two these are.
#
# Same defect class as three of the four the cut rerun harness found: nothing implemented
# config/paths.yml's promise that OUT is "created on demand", and seven scripts wrote into
# data_mashr/analysis_cache/ that none of them created -- which cost 75 minutes of completed
# ORF scan. A script must create what it writes into.
for (.d in file.path(.nmd_root, c(
       "figures/multipanel/figure3_isopair_and_ptc/data",
       "figures/multipanel/figure4_ptcneg_and_model/data",
       "figures/supplemental/SF33_CdsAnd3UTR_GENCODE/data"))) {
  if (!dir.exists(.d)) dir.create(.d, recursive = TRUE, showWarnings = FALSE)
}

stopifnot(
  "ISOPAIR_WRAPPER does not resolve"                   = dir.exists(ISOPAIR_WRAPPER),
  "DM does not resolve"                                = dir.exists(DM),
  "FASTA_PATH does not exist (needed for §4b Kozak PWM)" = file.exists(FASTA_PATH)
)

library(Isopair)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

stopifnot(packageVersion("Isopair") >= "0.99.4")

source(file.path(ISOPAIR_WRAPPER, "analysis_functions.R"))

profiles_c2 <- readRDS(file.path(DM, "profiles_c2_allsamples.rds"))
profiles_c4 <- readRDS(file.path(DM, "profiles_c4_allsamples.rds"))
cds         <- as.data.table(readRDS(file.path(DM, "cds.rds")))
structures  <- as.data.table(readRDS(file.path(DM, "structures.rds")))
ref_atg     <- readRDS(file.path(DM, "analysis_cache/ref_atg_analysis.rds"))
utr5_refaug <- readRDS(file.path(DM, "analysis_cache/utr5_features_refaug.rds"))
utr5_all    <- readRDS(file.path(DM, "analysis_cache/utr5_features_all.rds"))
expr_mat    <- readRDS(file.path(DM, "expression_data.rds"))
nmd_class   <- readRDS(file.path(DM, "nmd_classification.rds"))
dmso_samples <- readRDS(file.path(DM, "dmso_samples.rds"))

need_in <- function(obj, cols, label) {
  missing <- setdiff(cols, colnames(obj))
  if (length(missing) > 0)
    stop(sprintf("%s missing columns: %s", label,
                  paste(missing, collapse = ", ")))
  invisible(TRUE)
}

# profiles_c{2,4} schema
need_in(profiles_c2,
        c("gene_id", "reference_isoform_id", "comparator_isoform_id"),
        "profiles_c2")
need_in(profiles_c4,
        c("gene_id", "reference_isoform_id", "comparator_isoform_id"),
        "profiles_c4")

# cds schema (used in §2a own-stop conversion, §2d CDS length, §4 TD2 ORF)
need_in(cds,
        c("isoform_id", "coding_status", "cds_start", "cds_stop",
          "strand", "orf_length"),
        "cds")

# structures schema (exon coords for genomicToTranscript + tx-length)
need_in(structures,
        c("isoform_id", "exon_starts", "exon_ends", "strand"),
        "structures")

# utr5 caches
need_in(utr5_refaug$isoform_features,
        c("isoform_id", "utr5_length", "longest_orf_nt"),
        "utr5_refaug$isoform_features")
need_in(utr5_all$isoform_features,
        c("isoform_id", "utr5_length", "longest_orf_nt"),
        "utr5_all$isoform_features")

# ref_atg slots + columns
stopifnot("ref_atg missing c2 slot" = "c2" %in% names(ref_atg))
stopifnot("ref_atg missing c4 slot" = "c4" %in% names(ref_atg))
need_in(as.data.frame(ref_atg$c2),
        c("comparator_isoform_id", "reference_isoform_id", "gene_id",
          "category", "comp_orf_length", "ref_orf_length",
          "comp_orf_length_from_ref_atg", "ref_atg_genomic",
          "ref_atg_exonic_in_comp", "original_ptc",
          "utr3_via_non_ptc_stop_nt"),
        "ref_atg$c2")
need_in(as.data.frame(ref_atg$c4),
        c("comparator_isoform_id", "reference_isoform_id", "gene_id",
          "category", "utr3_via_non_ptc_stop_nt"),
        "ref_atg$c4")

format_p <- function(p, digits = 2) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) formatC(p, format = "e", digits = digits)
  else formatC(signif(p, digits + 1), format = "g")
}
```

```{r mechanism-helpers, include=FALSE}
# Inline-replicated from figures/lib/mechanism_class.R (per v4 plan C2: kept
# inline so the Rmd is self-contained; the two copies are simple enough to
# compare in one diff if they ever drift).

MECHANISM_CLASS_LEVELS <- c(
  "NMD+/PTC+",
  "NMD+/PTC- ORF same",
  "NMD+/PTC- ORF extended",
  "NMD+/PTC- ORF truncated",
  "Ref AUG absent"
)

MECHANISM_CLASS_4_LEVELS <- c(
  "NMD+/PTC+",
  "NMD+/PTC- ORF match",
  "NMD+/PTC- ORF diff",
  "Ref AUG absent"
)

REF_AUG_PRESENT_CATEGORIES <- c(
  "effectively_ptc", "no_downstream_ejc", "truncated_no_ejc"
)

mechanism_class <- function(category, comp_orf_length, ref_orf_length) {
  stopifnot(length(category) == length(comp_orf_length),
            length(category) == length(ref_orf_length))
  orf_diff <- comp_orf_length - ref_orf_length
  cls <- rep(NA_character_, length(category))
  cls[is.na(category) | !category %in% REF_AUG_PRESENT_CATEGORIES] <-
    "Ref AUG absent"
  cls[!is.na(category) & category == "effectively_ptc"] <- "NMD+/PTC+"
  ndj_mask <- !is.na(category) & category == "no_downstream_ejc"
  trc_mask <- !is.na(category) & category == "truncated_no_ejc"
  cls[ndj_mask & !is.na(orf_diff) & orf_diff == 0] <- "NMD+/PTC- ORF same"
  cls[ndj_mask & !is.na(orf_diff) & orf_diff > 0]  <- "NMD+/PTC- ORF extended"
  cls[trc_mask & !is.na(orf_diff) & orf_diff < 0]  <- "NMD+/PTC- ORF truncated"
  # ndj/trc with NA orf_diff: route to PTC- (mech4 merges to NMD+/PTC-).
  # Synced with figures/lib/mechanism_class.R.
  cls[ndj_mask & is.na(orf_diff)] <- "NMD+/PTC- ORF same"
  cls[trc_mask & is.na(orf_diff)] <- "NMD+/PTC- ORF truncated"
  cls[is.na(cls)] <- "Ref AUG absent"
  factor(cls, levels = MECHANISM_CLASS_LEVELS)
}

mechanism_class_4 <- function(cls) {
  cls_char <- as.character(cls)
  out <- ifelse(cls_char %in% c("NMD+/PTC- ORF extended",
                                "NMD+/PTC- ORF truncated"),
                "NMD+/PTC- ORF diff",
                ifelse(cls_char == "NMD+/PTC- ORF same",
                       "NMD+/PTC- ORF match",
                       cls_char))
  factor(out, levels = MECHANISM_CLASS_4_LEVELS)
}

ref_source <- function(reference_isoform_id) {
  out <- ifelse(grepl("^ENST", reference_isoform_id), "GENCODE", "TD2")
  factor(out, levels = c("GENCODE", "TD2"))
}

TES_CONCORDANT_THRESHOLD_NT <- 50L
tes_concordant <- function(comp_tx_len, ref_tx_len,
                           threshold = TES_CONCORDANT_THRESHOLD_NT) {
  abs(comp_tx_len - ref_tx_len) <= threshold
}
```

# Overview {.unnumbered}

This report supersedes `05_final_report_mashr.Rmd` (now legacy-bannered) for the
two canonical scopes used in the manuscript's Figure 3 and Figure 4:

1. **all-3-ENST + coding-CDS (n=130)** — both NMD+ and Control comparators
   have ENST reference isoforms with a complete coding CDS. Drives Figure 3
   D/E/F and Figure 4 Panels A/B + the `CDSand3UTR_GENCODEonly` supplement.
2. **ref-AUG-traceable, re-intersected on `(gene_id, reference_isoform_id)`
   after category filter (n=819/819)** — drives Figure 4 Panels C/D and
   the `TD2BiasEvidence` supplement (`n=819` broad, `n=348` occult-PTC).

Analyses kept in the legacy Rmd (per v4 plan): per-cell-type reclassification,
two-pass TD2→ref-AUG handoff, the "85% combined" framework, the n=1,904/22,335
GENCODE-only 3'UTR outside-set, and all ORF / model analyses
(`06_orf_analysis_mashr.Rmd`).

# §1 — Isopair construction and pair sets

The Stage-2 gene-matched pair cohort (`pop_BC`) is the foundation of every
analysis in this report. Construction proceeds in three stages, replicated
inline so the cohort can be regenerated without re-running upstream scripts:

1. **Upstream isoform calling and filtering** (`01_prepare_data_mashr.R`) —
   the long-read isocall count matrix (`nmd_isocall.count_matrix.txt`,
   `r format(645273L, big.mark=",")` isoforms × 38 samples; constant defined
   at the head of `01_prepare_data_mashr.R`) is reduced by removing the 2 DO
   outlier samples, restricting to the four manuscript cell types, then applying
   TMM normalization, the condition-stratified 5% isoform filter, and
   `edgeR::filterByExpr`. Output: `expression_data.rds` with
   `r format(nrow(expr_mat), big.mark=",")` filtered isoforms ×
   `r ncol(expr_mat)` samples spanning the four manuscript cell types
   (AT, DD, FB, MV): AT 6 (3 DMSO + 3 SMG1i), DD 8 (4 + 4), FB 6 (3 + 3),
   MV 6 (3 + 3). The cell-type restriction is applied at the source, so the
   isoform universe, normalization, reference selection, reference-share floor,
   and pair construction are all on the four cell types.

2. **Pair construction** (`02_build_profiles_mashr.R`) — `profiles_c2`
   pairs the top NMD-sensitive isoform per gene against its dominant non-NMD
   reference; `profiles_c4` pairs the top-two non-NMD isoforms per gene.
   Cache: `profiles_c{2,4}_allsamples.rds` (c2 = `r format(nrow(profiles_c2), big.mark=",")`
   pre-match rows; c4 = `r format(nrow(profiles_c4), big.mark=",")`).

3. **Gene-matching** (inline below; replicates `03_nmd_analysis_mashr.Rmd`
   Stage-2 filter) — restrict both arms to the shared `(gene_id,
   reference_isoform_id)` set so every gene contributes both a NMD pair and
   a Control pair under the same reference.

```{r sec1-pop-bc, include=FALSE}
pop_bc_shared <- dplyr::inner_join(
  unique(profiles_c2[, c("gene_id", "reference_isoform_id")]),
  unique(profiles_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id")
)
pop_bc_c2 <- dplyr::semi_join(profiles_c2, pop_bc_shared,
                              by = c("gene_id", "reference_isoform_id"))
pop_bc_c4 <- dplyr::semi_join(profiles_c4, pop_bc_shared,
                              by = c("gene_id", "reference_isoform_id"))

ref_ids_pop  <- unique(pop_bc_c2$reference_isoform_id)
nmd_ids_pop  <- unique(pop_bc_c2$comparator_isoform_id)
ctrl_ids_pop <- unique(pop_bc_c4$comparator_isoform_id)

# Claim 4.1.2 — the analysis population. The paper's four selection criteria (>= 3 expressed
# isoforms, 1 NMD susceptible, 2 non-NMD, dominant non-NMD >= 25% of gene expression) are
# applied UPSTREAM in profile construction, not here, so this chunk can only speak to the
# resulting count; the criteria themselves are not emitted and will show as unclaimed.
claim_emit("4.1.2", "genes in the analysis population", nrow(pop_bc_shared),
           published = 1578, n = nrow(pop_bc_shared),
           population = paste("genes with a shared (gene, reference isoform) pair in BOTH the",
                              "NMD-comparator and Control-comparator profile sets, so every",
                              "gene contributes exactly one NMD and one Control pair; no CDS",
                              "or GENCODE-annotation constraint applied"))
```

`pop_BC` = `r format(nrow(pop_bc_shared), big.mark=",")` shared `(gene,
reference)` pairs; `r format(nrow(pop_bc_c2), big.mark=",")` NMD-comp pairs
and `r format(nrow(pop_bc_c4), big.mark=",")` Control-comp pairs — i.e.,
every gene contributes exactly one of each.

### Cohort cascade — pop_BC through Section A / Section C / TD2 branches

<!-- ── hoisted 2026-07-26 ─────────────────────────────────────────────────────────
     These four chunks compute the Section A / Section C / PTC / attribution cohorts.
     They used to sit AFTER the flowchart, which is why the flowchart hardcoded its
     numbers: chunks evaluate in order, so it could not see them. All four are
     include=FALSE and verified to use nothing from §1a-§1c, so hoisting them changes
     no output and lets the diagram label itself from the same objects the tables use.
     ──────────────────────────────────────────────────────────────────────────── -->

```{r sec2-scope, include=FALSE}
profiles_c2_dt <- as.data.table(profiles_c2)
profiles_c4_dt <- as.data.table(profiles_c4)
cds_dt         <- cds

is_enst  <- function(x) grepl("^ENST", x)
make_key <- function(d) paste(d$gene_id, d$reference_isoform_id, sep = "::")

# Step 1 — pop_BC (gene-matched on shared (gene, ref))
pop_bc_keys <- intersect(unique(make_key(profiles_c2_dt)),
                          unique(make_key(profiles_c4_dt)))
pop_bc_c2   <- profiles_c2_dt[make_key(profiles_c2_dt) %in% pop_bc_keys]
pop_bc_c4   <- profiles_c4_dt[make_key(profiles_c4_dt) %in% pop_bc_keys]

# Step 2 — all-ENST on both reference AND comparator
pop_bc_c2_E <- pop_bc_c2[is_enst(reference_isoform_id) &
                           is_enst(comparator_isoform_id)]
pop_bc_c4_E <- pop_bc_c4[is_enst(reference_isoform_id) &
                           is_enst(comparator_isoform_id)]
joint1      <- intersect(make_key(pop_bc_c2_E), make_key(pop_bc_c4_E))
pop_bc_c2_E <- pop_bc_c2_E[make_key(pop_bc_c2_E) %in% joint1]
pop_bc_c4_E <- pop_bc_c4_E[make_key(pop_bc_c4_E) %in% joint1]

# Step 3 — coding-everywhere (curated CDS on both sides)
coding_ids <- cds_dt[coding_status == "coding", isoform_id]
have_cds   <- function(d) d$reference_isoform_id %in% coding_ids &
                          d$comparator_isoform_id %in% coding_ids
pop_bc_c2_E2 <- pop_bc_c2_E[have_cds(pop_bc_c2_E)]
pop_bc_c4_E2 <- pop_bc_c4_E[have_cds(pop_bc_c4_E)]

# Step 4 — re-intersect on (gene, ref) after the coding filter, so both arms
#          share the same reference set
joint2 <- intersect(make_key(pop_bc_c2_E2), make_key(pop_bc_c4_E2))
gencode_all3_c2 <- pop_bc_c2_E2[make_key(pop_bc_c2_E2) %in% joint2]
gencode_all3_c4 <- pop_bc_c4_E2[make_key(pop_bc_c4_E2) %in% joint2]

sec2_cascade <- data.frame(
  Step = c("pop_BC (gene-matched)",
           "+ all-ENST (ref & comp)",
           "  re-intersect on (gene, ref)",
           "+ coding-CDS everywhere",
           "  re-intersect on (gene, ref)"),
  c2_n = c(nrow(pop_bc_c2),
           length(unique(c(make_key(pop_bc_c2_E), make_key(pop_bc_c4_E)))),
           nrow(pop_bc_c2_E),
           nrow(pop_bc_c2_E2),
           nrow(gencode_all3_c2)),
  c4_n = c(nrow(pop_bc_c4),
           NA,
           nrow(pop_bc_c4_E),
           nrow(pop_bc_c4_E2),
           nrow(gencode_all3_c4))
)

# Claim 4.1.3 — the GENCODE reference set ("all three analyzed isoforms present in the GENCODE
# reference"). This is the population n=130 refers to, and it is the denominator claims 4.4.7
# and 4.5.2 report their PTC rates over.
SEC_A_POPULATION <- paste("genes whose reference, NMD comparator and Control comparator are ALL",
                          "GENCODE (ENST) isoforms with a curated coding CDS, re-intersected on",
                          "(gene, reference) after the coding filter so the NMD and Control arms",
                          "share one gene-matched universe; a subset of the gene-matched pair cohort (one reference, one NMD comparator and one Control comparator per gene)")
claim_emit("4.1.3", "genes in the GENCODE reference set", nrow(gencode_all3_c2),
           published = 132, n = nrow(gencode_all3_c2), population = SEC_A_POPULATION)
```
```{r sec2a-ptc, include=FALSE}
coding_cds <- cds[coding_status == "coding"]
struct_idx <- match(coding_cds$isoform_id, structures$isoform_id)

own_stop_tx_lookup <- (function() {
  stop_tx <- rep(NA_integer_, nrow(coding_cds))
  for (i in seq_len(nrow(coding_cds))) {
    si <- struct_idx[i]
    if (is.na(si)) next
    strand <- coding_cds$strand[i]
    if (is.na(strand) || strand == "") next
    stop_g <- if (strand == "+") coding_cds$cds_stop[i] else coding_cds$cds_start[i]
    stop_tx[i] <- Isopair::genomicToTranscript(
      stop_g, structures$exon_starts[[si]], structures$exon_ends[[si]], strand)
  }
  data.table(isoform_id = coding_cds$isoform_id,
             own_stop_tx_pos = stop_tx,
             strand = coding_cds$strand)
})()

junctions_for <- function(iso_id) {
  si <- match(iso_id, structures$isoform_id)
  if (is.na(si)) return(integer(0))
  exon_starts <- structures$exon_starts[[si]]
  exon_ends   <- structures$exon_ends[[si]]
  strand      <- structures$strand[si]
  if (is.na(strand) || strand == "") return(integer(0))
  exon_lens <- exon_ends - exon_starts + 1L
  if (strand == "-") exon_lens <- rev(exon_lens)
  if (length(exon_lens) <= 1L) return(integer(0))
  cumsum(exon_lens[-length(exon_lens)])
}

augment_own_stop <- function(pop) {
  merge(pop[, .(gene_id, reference_isoform_id, comparator_isoform_id)],
        own_stop_tx_lookup,
        by.x = "comparator_isoform_id", by.y = "isoform_id",
        all.x = TRUE)
}
compute_ptc_info <- function(d) {
  d[, last_ejc_tx_pos := vapply(comparator_isoform_id,
                                 function(iso) {
                                   junc <- junctions_for(iso)
                                   if (length(junc) == 0L) NA_integer_ else max(junc)
                                 }, integer(1))]
  d[, distance := last_ejc_tx_pos - own_stop_tx_pos]
  d[, has_ptc := !is.na(distance) & distance > 50L]
  d
}

trace_c2 <- compute_ptc_info(augment_own_stop(gencode_all3_c2))
trace_c4 <- compute_ptc_info(augment_own_stop(gencode_all3_c4))

n_ptc_nmd     <- sum(trace_c2$has_ptc, na.rm = TRUE)
n_ptc_ctrl    <- sum(trace_c4$has_ptc, na.rm = TRUE)
n_avail_nmd   <- sum(!is.na(trace_c2$distance))
n_avail_ctrl  <- sum(!is.na(trace_c4$distance))
ptc_rate_nmd  <- 100 * n_ptc_nmd  / n_avail_nmd
ptc_rate_ctrl <- 100 * n_ptc_ctrl / n_avail_ctrl
fold_enrich   <- ptc_rate_nmd / ptc_rate_ctrl

ptc_fisher <- fisher.test(matrix(c(
  n_ptc_nmd,            n_avail_nmd  - n_ptc_nmd,
  n_ptc_ctrl,           n_avail_ctrl - n_ptc_ctrl
), nrow = 2, byrow = TRUE))

pop_ptc_c2 <- merge(
  gencode_all3_c2,
  trace_c2[has_ptc == TRUE,
           .(comparator_isoform_id, own_stop_tx_pos, last_ejc_tx_pos, distance)],
  by = "comparator_isoform_id"
)

# ── Claims 4.4.7 and 4.5.2 — the PTC rate in the GENCODE reference set ────────────────
# The denominator here is pairs with a COMPUTABLE stop-to-last-EJC distance, which is not
# automatically the whole Section A cohort: a single-exon comparator has no junction, so
# junctions_for() returns integer(0) and the distance is NA. Stating that is the point of the
# population field -- the published "48/130" does not say which 130.
PTC_POP_NMD <- paste("NMD-comparator pairs in the GENCODE reference set (all three isoforms",
                     "ENST with a curated coding CDS) that have a computable stop-to-last-EJC",
                     "distance; comparators with no exon junction are excluded as NA")
PTC_POP_CTRL <- sub("NMD-comparator", "Control-comparator", PTC_POP_NMD)

claim_emit("4.4.7", "PTC rate in NMD comparators (%)", ptc_rate_nmd,
           published = 36.4, n = n_avail_nmd, population = PTC_POP_NMD)
claim_emit("4.4.7", "NMD comparators with a PTC (count)", n_ptc_nmd,
           published = 48, n = n_avail_nmd, population = PTC_POP_NMD)
claim_emit("4.4.7", "NMD comparators assessable for a PTC (denominator)", n_avail_nmd,
           published = 132, n = n_avail_nmd, population = PTC_POP_NMD)
claim_emit("4.4.7", "PTC rate in Control comparators (%)", ptc_rate_ctrl,
           published = 1.5, n = n_avail_ctrl, population = PTC_POP_CTRL)
claim_emit("4.4.7", "Control comparators with a PTC (count)", n_ptc_ctrl,
           published = 2, n = n_avail_ctrl, population = PTC_POP_CTRL)

claim_emit("4.5.2", "PTC fold enrichment, NMD over Control", fold_enrich,
           published = 24, n = n_avail_nmd + n_avail_ctrl,
           population = paste("ratio of the two PTC rates above, over the GENCODE reference",
                              "set; both arms gene-matched, one NMD and one Control comparator",
                              "per gene"))
# The paper states this as an INEQUALITY ("p < 10^-13"), not as a value. Emitting it against a
# point value would score a disagreement for a claim that is satisfied by any smaller p, so the
# computed p is recorded as a SUPPORTING number with the paper's threshold named in the
# quantity, and the inequality is checked by reading it.
claim_emit("4.5.2", "PTC rate in NMD comparators (%)", ptc_rate_nmd,
           published = 36.4, n = n_avail_nmd, population = PTC_POP_NMD)
claim_emit("4.5.2", "PTC rate in Control comparators (%)", ptc_rate_ctrl,
           published = 1.5, n = n_avail_ctrl, population = PTC_POP_CTRL)
claim_emit("4.5.2", "Fisher p for PTC enrichment (paper states p < 1e-13)",
           ptc_fisher$p.value, n = n_avail_nmd + n_avail_ctrl,
           population = paste("Fisher exact test on PTC-positive vs PTC-negative counts in the",
                              "two arms of the GENCODE reference set"))
```
```{r sec2b-attribution, include=FALSE}
fc_c2 <- readRDS(file.path(DM, "analysis_cache/fc_c2_allsamples.rds"))
fw_c2 <- readRDS(file.path(DM, "analysis_cache/fw_c2_allsamples.rds"))

own_stop_genomic_lookup <- setNames(
  ifelse(coding_cds$strand == "+", coding_cds$cds_stop, coding_cds$cds_start),
  coding_cds$isoform_id)
own_atg_genomic_lookup <- setNames(
  ifelse(coding_cds$strand == "+", coding_cds$cds_start, coding_cds$cds_stop),
  coding_cds$isoform_id)
strand_lookup <- setNames(coding_cds$strand, coding_cds$isoform_id)

ep_ids <- pop_ptc_c2$comparator_isoform_id
fc_ps <- as.data.table(fc_c2$pair_summary)
fc_ep <- fc_ps[comparator_isoform_id %in% ep_ids]
frameshift_cats <- c("same_start_frameshift",
                     "diff_start_diff_frame_with_frameshift")
fc_ep[, is_frameshift_cat := frame_category %in% frameshift_cats]

ptc_genomic_pos <- own_stop_genomic_lookup[fc_ep$comparator_isoform_id]
atg_genomic_pos <- own_atg_genomic_lookup[fc_ep$comparator_isoform_id]
strand_vec      <- strand_lookup[fc_ep$comparator_isoform_id]
names(ptc_genomic_pos) <- fc_ep$comparator_isoform_id
names(atg_genomic_pos) <- fc_ep$comparator_isoform_id
names(strand_vec)      <- fc_ep$comparator_isoform_id
is_fs_lookup <- setNames(fc_ep$is_frameshift_cat, fc_ep$comparator_isoform_id)

attr_b <- as.data.table(attribute_ptc_events(
  pairs           = fc_ep[, .(comparator_isoform_id, reference_isoform_id)],
  fw_events       = fw_c2$events,
  profiles        = as.data.frame(pop_ptc_c2),
  ptc_genomic_pos = ptc_genomic_pos,
  atg_genomic_pos = atg_genomic_pos,
  strand_vec      = strand_vec,
  is_frameshift_vec = is_fs_lookup
))
n_direct_b     <- sum(attr_b$attribution == "direct")
n_unresolved_b <- sum(attr_b$attribution == "unresolved")

# Same-stop 3'UTR splice handling
fc_ep[, strand_s     := strand_lookup[comparator_isoform_id]]
fc_ep[, comp_stop_s  := own_stop_genomic_lookup[comparator_isoform_id]]
fc_ep[, ref_stop_s   := own_stop_genomic_lookup[reference_isoform_id]]
fc_ep[, same_stop_s  := comp_stop_s == ref_stop_s]
same_stop_b <- fc_ep[same_stop_s == TRUE]

ss_events_b <- data.table()
if (nrow(same_stop_b) > 0) {
  ss_stop_vec   <- setNames(same_stop_b$comp_stop_s, same_stop_b$comparator_isoform_id)
  ss_strand_vec <- setNames(same_stop_b$strand_s,    same_stop_b$comparator_isoform_id)
  ss_events_b <- as.data.table(attribute_3utr_splice(
    pairs            = same_stop_b[, .(comparator_isoform_id, reference_isoform_id)],
    profiles         = as.data.frame(pop_ptc_c2),
    stop_genomic_pos = ss_stop_vec,
    strand_vec       = ss_strand_vec
  ))[attribution == "direct"]
}
n_same_stop_direct <- nrow(ss_events_b)

# Combine: same-stop precedence over diff-stop
same_stop_ids <- if (nrow(ss_events_b) > 0) ss_events_b$comparator_isoform_id else character(0)
diffstop_b    <- attr_b[attribution == "direct" &
                        !comparator_isoform_id %in% same_stop_ids]
all_attr_new  <- rbind(diffstop_b, ss_events_b, fill = TRUE)
all_attr_new  <- all_attr_new[attribution == "direct"]
all_attr_new$event_short <- shorten_event_labels(all_attr_new$ptc_causing_event,
                                                  collapse_rare = 3)
n_unique_direct <- nrow(all_attr_new)

# Mechanism breakdown (38 / 23 / 8 target)
all_attr_new[, mechanism := factor(mechanism,
                                   levels = c("Frameshift", "In-frame stop", "3'UTR splice"))]
mech_tab <- as.data.frame(table(all_attr_new$mechanism, useNA = "ifany"))
names(mech_tab) <- c("Mechanism", "n")

# ── Claims 4.4.10 and 4.6.1 — the mechanism split of the attributed PTC events ─────────
# EMITTED HERE, AT THE POINT OF COMPUTATION, not in `sec2b-mech-table` where the ledger
# attributes them: that chunk only calls knitr::kable on this object. The contract requires the
# call site because the population is known here and nowhere else.
MECH_POP <- paste("splicing events attributed as the direct cause of a PTC, one per PTC-positive",
                  "pair, in the GENCODE reference set; same-stop 3'UTR-splice attributions take",
                  "precedence over different-stop ones, and unresolved attributions are excluded")
mech_n <- function(m) {
  i <- match(m, as.character(mech_tab$Mechanism))
  if (is.na(i)) 0L else as.integer(mech_tab$n[i])
}
n_fs <- mech_n("Frameshift"); n_ifs <- mech_n("In-frame stop"); n_u3 <- mech_n("3'UTR splice")

claim_emit("4.4.10", "attributed PTC-causing events", n_unique_direct,
           published = 48, n = n_unique_direct, population = MECH_POP)
claim_emit("4.4.10", "frameshift events (count)", n_fs,
           published = 26, n = n_unique_direct, population = MECH_POP)
claim_emit("4.4.10", "in-frame stop events (count)", n_ifs,
           published = 18, n = n_unique_direct, population = MECH_POP)
claim_emit("4.4.10", "3'UTR splice events (count)", n_u3,
           published = 4, n = n_unique_direct, population = MECH_POP)

# The three percentages are asserted by BOTH 4.4.10 (the figure legend) and 4.6.1 (the Results
# sentence). Two claims, so two emissions each -- the S4 key is (claim_id, quantity), and one
# writer per number is preserved because both come from this single computation.
#
# WRITTEN OUT RATHER THAN LOOPED OVER c("4.4.10", "4.6.1"): with the id in a loop variable the
# claim id is not a literal at the call site, and tools/check_claim_emit_sites.py -- correctly --
# scored 4.6.1 as uninstrumented, because a call site no static check can find cannot be
# verified before the run.
pct_fs  <- 100 * n_fs  / n_unique_direct
pct_ifs <- 100 * n_ifs / n_unique_direct
pct_u3  <- 100 * n_u3  / n_unique_direct
claim_emit("4.4.10", "frameshift share of attributed events (%)", pct_fs,
           published = 54, n = n_unique_direct, population = MECH_POP)
claim_emit("4.4.10", "in-frame stop share of attributed events (%)", pct_ifs,
           published = 38, n = n_unique_direct, population = MECH_POP)
claim_emit("4.4.10", "3'UTR splice share of attributed events (%)", pct_u3,
           published = 8, n = n_unique_direct, population = MECH_POP)
claim_emit("4.6.1", "frameshift share of attributed events (%)", pct_fs,
           published = 54, n = n_unique_direct, population = MECH_POP)
claim_emit("4.6.1", "in-frame stop share of attributed events (%)", pct_ifs,
           published = 38, n = n_unique_direct, population = MECH_POP)
claim_emit("4.6.1", "3'UTR splice share of attributed events (%)", pct_u3,
           published = 8, n = n_unique_direct, population = MECH_POP)
```
```{r sec3-scope, include=FALSE}
ref_atg_c2 <- as.data.table(ref_atg$c2)
ref_atg_c4 <- as.data.table(ref_atg$c4)

# Step 1 — ENST-reference subset of pop_BC (no constraint on comparator)
sec_C_c2_base <- pop_bc_c2[is_enst(reference_isoform_id)]
sec_C_c4_base <- pop_bc_c4[is_enst(reference_isoform_id)]

# Step 2 — restrict ref_atg tables to the ENST-reference pop_BC comparators
c2_full <- ref_atg_c2[comparator_isoform_id %in%
                        sec_C_c2_base$comparator_isoform_id]
c4_full <- ref_atg_c4[comparator_isoform_id %in%
                        sec_C_c4_base$comparator_isoform_id]

# Step 3 — ref-AUG-traceable category filter (per side)
TR    <- REF_AUG_PRESENT_CATEGORIES   # from mechanism-helpers chunk
c2_tr <- c2_full[category %in% TR]
c4_tr <- c4_full[category %in% TR]

# Step 4 — re-intersect on (gene, reference) AFTER the category filter so
#          NMD and Control panels share the same 1:1 gene-matched universe
shared_keys <- merge(
  unique(c2_tr[, .(gene_id, reference_isoform_id)]),
  unique(c4_tr[, .(gene_id, reference_isoform_id)]),
  by = c("gene_id", "reference_isoform_id")
)
refaug_matched_c2 <- merge(c2_tr, shared_keys,
                            by = c("gene_id", "reference_isoform_id"))
refaug_matched_c4 <- merge(c4_tr, shared_keys,
                            by = c("gene_id", "reference_isoform_id"))

# Classification: 5-group → 4-group → 3-group merged
refaug_matched_c2[, mech5 := mechanism_class(category, comp_orf_length, ref_orf_length)]
refaug_matched_c2[, mech4 := mechanism_class_4(mech5)]
refaug_matched_c2[, group3 := data.table::fcase(
  mech4 == "NMD+/PTC+",                                       "NMD+/PTC+",
  mech4 %in% c("NMD+/PTC- ORF match", "NMD+/PTC- ORF diff"),  "NMD+/PTC-",
  default = "Ref AUG absent"
)]

sec3_cascade <- data.frame(
  Step = c("pop_BC (gene-matched)",
           "+ ENST-reference (no comp constraint)",
           "+ ref-AUG-traceable category (per side)",
           "  re-intersect on (gene, ref)"),
  c2_n = c(nrow(pop_bc_c2),
           nrow(sec_C_c2_base),
           nrow(c2_tr),
           nrow(refaug_matched_c2)),
  c4_n = c(nrow(pop_bc_c4),
           nrow(sec_C_c4_base),
           nrow(c4_tr),
           nrow(refaug_matched_c4))
)

sec3_class <- as.data.frame(table(refaug_matched_c2$group3,
                                   useNA = "ifany"))
names(sec3_class) <- c("group", "n")

# Claim 4.1.4 — the reference AUG-traceable set.
SEC_C_POPULATION <- paste("NMD-comparator pairs whose REFERENCE isoform is GENCODE (ENST) and",
                          "whose comparator falls in a ref-AUG-traceable category, so the CDS",
                          "can be projected onto novel comparators by tracing the reference AUG;",
                          "re-intersected on (gene, reference) after the category filter so both",
                          "arms share one gene-matched universe. No requirement that the",
                          "comparator itself be in GENCODE, which is what makes it ~6x larger",
                          "than the GENCODE reference set")
claim_emit("4.1.4", "NMD pairs in the reference AUG-traceable set", nrow(refaug_matched_c2),
           published = 833, n = nrow(refaug_matched_c2), population = SEC_C_POPULATION)
```

```{r sec1-flowchart-data, include=FALSE}
# Pull canonical counts for the flowchart (NMD classification + pair
# construction stages) from the loaded caches. The §2 / §3 scope counts and
# the TD2 broad / occult-PTC scope counts are taken from the canonical Pass-7
# verifier manifest at reproducibility/verify_pass7_new_rmd.R — these are
# guarded against drift by both verifiers.
n_filtered    <- nrow(expr_mat)                       # 123,790 isoforms
n_samples_4ct <- 26L                                  # 4-CT manuscript scope
n_dmso_4ct    <- 13L                                  # AT 3 + DD 4 + FB 3 + MV 3
n_smg1i_4ct   <- 13L
n_nmd_as      <- length(nmd_class[["all_samples"]]$nmd)
n_nonnmd_as   <- length(nmd_class[["all_samples"]]$non_nmd)
n_c2_before   <- nrow(profiles_c2)
n_c4_before   <- nrow(profiles_c4)
fmt <- function(x) format(x, big.mark = ",")

# --- Flowchart cohort labels, 2026-07-26 ----------------------------------------------
# These were HARDCODED into the diagram's dot spec (1,548 / 130 / 819 / 48 / 36.9% / OR
# 37.06 / 756 ...). They could not be computed here because the chunks that derive them ran
# AFTER the diagram; those four chunks are now hoisted above it, so the diagram labels
# itself from the same objects the tables use and cannot drift from them again.
#
# The rebuild is what made the drift visible: the diagram still said 130/130 while the text
# said 132/132, in one document.
fd_popbc_c2  <- nrow(pop_bc_c2);        fd_popbc_c4  <- nrow(pop_bc_c4)
fd_enst_c2   <- nrow(pop_bc_c2_E);      fd_enst_c4   <- nrow(pop_bc_c4_E)
fd_all3_c2   <- nrow(gencode_all3_c2);  fd_all3_c4   <- nrow(gencode_all3_c4)
fd_ptc_n     <- n_ptc_nmd
fd_ptc_pct   <- ptc_rate_nmd
fd_nptc_n    <- n_avail_nmd - n_ptc_nmd
fd_or        <- unname(ptc_fisher$estimate)
fd_p         <- ptc_fisher$p.value
fd_fold      <- fold_enrich
fd_scC_base_c2 <- nrow(sec_C_c2_base);  fd_scC_base_c4 <- nrow(sec_C_c4_base)
fd_scC_tr_c2   <- nrow(c2_tr);          fd_scC_tr_c4   <- nrow(c4_tr)
fd_scC_c2      <- nrow(refaug_matched_c2); fd_scC_c4   <- nrow(refaug_matched_c4)
.g <- function(g) { v <- sec3_class$n[sec3_class$group == g]; if (length(v)) v[1] else NA_integer_ }
fd_scC_ptc_n   <- .g("NMD+/PTC+")
fd_scC_nptc_n  <- .g("NMD+/PTC-")
fd_scC_ptc_pct <- 100 * fd_scC_ptc_n / fd_scC_c2
```

```{r sec1-flowchart, fig.width=11, fig.height=14, echo=FALSE}
dot_spec <- sprintf('
digraph cohort_cascade {
  graph [rankdir=TB, fontname="Helvetica", nodesep=0.5, ranksep=0.6, splines=ortho]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=12, margin="0.18,0.10"]
  edge  [color="#666666", penwidth=1.6, fontname="Helvetica", fontsize=10, fontcolor="#333"]

  // ─── Upstream chain ────────────────────────────────────────────────
  A [label=<<b>4-CT manuscript-scope cohort</b><br/>
            %d samples (AT 6  ·  DD 8  ·  FB 6  ·  MV 6)<br/>
            %s expressed isoforms (after TMM + 5%% filter)>,
     fillcolor="#e8eef4", color="#2f4858"]

  B [label=<<b>mashr NMD classification</b><br/>
            NMD-sensitive: %s isoforms (union across CTs)<br/>
            Non-NMD: %s isoforms (intersection across CTs)>,
     fillcolor="#fff7bc", color="#d95f0e"]

  C [label=<<b>Pair construction</b> (Isopair)<br/>
            c4: top-two non-NMD isoforms per gene<br/>
            c2: top NMD isoform vs the same reference<br/>
            <font point-size="11">c2 pre-match: %s pairs  ·  c4 pre-match: %s pairs</font>>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  HUB [label=<<b><font point-size="15">pop_BC</font></b><br/>
              Stage-2 gene-matched on (gene, reference)<br/>
              <font point-size="14"><b>%s NMD pairs    /    %s Control pairs</b></font>>,
       fillcolor="#1f3a5f", fontcolor="white", color="#0a1a30", penwidth=2.0]

  A -> B [label=" per-CT mashr + 4-CT union "]
  B -> C [label=" generatePairsExpression "]
  C -> HUB [label=" inner_join on (gene, reference) "]

  // ─── Section A branch (left) ───────────────────────────────────────
  SA_H [label=<<b>SECTION A scope</b><br/>
                all-3-ENST + coding-CDS<br/>
                <font point-size="11">re-intersect on (gene, reference)</font>>,
        fillcolor="#ef8a62", color="#7d4030", fontcolor="#1a0a05"]
  SA1 [label=<+ all-3-ENST (ref AND comp)<br/>
              re-intersect → <b>%s / %s</b>>,
       fillcolor="#f5b294", color="#a2604a"]
  SA2 [label=<+ coding-CDS everywhere<br/>
              re-intersect → <b>%s / %s</b>>,
       fillcolor="#f5b294", color="#a2604a"]
  SA_CL [label=<<b>own-GENCODE-stop classifier</b><br/>
                50-nt last-EJC rule>,
         fillcolor="#fff7bc", color="#d95f0e"]
  SA_PTC  [label=<<b>NMD+/PTC+</b><br/>n = %s  (%.1f%%)>,
           fillcolor="#ef8a62", color="#7d4030"]
  SA_NPTC [label=<<b>NMD+/PTC−</b><br/>n = %s>,
           fillcolor="#fee08b", color="#a07a30"]
  SA_CTL  [label=<<b>Control</b><br/>n = %s>,
           fillcolor="#92c5de", color="#406080"]
  SA_STAT [label=<Fisher OR = %.2f  ·  p = %.2e  ·  %.0f× enrichment>,
           shape=plaintext, fillcolor="white"]
  SA_F3 [label=<<b>Figure 3</b><br/>Panels D / E / F>,
         fillcolor="#dddddd", color="#666"]
  SA_F4 [label=<<b>Figure 4</b><br/>Panels A / B<br/>5\\\'UTR length + ORF>,
         fillcolor="#dddddd", color="#666"]
  SA_SUP [label=<<b>CDS / 3\\\'UTR supplement</b><br/>Row 1 (A / B / C)<br/>n = %s>,
          fillcolor="#dddddd", color="#666"]

  HUB -> SA_H [label="  ENST + coding-CDS  ", color="#7d4030"]
  SA_H -> SA1 -> SA2 -> SA_CL -> {SA_PTC, SA_NPTC, SA_CTL} [color="#7d4030"]
  SA_CL -> SA_STAT [style=dotted, arrowhead=none, color="#999"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_F3 [style=dashed, color="#888"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_F4 [style=dashed, color="#888"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_SUP [style=dashed, color="#888"]

  // ─── Section C branch (right) ──────────────────────────────────────
  SC_H [label=<<b>SECTION C scope</b><br/>
                ENST-reference + ref-AUG-traceable<br/>
                <font point-size="11">re-intersect AFTER category filter</font>>,
        fillcolor="#3690c0", color="#1f5570", fontcolor="white"]
  SC1 [label=<+ ENST-reference only<br/>
              pop_BC ENST-ref subset → <b>%s / %s</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC2 [label=<+ ref-AUG-traceable category per side<br/>
              (effectively_ptc / no_downstream_ejc / truncated_no_ejc)<br/>
              pre re-intersect: <b>%s NMD  /  %s Control</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC3 [label=<re-intersect on (gene, ref) → <b>%s / %s</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC_CL [label=<<b>mechanism_class_4 classifier</b><br/>
                (figures/lib/mechanism_class.R)>,
         fillcolor="#fff7bc", color="#d95f0e"]
  SC_PTC  [label=<<b>NMD+/PTC+</b><br/>n = %s  (%.1f%%)>,
           fillcolor="#ef8a62", color="#7d4030"]
  SC_NPTC [label=<<b>NMD+/PTC−</b><br/>n = %s>,
           fillcolor="#fee08b", color="#a07a30"]
  SC_CTL  [label=<<b>Control</b><br/>n = %s>,
           fillcolor="#92c5de", color="#406080"]
  SC_F4 [label=<<b>Figure 4</b><br/>Panels C / D<br/>ref-AUG 5\\\'UTR + ORF>,
         fillcolor="#dddddd", color="#666"]
  SC_SUP [label=<<b>CDS / 3\\\'UTR supplement</b><br/>Row 2 (D / E / F)<br/>n = %s>,
          fillcolor="#dddddd", color="#666"]

  HUB -> SC_H [label="  ENST-reference  ", color="#1f5570"]
  SC_H -> SC1 -> SC2 -> SC3 -> SC_CL -> {SC_PTC, SC_NPTC, SC_CTL} [color="#1f5570"]
  {SC_PTC, SC_NPTC, SC_CTL} -> SC_F4 [style=dashed, color="#888"]
  {SC_PTC, SC_NPTC, SC_CTL} -> SC_SUP [style=dashed, color="#888"]

  // ─── TD2 cascade (branches off Section C scope) ────────────────────
  TD2_BR [label=<<b>TD2BiasEvidence — broad</b><br/>
                  Section C ∩ TD2 CDS ∩ ref-AUG ORF ∩ ref AUG exonic<br/>
                  <b>n = 819</b>>,
          fillcolor="#a48bbf", color="#4a3556"]
  TD2_OC [label=<<b>TD2BiasEvidence — occult-PTC</b><br/>
                  broad ∩ effectively_ptc ∩ original_ptc = FALSE<br/>
                  <b>n = 348</b><br/>
                  <font point-size="11">TD2/ref ORF 7.67×  ·  ref &gt; TD2 Kozak 286 / 348 (82%%)<br/>
                  paired p = 9.3e−36  ·  TD2 downstream of ref AUG 99.1%%</font>>,
          fillcolor="#7a3f8f", color="#3a1e44", fontcolor="white"]

  SC3 -> TD2_BR [label="  + TD2 CDS available  ", color="#4a3556"]
  TD2_BR -> TD2_OC [label="  + occult-PTC filter  ", color="#4a3556"]

  // ─── Rank constraints — encourage Section A and Section C side-by-side
  { rank=same; SA_H; SC_H }
  { rank=same; SA1; SC1 }
  { rank=same; SA2; SC2 }
  { rank=same; SA_CL; SC_CL }
  { rank=same; SA_PTC, SA_NPTC, SA_CTL; SC_PTC, SC_NPTC, SC_CTL }
}',
  n_samples_4ct, fmt(n_filtered),
  fmt(n_nmd_as), fmt(n_nonnmd_as),
  fmt(n_c2_before), fmt(n_c4_before),
  # HUB (pop_BC)
  fmt(fd_popbc_c2), fmt(fd_popbc_c4),
  # Section A cascade
  fmt(fd_enst_c2), fmt(fd_enst_c4),
  fmt(fd_all3_c2), fmt(fd_all3_c4),
  fmt(fd_ptc_n), fd_ptc_pct,
  fmt(fd_nptc_n), fmt(fd_all3_c4),
  fd_or, fd_p, fd_fold,
  fmt(fd_all3_c2),
  # Section C cascade
  fmt(fd_scC_base_c2), fmt(fd_scC_base_c4),
  fmt(fd_scC_tr_c2),   fmt(fd_scC_tr_c4),
  fmt(fd_scC_c2),      fmt(fd_scC_c4),
  fmt(fd_scC_ptc_n),   fd_scC_ptc_pct,
  fmt(fd_scC_nptc_n),  fmt(fd_scC_c4),
  fmt(fd_scC_c2))   # SC_SUP

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1100, height = 1400)
} else {
  message("DiagrammeR not installed — flowchart not rendered")
}
```

The figure-consumer endpoints (gray boxes) tie each cohort to the figures and
supplements that consume it: Figure 3 D/E/F and Figure 4 A/B + the
CDS/3'UTR supplement Row 1 from Section A; Figure 4 C/D + the
CDS/3'UTR supplement Row 2 from Section C; and the TD2BiasEvidence
supplement from the n=819 → n=348 occult-PTC chain.

## §1a — Transcript-length comparison

Lengths are recomputed inline from `structures.rds`. The legacy Rmd reported
slightly different values because it used the full pre-match c4 set; the
pop_BC-restricted Control median is the correct comparator.

```{r sec1-tx-length, include=FALSE}
tx_lengths_all <- setNames(
  vapply(seq_len(nrow(structures)), function(i)
    sum(structures$exon_ends[[i]] - structures$exon_starts[[i]] + 1L),
    integer(1)),
  structures$isoform_id
)

# pop_BC is gene-matched: each gene contributes exactly one (ref, NMD comp,
# Control comp) triplet. Build the triplet table so Wilcoxon tests are
# PROPERLY PAIRED on the gene-matched design (Pass 2 / M1 fix).
pop_bc_triplets <- merge(
  pop_bc_c2[, c("gene_id", "reference_isoform_id", "comparator_isoform_id")],
  pop_bc_c4[, c("gene_id", "reference_isoform_id", "comparator_isoform_id")],
  by = c("gene_id", "reference_isoform_id"),
  suffixes = c("_nmd", "_ctrl"))
pop_bc_triplets$ref_len  <- tx_lengths_all[pop_bc_triplets$reference_isoform_id]
pop_bc_triplets$nmd_len  <- tx_lengths_all[pop_bc_triplets$comparator_isoform_id_nmd]
pop_bc_triplets$ctrl_len <- tx_lengths_all[pop_bc_triplets$comparator_isoform_id_ctrl]
pop_bc_triplets <- pop_bc_triplets[
  complete.cases(pop_bc_triplets[, c("ref_len", "nmd_len", "ctrl_len")]), ]

ref_lens  <- pop_bc_triplets$ref_len
nmd_lens  <- pop_bc_triplets$nmd_len
ctrl_lens <- pop_bc_triplets$ctrl_len

len_summary <- data.frame(
  Group   = c("Reference (non-NMD)", "NMD comparator", "Control comparator"),
  n       = c(length(ref_lens), length(nmd_lens), length(ctrl_lens)),
  Median  = c(median(ref_lens), median(nmd_lens), median(ctrl_lens)),
  Q25     = c(unname(quantile(ref_lens, 0.25)),
              unname(quantile(nmd_lens, 0.25)),
              unname(quantile(ctrl_lens, 0.25))),
  Q75     = c(unname(quantile(ref_lens, 0.75)),
              unname(quantile(nmd_lens, 0.75)),
              unname(quantile(ctrl_lens, 0.75))),
  row.names = NULL
)
len_pw <- data.frame(
  Comparison = c("NMD vs Reference", "NMD vs Control", "Control vs Reference"),
  p_value    = c(
    wilcox.test(nmd_lens,  ref_lens,  paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(nmd_lens,  ctrl_lens, paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(ctrl_lens, ref_lens,  paired = TRUE, exact = FALSE)$p.value
  )
)
```

Median transcript length (nt):
**Reference** = `r format(median(ref_lens), big.mark=",")`,
**NMD comparator** = `r format(median(nmd_lens), big.mark=",")`,
**Control comparator** = `r format(median(ctrl_lens), big.mark=",")`.

Paired Wilcoxon (each gene contributes one (ref, NMD comp, Control comp)
triplet, so paired tests are the correct comparator on this gene-matched
design): NMD vs Reference `r format_p(len_pw$p_value[1])`;
NMD vs Control `r format_p(len_pw$p_value[2])`;
Control vs Reference `r format_p(len_pw$p_value[3])`.

```{r sec1-tx-length-table, echo=FALSE}
knitr::kable(len_summary, caption = "Transcript length by isoform role (pop_BC scope, nt)",
             digits = 0, format.args = list(big.mark = ","))
knitr::kable(len_pw, caption = "Paired Wilcoxon: transcript length (gene-matched pairs)")
```

## §1b — Annotation status by role

```{r sec1-annotation, include=FALSE}
is_novel_iso <- function(ids) grepl("[.]novel[0-9]+$", ids)

ann_table <- data.frame(
  Role = c("Reference (non-NMD)", "NMD comparator", "Control comparator"),
  n    = c(length(ref_ids_pop), length(nmd_ids_pop), length(ctrl_ids_pop)),
  ENST = c(sum(!is_novel_iso(ref_ids_pop)),
           sum(!is_novel_iso(nmd_ids_pop)),
           sum(!is_novel_iso(ctrl_ids_pop))),
  Novel = c(sum(is_novel_iso(ref_ids_pop)),
            sum(is_novel_iso(nmd_ids_pop)),
            sum(is_novel_iso(ctrl_ids_pop)))
)
ann_table$pct_ENST  <- round(100 * ann_table$ENST  / ann_table$n, 1)
ann_table$pct_Novel <- round(100 * ann_table$Novel / ann_table$n, 1)
```

References are predominantly GENCODE-annotated
(`r ann_table$pct_ENST[ann_table$Role == "Reference (non-NMD)"]`% ENST,
`r format(ann_table$ENST[ann_table$Role == "Reference (non-NMD)"], big.mark=",")` /
`r format(ann_table$n[ann_table$Role == "Reference (non-NMD)"], big.mark=",")`),
while NMD comparators are predominantly novel
(`r ann_table$pct_Novel[ann_table$Role == "NMD comparator"]`%) — consistent with
NMD-sensitive transcripts being underrepresented in reference annotations.

```{r sec1-annotation-table, echo=FALSE}
knitr::kable(ann_table, caption = "Annotation status by role (pop_BC scope)")
```

The Reference %ENST and the NMD-comparator %Novel define the scope reduction
that motivates §2 (the GENCODE-only n=130 scope: where ALL three isoforms are
ENST-annotated, which constrains the cohort sharply).

## §1c — Per-gene isoform diversity and reference-isoform dominance

The §4¶1 descriptive numbers (median expressed-isoforms per gene; reference-
isoform share of parent-gene expression) are computed here from the same
pop_BC universe used for transcript-length in §1a, so the SF
*PairSetDescriptives* draws from these canonical chunks rather than
recomputing.

```{r sec1-iso-per-gene, include=FALSE}
# Expressed non-NMD isoforms per gene across the pop_BC universe.
# Filter: gene in pop_BC; isoform has a row in the TMM-normalized count matrix
# (expression_data.rds); NOT classified as NMD-sensitive by mashr 4-CT union.
nmd_iso_set <- nmd_class[["all_samples"]]$nmd
expr_iso    <- rownames(expr_mat)

sec1_iso <- data.table::as.data.table(structures)[
  gene_id %in% pop_bc_shared$gene_id &
    isoform_id %in% expr_iso &
    !isoform_id %in% nmd_iso_set,
  .(isoform_id, gene_id)]

iso_per_gene <- sec1_iso[, .(n_isoforms = .N), by = gene_id]
# Genes with zero matches keep a 0 count (rare; preserves denominator).
all_pop_genes <- data.table::data.table(gene_id = unique(pop_bc_shared$gene_id))
iso_per_gene <- merge(all_pop_genes, iso_per_gene, by = "gene_id", all.x = TRUE)
iso_per_gene[is.na(n_isoforms), n_isoforms := 0L]

iso_med <- as.integer(median(iso_per_gene$n_isoforms))
iso_iqr <- as.integer(IQR(iso_per_gene$n_isoforms))
```

Across the `r format(nrow(iso_per_gene), big.mark=",")` pop_BC genes, the
median number of expressed non-NMD isoforms per gene is **`r iso_med`**
(IQR = `r iso_iqr`). The pop_BC ≥3-isoform inclusion criterion sets the floor.

```{r sec1-ref-share, include=FALSE}
# Reference-isoform DMSO 4-CT mean expression as a fraction of the gene's
# total non-NMD DMSO 4-CT mean expression. The same denominator (non-NMD
# expressed isoforms of the gene) used in sec1-iso-per-gene.
dmso_4ct_samples <- unique(c(dmso_samples$AT, dmso_samples$DD,
                              dmso_samples$FB, dmso_samples$MV))
stopifnot(all(dmso_4ct_samples %in% colnames(expr_mat)))

dmso_means <- rowMeans(expr_mat[, dmso_4ct_samples, drop = FALSE])
sec1_iso[, mean_dmso := dmso_means[isoform_id]]
sec1_iso[is.na(mean_dmso), mean_dmso := 0]
gene_total_dmso <- sec1_iso[, .(gene_total = sum(mean_dmso)), by = gene_id]

ref_share <- data.table::data.table(
  gene_id              = pop_bc_c2$gene_id,
  reference_isoform_id = pop_bc_c2$reference_isoform_id
)
ref_share <- merge(ref_share, gene_total_dmso, by = "gene_id", all.x = TRUE)
ref_share[, ref_dmso := dmso_means[reference_isoform_id]]
ref_share[is.na(ref_dmso), ref_dmso := 0]
ref_share[, share := ifelse(gene_total > 0, ref_dmso / gene_total, NA_real_)]

ref_share_med_pct  <- round(100 * median(ref_share$share, na.rm = TRUE))
ref_share_ge50_pct <- round(100 * mean(ref_share$share >= 0.50, na.rm = TRUE))
```

The median reference-isoform DMSO 4-CT mean expression captures
**`r ref_share_med_pct`%** of the parent gene's total non-NMD expression;
**`r ref_share_ge50_pct`%** of references are above the 50% threshold.

## §1d — Figure 3 Panels A/B/C

Panels A (pair-set concept), B (sequence similarity), and C (event prevalence)
characterize the pop_BC pair structure.

**Panel C's numbers are computed HERE**, in the chunk below, and the figure only
renders them (S13: no result may originate in `figures/`). Until 2026-07-27 this
report embedded Panel C as an image while its values were derived in
`figures/multipanel/figure3_isopair_and_ptc/data_export.R` — so re-rendering this
report could not check them, and a restyle of the figure could move a published
number silently. Panels A and B are unchanged and still render from the figure
pipeline.

```{r sec1d-panelB-sequence-similarity}
# ── Panel B — sequence similarity of the comparator to the reference ──────────────────
# S13. Panel B was a RESHAPE rather than a derivation -- data_export.R read the div_*
# caches and stacked pct_shared into long form -- but claim 4.2.2 ("the NMD isoforms
# shared more sequence content with the reference") is a RESULT asserted from that
# distribution, and it was stated in no report. It had no producer outside figure code and
# no asserted value at all, so nothing could check it.
#
# The panel data and the comparison behind the sentence now both live here.
div_c2 <- readRDS(file.path(DM, "analysis_cache/div_c2_allsamples.rds"))
div_c4 <- readRDS(file.path(DM, "analysis_cache/div_c4_allsamples.rds"))

# div_* is row-aligned to the gene-matched pop_BC arms; assert rather than assume, because a
# silent misalignment would pair each comparator with another pair's similarity value.
stopifnot("div_c2 not row-aligned to pop_bc_c2" = nrow(div_c2) == nrow(pop_bc_c2),
          "div_c4 not row-aligned to pop_bc_c4" = nrow(div_c4) == nrow(pop_bc_c4))

panelB <- data.table::rbindlist(list(
  data.table::data.table(comparison = "NMD",     pct_shared = div_c2$pct_shared),
  data.table::data.table(comparison = "Control", pct_shared = div_c4$pct_shared)))
panelB <- panelB[!is.na(pct_shared)]

data.table::fwrite(panelB, file.path(.nmd_root,
  "figures/multipanel/figure3_isopair_and_ptc/data/panelB_sequence_similarity.tsv"), sep = "\t")

# The comparison claim 4.2.2 actually makes:
b_med_nmd  <- median(panelB[comparison == "NMD"]$pct_shared)
b_med_ctrl <- median(panelB[comparison == "Control"]$pct_shared)
b_wilcox   <- wilcox.test(pct_shared ~ comparison, data = panelB)

# Claim 4.2.2 is RELATIONAL -- "the NMD isoforms shared more sequence content with the
# reference" -- and states no number, so these are SUPPORTING values with no `published`. The
# assertion is the ORDERING of the two medians, and a reader checks it by reading them.
PANEL_B_POP <- paste("pairs from the gene-matched pair cohort (one reference, one NMD comparator and one Control comparator per gene) with a non-missing exonic-sequence-shared value, one NMD and",
                     "one Control comparator per gene; percentage of the comparator's exonic",
                     "sequence shared with the reference isoform")
claim_emit("4.2.2", "median exonic sequence shared with the reference, NMD (%)", b_med_nmd,
           n = sum(panelB$comparison == "NMD"), population = PANEL_B_POP)
claim_emit("4.2.2", "median exonic sequence shared with the reference, Control (%)", b_med_ctrl,
           n = sum(panelB$comparison == "Control"), population = PANEL_B_POP)
claim_emit("4.2.2", "Wilcoxon p, NMD vs Control sequence shared", b_wilcox$p.value,
           n = nrow(panelB), population = PANEL_B_POP)
```

Median exonic sequence shared with the reference is
**`r sprintf("%.1f%%", b_med_nmd)`** for NMD comparators against
**`r sprintf("%.1f%%", b_med_ctrl)`** for Controls
(Wilcoxon rank-sum p = `r signif(b_wilcox$p.value, 3)`;
n = `r format(sum(panelB$comparison == "NMD"), big.mark=",")` NMD and
`r format(sum(panelB$comparison == "Control"), big.mark=",")` Control pairs with a
non-missing value). This is the comparison behind the Results sentence that NMD isoforms
share more sequence with the reference than Controls do.

```{r sec1d-panelC-event-prevalence}
# ── Data provenance ───────────────────────────────────────────────────────────────────
# Starting data: pop_bc_c2 / pop_bc_c4 from sec1-pop-bc above -- the shared (gene,
# reference) pairs, one NMD comparator and one Control comparator per gene. Event booleans
# (tss_changed, n_se, n_ir, ...) come from buildProfiles -> compareSplicingChoices; the
# Partial_IR_5/3 flags come from the detailed_events list-column.
#
# ── ONE CONVENTION, LOCAL TO THIS TABLE (Pete, 2026-07-27) ────────────────────────────
# IR is reported as the UNION of IR and IR_diff. The justification is biological, and it
# comes from Isopair itself: per cooccurrence.R the two are the same mechanism, and Isopair
# defines has_ir = n_ir + n_ir_diff. The merge is a per-pair OR rather than a sum of bars,
# so a pair carrying BOTH an IR and an IR_diff event is counted once, not twice.
#
# The event caller still separates them, and this collapse is deliberately NOT applied
# globally -- every other event table in the isopair reports keeps IR_diff as its own row.
# Local to this table (Pete, 2026-07-27).
#
# The Control arm needs no adjustment: pop_bc_c4 is already one pair per gene, matched to
# pop_bc_c2. Verified identical to the figure's own gene-matching, 1,578 vs 1,578 with zero
# difference in either direction (verification/phaseC/panelC_from_report_logic.R).
event_counts <- function(pr) c(
  Alt_TSS          = sum(pr$tss_changed),
  Alt_TES          = sum(pr$tes_changed),
  A5SS             = sum(pr$n_a5ss > 0),
  A3SS             = sum(pr$n_a3ss > 0),
  SE               = sum(pr$n_se > 0),
  Missing_Internal = sum(pr$n_missing_internal > 0),
  IR               = sum(pr$n_ir > 0 | pr$n_ir_diff > 0),   # <- the convention
  Partial_IR_5     = sum(sapply(pr$detailed_events, function(d) any(d$event_type == "Partial_IR_5"))),
  Partial_IR_3     = sum(sapply(pr$detailed_events, function(d) any(d$event_type == "Partial_IR_3"))))

evt_nmd  <- event_counts(pop_bc_c2)
evt_ctrl <- event_counts(pop_bc_c4)
n_evt_nmd  <- nrow(pop_bc_c2)
n_evt_ctrl <- nrow(pop_bc_c4)

panelC <- data.frame(
  event_type              = names(evt_nmd),
  n_pairs_with_event_NMD  = as.integer(evt_nmd),
  pct_of_pairs_NMD        = round(100 * evt_nmd  / n_evt_nmd,  1),
  n_pairs_with_event_Ctrl = as.integer(evt_ctrl),
  pct_of_pairs_Ctrl       = round(100 * evt_ctrl / n_evt_ctrl, 1),
  stringsAsFactors = FALSE)

panelC$fisher_OR <- NA_real_; panelC$fisher_p <- NA_real_
for (i in seq_len(nrow(panelC))) {
  ft <- tryCatch(fisher.test(matrix(c(
          panelC$n_pairs_with_event_NMD[i],  n_evt_nmd  - panelC$n_pairs_with_event_NMD[i],
          panelC$n_pairs_with_event_Ctrl[i], n_evt_ctrl - panelC$n_pairs_with_event_Ctrl[i]),
          nrow = 2)), error = function(e) NULL)
  if (!is.null(ft)) { panelC$fisher_OR[i] <- round(ft$estimate, 2); panelC$fisher_p[i] <- ft$p.value }
}
panelC$direction <- ifelse(panelC$fisher_p > 0.05, "Non-significant",
                    ifelse(panelC$fisher_OR > 1, "Enriched in NMD", "Enriched in Control"))
panelC <- panelC[order(panelC$fisher_p), ]

# The figure READS this; it must not recompute it.
panelC_path <- file.path(.nmd_root,
  "figures/multipanel/figure3_isopair_and_ptc/data/panelC_event_prevalence.tsv")
write.table(panelC, panelC_path, sep = "\t", row.names = FALSE, quote = FALSE)

# ── Claims 4.4.3, 4.4.4, 4.2.3 ────────────────────────────────────────────────────────
# THE POPULATION IS THE FINDING HERE. This panel is computed over pop_BC -- every gene's NMD
# and Control pair, no CDS or GENCODE constraint, because splice-event detection needs no CDS.
# The manuscript's Panel C legend says "pairs where all three isoforms are GENCODE annotated
# (n=130)", which is the Section A scope belonging to panels D/E/F. The emission below declares
# the real denominator against that published 130, so the ~12x population error is recorded as a
# disagreement rather than left in prose.
PANEL_C_POP <- paste("pairs from the gene-matched pair cohort (one reference, one NMD comparator and one Control comparator per gene), one NMD comparator and one gene-matched Control comparator",
                     "per gene; NO CDS or GENCODE-annotation constraint, because splice-event",
                     "detection does not require a CDS. IR is reported as the union of IR and",
                     "IR_diff, a per-pair OR local to this table")
claim_emit("4.4.3", "pairs per arm in the Panel C population", n_evt_nmd,
           n = n_evt_nmd, population = PANEL_C_POP)

se_row      <- panelC[panelC$event_type == "SE", ]
se_pct_nmd  <- se_row$pct_of_pairs_NMD
se_pct_ctrl <- se_row$pct_of_pairs_Ctrl
# `published` AS A STRING, NOT A DOUBLE. R renders as.character(51.0) as "51", which drops the
# paper's stated decimal place -- and the checker takes its tolerance from the literal, so the
# comparison silently loosened from +/-0.05 to +/-0.5 and scored the computed 51.3 as AGREEING
# with a published 51.0 that the ledger records as a difference. Measured, not theorised.
claim_emit("4.4.4", "SE prevalence in NMD pairs (%)", se_pct_nmd,
           published = 51.3, n = n_evt_nmd, population = PANEL_C_POP)
claim_emit("4.4.4", "SE prevalence in Control pairs (%)", se_pct_ctrl,
           published = 21.7, n = n_evt_ctrl, population = PANEL_C_POP)
claim_emit("4.4.4", "Fisher p for SE prevalence, NMD vs Control", se_row$fisher_p,
           published = 1e-68, n = n_evt_nmd + n_evt_ctrl, population = PANEL_C_POP)

# 4.2.3 says SE is "twice as frequent" in NMD pairs. The "2" the ledger extracted is an English
# intensifier, not a printed point estimate, so the ratio is a SUPPORTING number: emitting it
# against published = 2 would score the adjudicated-as-reproduced 2.4x as a disagreement.
claim_emit("4.2.3", "SE prevalence ratio, NMD over Control", se_pct_nmd / se_pct_ctrl,
           n = n_evt_nmd + n_evt_ctrl, population = PANEL_C_POP)
```

Panel C is computed over **`r format(n_evt_nmd, big.mark=",")`** NMD pairs and
**`r format(n_evt_ctrl, big.mark=",")`** gene-matched Control pairs. Skipped exon is
**`r panelC$pct_of_pairs_NMD[panelC$event_type == "SE"]`%** of NMD pairs against
**`r panelC$pct_of_pairs_Ctrl[panelC$event_type == "SE"]`%** of Control pairs.

```{r sec1d-panelC-table, echo=FALSE}
knitr::kable(transform(panelC, fisher_p = signif(fisher_p, 3)),
             caption = sprintf("Figure 3 Panel C — splice event prevalence, pop_BC (n=%s NMD / %s Control pairs). IR is the union of IR and IR_diff, a convention local to this table.",
                               format(n_evt_nmd, big.mark=","), format(n_evt_ctrl, big.mark=",")),
             format.args = list(big.mark = ","),
             row.names = FALSE)
```

```{r sec1-fig3abc, echo=FALSE, fig.cap=c("Figure 3 Panel A — Isoform pair construction (schematic)", "Figure 3 Panel B — Sequence similarity to reference", "Figure 3 Panel C — Splice event prevalence"), out.width="100%"}
inc_fig(c(
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelA_pair_concept.png"),
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelB_sequence_similarity.png"),
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelC_event_prevalence.png")
))
```

# §2 — PTC analysis at the GENCODE-only scope (n=130)

The Section A scope is the most stringent "complete GENCODE-only" subset:
every isoform in the pair (Reference, NMD comparator, Control comparator) is
GENCODE-annotated (ENST prefix) AND has a curated coding CDS. This eliminates
TD2-derived CDS calls from the comparison and gives the cleanest possible
PTC-vs-non-PTC contrast.

**Scope construction** (replicated from
[`figures/multipanel/figure4_ptcneg_and_model/data_export.R`](../../../../figures/multipanel/figure4_ptcneg_and_model/data_export.R) §"Section A scope"):

<!-- The computation chunk that was here is now hoisted above the §1 flowchart, so the
     flowchart can label itself from computed values instead of hardcoded literals.
     It is include=FALSE and emits nothing, so nothing moved visibly. 2026-07-26 -->

Filter cascade from pop_BC down to the n=130 Section A scope:

```{r sec2-scope-table, echo=FALSE}
knitr::kable(sec2_cascade,
             caption = sprintf("Section A scope cascade — pop_BC → all-ENST → coding-everywhere → gencode_all3 (n=%s / %s)",
                               format(nrow(gencode_all3_c2), big.mark=","),
                               format(nrow(gencode_all3_c4), big.mark=",")),
             format.args = list(big.mark = ","),
             col.names = c("Filter step", "c2 rows", "c4 rows"))
```

`gencode_all3_c2` = `r nrow(gencode_all3_c2)` NMD pairs;
`gencode_all3_c4` = `r nrow(gencode_all3_c4)` Control pairs. Both arms share
the same `r length(joint2)` `(gene, reference)` pairs.

**Why the cohort drops from 1,385/1,385 ENST-reference pairs to 130 coding-
everywhere pairs:** the ENST-on-comparator filter is the tightest constraint
— most NMD comparators in `pop_BC` are novel isoforms
(`r ann_table$pct_Novel[ann_table$Role == "NMD comparator"]`% per §1b), so
requiring ENST on the NMD-comparator side eliminates the bulk of the cohort.
Requiring a curated CDS on top reduces further to 130. This stringent scope
defines the Section A panels in Figure 3 (D/E/F) and Figure 4 (A/B), plus
the `CDSand3UTR_GENCODEonly` supplement.

## §2a — PTC determination

At this scope every isoform is GENCODE-annotated with a curated coding CDS, so
PTC status is determined per comparator from its own annotated stop's distance
to the last exon-exon junction (50-nt rule). No ref-AUG projection is needed.
This is the methodology used by Figure 3 Panels D/E/F and is replicated inline
from
[`figures/multipanel/figure3_isopair_and_ptc/data_export.R`](../../../../figures/multipanel/figure3_isopair_and_ptc/data_export.R).

<!-- The computation chunk that was here is now hoisted above the §1 flowchart, so the
     flowchart can label itself from computed values instead of hardcoded literals.
     It is include=FALSE and emits nothing, so nothing moved visibly. 2026-07-26 -->

**PTC+ rates** (own GENCODE stop + 50-nt rule, n=`r nrow(gencode_all3_c2)`
NMD pairs / `r nrow(gencode_all3_c4)` Control pairs):

- NMD: `r n_ptc_nmd` / `r n_avail_nmd` = **`r sprintf("%.1f", ptc_rate_nmd)`%**
- Control: `r n_ptc_ctrl` / `r n_avail_ctrl` = **`r sprintf("%.1f", ptc_rate_ctrl)`%**
- Fold enrichment: **`r sprintf("%.1f×", fold_enrich)`**
- Fisher: OR = `r sprintf("%.1f", ptc_fisher$estimate)`, p = `r format_p(ptc_fisher$p.value)`

```{r sec2a-panelD-export}
# ── Panel D — stop-to-last-EJC distance, all-3-ENST scope ─────────────────────────────
# S13. Like Panel B this is a reshape rather than a new quantity: `distance` is computed by
# compute_ptc_info() in sec2-scope above. It moves here so the figure has no reason to
# re-derive it, and so the panel and the report cannot drift.
panelD <- data.table::rbindlist(list(
  data.table::as.data.table(trace_c2)[!is.na(distance),
    .(comparator_isoform_id, gene_id, reference_isoform_id,
      comparison = "NMD", distance, last_ejc_tx_pos, own_stop_tx_pos)],
  data.table::as.data.table(trace_c4)[!is.na(distance),
    .(comparator_isoform_id, gene_id, reference_isoform_id,
      comparison = "Control", distance, last_ejc_tx_pos, own_stop_tx_pos)]))

data.table::fwrite(panelD, file.path(.nmd_root,
  "figures/multipanel/figure3_isopair_and_ptc/data/panelD_stop_codon_distance.tsv"), sep = "\t")
```

```{r sec2a-fig3-panelD, echo=FALSE, fig.cap="Figure 3 Panel D — Stop-to-last-EJC distance at the n=130 scope", out.width="80%"}
inc_fig(
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelD_stop_codon_distance.png")
)
```

## §2b — PTC-causing event attribution

For each of the `r n_ptc_nmd` PTC+ NMD pairs, we attribute the splice event
that introduced the PTC. Two attribution functions are applied:

- `attribute_ptc_events()` — matches a `detailed_events` row to the stop
  position via `fw_events` (forward-walked junctions); handles the diff-stop
  case where the comparator's PTC is at a different genomic position from
  the reference's stop.
- `attribute_3utr_splice()` — for same-stop pairs (comparator's own GENCODE
  stop equals the reference's stop), looks for splice events DOWNSTREAM of
  the stop in the 3'UTR.

For the small set of pairs where both functions return a `"direct"` row, the
same-stop call takes precedence (a 3'UTR splice is a more specific
attribution than a diff-stop event when the stops coincide).

<!-- The computation chunk that was here is now hoisted above the §1 flowchart, so the
     flowchart can label itself from computed values instead of hardcoded literals.
     It is include=FALSE and emits nothing, so nothing moved visibly. 2026-07-26 -->

**Attribution accounting:**
`r n_direct_b` direct (from `attribute_ptc_events`) +
`r n_same_stop_direct` same-stop 3'UTR splice =
`r n_direct_b + n_same_stop_direct` total attribution rows →
`r n_unique_direct` unique direct (same-stop takes precedence over diff-stop;
`r n_direct_b + n_same_stop_direct - n_unique_direct` rows were attributed by
both functions, with the same-stop call kept) →
`r n_ptc_nmd - n_unique_direct` unresolved out of `r n_ptc_nmd` PTC+ pairs
(`r sprintf("%.1f%%", 100 * n_unique_direct / n_ptc_nmd)` attribution coverage).

The pre-combine `attribute_ptc_events` pass left `r n_unresolved_b`
unresolved; the same-stop pass rescued
`r n_unresolved_b - (n_ptc_nmd - n_unique_direct)`
of those, leaving `r n_ptc_nmd - n_unique_direct` finally unresolved.

**Mechanism breakdown** of the `r n_unique_direct` attributed PTC events:

```{r sec2b-mech-table, echo=FALSE}
knitr::kable(mech_tab,
             caption = sprintf("Mechanism breakdown for the %d attributed PTC events (0 of 48 PTC+ unresolved; Section A n=130 scope)", n_unique_direct),
             col.names = c("Mechanism", "n"))
```

**Panels E and F are computed HERE** (S13), from the attribution already performed in
`sec2b-attribution` above; the figure only renders them. Until 2026-07-27 they were
derived in `figures/multipanel/figure3_isopair_and_ptc/panel_e_compute.R`, which read
the same two caches and repeated the same attribution.

```{r sec2b-panelEF-export}
# ── Panel E — PTC-causing event enrichment vs the Control baseline ────────────────────
# NMD side: the attributed PTC-causing events from sec2b-attribution (all_attr_new).
# Control side: every detailed_event in the all-3-ENST Control arm (gencode_all3_c4).
#
# ONE CONVENTION, LOCAL TO THIS TABLE: IR_diff collapses into IR on the CONTROL side, so
# the two arms are counted on the same event vocabulary. Same biological justification as
# Panel C (Isopair cooccurrence.R: has_ir = n_ir + n_ir_diff).
#
# NOTE THE DENOMINATORS DIFFER BY CONSTRUCTION and that is deliberate: the NMD side counts
# ATTRIBUTED events (one per PTC+ pair) while the Control side counts ALL events in the arm.
# It is an event-composition comparison, not a per-pair rate, which is why `enrichment` is a
# ratio of percentages rather than an odds ratio.
# MERGE IR_diff* INTO IR AT THE SOURCE, on a LOCAL COPY. The figure does this at
# panel_e_compute.R:113, BEFORE deriving event_short, so both the Panel E proportion table
# and the Panel F mechanism breakdown treat IR_diff as IR -- same mechanism, same Isopair
# justification as Panel C (has_ir = n_ir + n_ir_diff).
#
# On a COPY, not on all_attr_new itself: the convention is local to these two tables (Pete),
# and the report's own event_short at the attribution chunk keeps IR_diff separate. Applying
# it in place would silently redefine an object other chunks share.
#
# Found by the byte-identity test, not by reading: collapsing only the CONTROL side left
# Panel E with IR n_ptc=3 plus a stray IR_diff_3 row against the figure's IR n_ptc=4.
attr_EF <- data.table::copy(data.table::as.data.table(all_attr_new))
attr_EF[grepl("^IR_diff", ptc_causing_event), ptc_causing_event := "IR"]
attr_EF$event_short <- shorten_event_labels(attr_EF$ptc_causing_event, collapse_rare = 3)

n_ptc_attr_e <- nrow(attr_EF)
ptc_evt_freq <- sort(table(attr_EF$ptc_causing_event), decreasing = TRUE)
panelE <- data.table::data.table(
  event_type   = names(ptc_evt_freq),
  n_ptc_events = as.integer(ptc_evt_freq),
  pct_of_ptc   = round(100 * as.integer(ptc_evt_freq) / n_ptc_attr_e, 1))

ctrl_all_events <- do.call(rbind, lapply(gencode_all3_c4$detailed_events, as.data.frame))
ctrl_all_events$event_type[grepl("^IR_diff", ctrl_all_events$event_type)] <- "IR"
ctrl_event_freq <- table(ctrl_all_events$event_type)
ctrl_total_e    <- sum(ctrl_event_freq)

panelE[, n_ctrl_events := as.integer(ctrl_event_freq[event_type])]
panelE[is.na(n_ctrl_events), n_ctrl_events := 0L]
panelE[, pct_ctrl   := round(100 * n_ctrl_events / ctrl_total_e, 1)]
panelE[, enrichment := round(pct_of_ptc / pmax(pct_ctrl, 0.1), 1)]
panelE[, fisher_p := NA_real_][, direction := NA_character_]
for (i in seq_len(nrow(panelE))) {
  ft <- tryCatch(fisher.test(matrix(c(
          panelE$n_ptc_events[i],  n_ptc_attr_e - panelE$n_ptc_events[i],
          panelE$n_ctrl_events[i], ctrl_total_e - panelE$n_ctrl_events[i]), nrow = 2)),
        error = function(e) NULL)
  if (!is.null(ft)) {
    panelE$fisher_p[i]  <- signif(ft$p.value, 3)
    panelE$direction[i] <- if (ft$p.value > 0.05) "Non-significant"
      else if (panelE$enrichment[i] > 1) "Enriched in PTC-causing" else "Depleted in PTC-causing"
  }
}
panelE <- panelE[, .(event_type, n_ptc_events, pct_of_ptc, n_ctrl_events, pct_ctrl,
                     enrichment, fisher_p, direction)]

# ── Panel F — mechanism breakdown x event type (long form for the stacked bar) ────────
attr_F <- data.table::copy(attr_EF)   # same IR-collapsed copy Panel E uses
attr_F[, mechanism := factor(mechanism,
        levels = c("Frameshift", "In-frame stop", "3'UTR splice"))]
sankey_agg <- attr_F[!is.na(mechanism), .(n = .N), by = .(event_short, mechanism)]
evt_totals <- sankey_agg[, .(total = sum(n)), by = event_short][order(-total)]
sankey_agg[, event_short := factor(event_short, levels = evt_totals$event_short)]
panelF <- sankey_agg[order(event_short, mechanism),
                     .(event_type = as.character(event_short),
                       mechanism  = as.character(mechanism), n)]

# The figures READ these; they must not recompute them.
data.table::fwrite(panelE, file.path(.nmd_root,
  "figures/multipanel/figure3_isopair_and_ptc/data/panelE_ptc_event_attribution.tsv"), sep = "\t")
data.table::fwrite(panelF, file.path(.nmd_root,
  "figures/multipanel/figure3_isopair_and_ptc/data/panelF_mechanism_breakdown.tsv"), sep = "\t")

# ── Claims 4.4.8, 4.4.9, 4.6.2 — Panel E event composition ────────────────────────────
# THE TWO DENOMINATORS ARE DIFFERENT BY CONSTRUCTION and the manuscript does not say so, which
# is why both are spelled out here: the NMD side counts one ATTRIBUTED event per PTC-positive
# pair, the Control side counts EVERY detailed event in the Control arm. `enrichment` is
# therefore a ratio of percentages, not an odds ratio.
PANEL_E_POP_NMD <- paste("splicing events attributed as the direct cause of a PTC, one per",
                         "PTC-positive pair in the GENCODE reference set; IR_diff collapsed",
                         "into IR on a local copy so both arms use one event vocabulary")
PANEL_E_POP_CTRL <- paste("ALL detailed splicing events in the Control arm of the GENCODE",
                          "reference set -- not one per pair -- with IR_diff collapsed into IR")

claim_emit("4.4.8", "attributed PTC-causing events", n_ptc_attr_e,
           published = 48, n = n_ptc_attr_e, population = PANEL_E_POP_NMD)
claim_emit("4.4.8", "total Control splicing events (Panel E baseline)", ctrl_total_e,
           published = 298, n = ctrl_total_e, population = PANEL_E_POP_CTRL)

# 4.4.9 and 4.6.2 assert the same three events at different rounding. One emission per number:
# a single row declaring nine published values would be matched set-wise, with no positional
# pairing, which is the property D19 voided 25 results over.
e_row <- function(ev) panelE[event_type == ev]
for (spec in list(
      list(ev = "SE",      pct = 50.0, ctrl = 14.4, p = 2e-7,  pct6 = 50, ctrl6 = 14),
      list(ev = "A5SS",    pct = 12.5, ctrl = 4.0,  p = 0.03,  pct6 = 13, ctrl6 = 4),
      list(ev = "Alt_TES", pct = 4.2,  ctrl = "26.2", p = 3e-4, pct6 = NA, ctrl6 = NA))) {
  r <- e_row(spec$ev)
  if (nrow(r) != 1L) next
  claim_emit("4.4.9", sprintf("%s share of PTC-causing events (%%)", spec$ev), r$pct_of_ptc,
             published = spec$pct, n = n_ptc_attr_e, population = PANEL_E_POP_NMD)
  claim_emit("4.4.9", sprintf("%s share of Control events (%%)", spec$ev), r$pct_ctrl,
             published = spec$ctrl, n = ctrl_total_e, population = PANEL_E_POP_CTRL)
  claim_emit("4.4.9", sprintf("Fisher p for %s, PTC-causing vs Control", spec$ev), r$fisher_p,
             published = spec$p, n = n_ptc_attr_e + ctrl_total_e,
             population = paste(PANEL_E_POP_NMD, "|| compared against:", PANEL_E_POP_CTRL))
  # 4.6.2 restates SE and A5SS at whole-percent precision. Alt_TES is not in that sentence.
  if (!is.na(spec$pct6)) {
    claim_emit("4.6.2", sprintf("%s share of PTC-causing events (%%)", spec$ev), r$pct_of_ptc,
               published = spec$pct6, n = n_ptc_attr_e, population = PANEL_E_POP_NMD)
    claim_emit("4.6.2", sprintf("%s share of Control events (%%)", spec$ev), r$pct_ctrl,
               published = spec$ctrl6, n = ctrl_total_e, population = PANEL_E_POP_CTRL)
  }
}
# 4.6.2 states both p-values as INEQUALITIES ("p < 10^-6", "p < 0.05"), so they are supporting
# numbers with the paper's threshold named -- see the note at claim 4.5.2.
claim_emit("4.6.2", "Fisher p for SE (paper states p < 1e-6)", e_row("SE")$fisher_p,
           n = n_ptc_attr_e + ctrl_total_e, population = PANEL_E_POP_NMD)
claim_emit("4.6.2", "Fisher p for A5SS (paper states p < 0.05)", e_row("A5SS")$fisher_p,
           n = n_ptc_attr_e + ctrl_total_e, population = PANEL_E_POP_NMD)
```

Panel E compares **`r n_ptc_attr_e`** attributed PTC-causing events against a Control
baseline of **`r format(ctrl_total_e, big.mark=",")`** events. Panel F breaks the same
**`r sum(panelF$n)`** attributed events down by mechanism.

```{r sec2b-panelE-table, echo=FALSE}
knitr::kable(panelE,
             caption = sprintf("Figure 3 Panel E — PTC-causing event enrichment (n=%d attributed events vs %s Control events)",
                               n_ptc_attr_e, format(ctrl_total_e, big.mark = ",")),
             row.names = FALSE)
```

```{r sec2b-fig3-panelE-F, echo=FALSE, fig.cap=c("Figure 3 Panel E — PTC-causing event attribution (per-event Fisher table)", "Figure 3 Panel F — Mechanism breakdown × event type"), out.width="80%"}
inc_fig(c(
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelE_ptc_event_attribution.png"),
  file.path(FIG_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelF_mechanism_breakdown.png")
))
```

## §2c — 5'UTR length + longest 5'UTR ORF at the n=130 scope

At the n=130 all-3-ENST scope, the 5'UTR is the comparator's annotated 5'UTR
(every isoform is GENCODE-annotated so the TD2-anchored `utr5_features_all`
cache value equals the GENCODE 5'UTR length at this scope; the Rmd reads it
from cache rather than recomputing). Three groups (using §2a `has_ptc`):

- **NMD+/PTC+** (`r n_ptc_nmd` pairs) — `has_ptc == TRUE`
- **NMD+/PTC−** (`r n_avail_nmd - n_ptc_nmd` pairs) — `has_ptc == FALSE`
- **Control** (`r n_avail_ctrl` pairs) — `gencode_all3_c4` comparators

The Control arm is a single group (not PTC-stratified) because PTC status is
a property of the NMD-comparator. This matches the figure-side panel
construction.

Interpretive caution: at the n=130 scope, NMD+/PTC+ comparators have markedly
short 5'UTRs (median 84.5 nt vs Control 138 nt) — this is most likely an
alternative-TSS effect (PTC+ ENSTs at this scope tend to be 5'-truncated
alternative transcripts) rather than 5'UTR cis-element divergence per se.
The NMD+/PTC− 5'UTR enrichment is therefore the more biologically meaningful
contrast for uORF-mediated NMD.

```{r sec2c-helpers, include=FALSE}
hodges_lehmann <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) < 2L || length(y) < 2L) return(c(NA_real_, NA_real_, NA_real_, NA_real_))
  wt <- tryCatch(wilcox.test(x, y, conf.int = TRUE, exact = FALSE),
                  error = function(e) NULL)
  if (is.null(wt)) return(c(NA_real_, NA_real_, NA_real_, NA_real_))
  c(unname(wt$estimate), unname(wt$conf.int[1]), unname(wt$conf.int[2]),
    unname(wt$p.value))
}
cliffs_delta <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) == 0L || length(y) == 0L) return(NA_real_)
  (sum(outer(x, y, ">")) - sum(outer(x, y, "<"))) / (length(x) * length(y))
}
pairwise_stats <- function(groups_list, lookup) {
  combos <- combn(names(groups_list), 2, simplify = FALSE)
  do.call(rbind, lapply(combos, function(p) {
    x <- lookup[groups_list[[p[1]]]]; y <- lookup[groups_list[[p[2]]]]
    x <- x[!is.na(x)]; y <- y[!is.na(y)]
    hl <- hodges_lehmann(x, y)
    d  <- cliffs_delta(x, y)
    data.frame(group_x = p[1], group_y = p[2],
               n_x = length(x), n_y = length(y),
               median_x = round(median(x), 1), median_y = round(median(y), 1),
               hl_shift = round(hl[1], 1),
               hl_ci = sprintf("[%.1f, %.1f]", hl[2], hl[3]),
               cliffs_delta = round(d, 3),
               wilcox_p = hl[4],
               stringsAsFactors = FALSE)
  }))
}
group_descriptives <- function(groups_list, lookup, order_names) {
  out <- do.call(rbind, lapply(order_names, function(g) {
    v <- lookup[groups_list[[g]]]
    v <- v[!is.na(v)]
    data.frame(group = g, n = length(v),
               median = round(median(v), 1),
               q25 = round(unname(quantile(v, 0.25)), 1),
               q75 = round(unname(quantile(v, 0.75)), 1),
               stringsAsFactors = FALSE)
  }))
  rownames(out) <- NULL
  out
}
```

```{r sec2c-secA-groups, include=FALSE}
ptc_pos_ids <- trace_c2[has_ptc == TRUE,  comparator_isoform_id]
ptc_neg_ids <- trace_c2[has_ptc == FALSE, comparator_isoform_id]
sec_A_groups <- list(
  "NMD+/PTC+" = ptc_pos_ids,
  "NMD+/PTC-" = ptc_neg_ids,
  "Control"   = gencode_all3_c4$comparator_isoform_id
)

u5_all_dt <- as.data.table(utr5_all$isoform_features)
u5_len_lookup <- setNames(as.numeric(u5_all_dt$utr5_length),    u5_all_dt$isoform_id)
u5_orf_lookup <- setNames(as.numeric(u5_all_dt$longest_orf_nt), u5_all_dt$isoform_id)

secA_utr5_desc <- group_descriptives(sec_A_groups, u5_len_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr5_pw   <- pairwise_stats(sec_A_groups, u5_len_lookup)
secA_orf_desc  <- group_descriptives(sec_A_groups, u5_orf_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_orf_pw    <- pairwise_stats(sec_A_groups, u5_orf_lookup)

secA_utr5_pw$wilcox_p_fmt <- vapply(secA_utr5_pw$wilcox_p, format_p, character(1))
secA_orf_pw$wilcox_p_fmt  <- vapply(secA_orf_pw$wilcox_p,  format_p, character(1))

# ── Claim 4.6.4 — three assertions, no published numbers ──────────────────────────────
# "NMD+/PTC- isoforms had longer 5'UTRs and uORFs than BOTH other groups, with no significant
# difference in CDS length." The paper prints no values, so every emission here is SUPPORTING:
# the claim is an ordering plus a null, and a reader checks it from these six numbers.
# BOTH comparisons matter, not just one -- "than both other groups" fails if PTC- exceeds only
# Control. The CDS half is emitted in `sec2d-cds` below, where secA_cds_pw is built.
SEC_A_FEATURE_POP <- paste("comparator isoforms of the GENCODE reference set, split into",
                           "NMD+/PTC+, NMD+/PTC- and Control; isoforms with a missing value for",
                           "the feature are dropped, so n differs between features")
med_of <- function(desc, g) desc$median[match(g, desc$group)]
for (spec in list(list(d = secA_utr5_desc, what = "5'UTR length (nt)"),
                  list(d = secA_orf_desc,  what = "longest 5'UTR ORF (nt)"))) {
  for (g in c("NMD+/PTC+", "NMD+/PTC-", "Control")) {
    claim_emit("4.6.4", sprintf("median %s, %s", spec$what, g), med_of(spec$d, g),
               n = spec$d$n[match(g, spec$d$group)], population = SEC_A_FEATURE_POP)
  }
}
```

### Panel A — 5'UTR length

```{r sec2c-panelA-tables, echo=FALSE}
knitr::kable(secA_utr5_desc,
             caption = "5'UTR length (nt) — Section A descriptives",
             col.names = c("Group", "n", "Median", "Q25", "Q75"))
knitr::kable(secA_utr5_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "5'UTR length — pairwise stats (Hodges–Lehmann shift, Cliff's δ, Wilcoxon p)",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### Panel B — Longest 5'UTR ORF

```{r sec2c-panelB-tables, echo=FALSE}
knitr::kable(secA_orf_desc,
             caption = "Longest 5'UTR ORF (nt) — Section A descriptives",
             col.names = c("Group", "n", "Median", "Q25", "Q75"))
knitr::kable(secA_orf_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "Longest 5'UTR ORF — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

```{r sec2c-fig4AB, echo=FALSE, fig.cap=c("Figure 4 Panel A — 5'UTR length at n=130 scope", "Figure 4 Panel B — Longest 5'UTR ORF at n=130 scope"), out.width="90%"}
inc_fig(c(
  file.path(FIG_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelA_5utr_length_all3enst.png"),
  file.path(FIG_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelB_longest_5utr_orf_all3enst.png")
))
```

## §2d — 3'UTR / CDS sanity at n=130

Three additional structural features for the same `sec_A_groups`: CDS length,
translation-based 3'UTR (`tx_len − own_stop_tx_pos`), and bias-corrected
3'UTR via the `utr3_via_non_ptc_stop_nt` field of `ref_atg` (walks past PTCs
to find the next non-PTC stop). The non-PTC measure is the more meaningful
3'UTR-length comparator because for PTC+ pairs the translation-based 3'UTR
inflates with PTC distance and is no longer comparable to the Control 3'UTR.

```{r sec2d-features, include=FALSE}
own_atg_tx_lookup <- (function() {
  atg_tx <- rep(NA_integer_, nrow(coding_cds))
  for (i in seq_len(nrow(coding_cds))) {
    si <- struct_idx[i]; if (is.na(si)) next
    strand <- coding_cds$strand[i]
    if (is.na(strand) || strand == "") next
    atg_g <- if (strand == "+") coding_cds$cds_start[i] else coding_cds$cds_stop[i]
    atg_tx[i] <- Isopair::genomicToTranscript(
      atg_g, structures$exon_starts[[si]], structures$exon_ends[[si]], strand)
  }
  setNames(atg_tx, coding_cds$isoform_id)
})()

tx_len_lookup <- setNames(
  vapply(seq_len(nrow(structures)), function(i)
    sum(structures$exon_ends[[i]] - structures$exon_starts[[i]] + 1L),
    integer(1)),
  structures$isoform_id
)

# Build per-comparator features for sec_A_groups
ra_c2_lookup <- setNames(
  as.data.table(ref_atg$c2)$utr3_via_non_ptc_stop_nt,
  as.data.table(ref_atg$c2)$comparator_isoform_id
)
ra_c4_lookup <- setNames(
  as.data.table(ref_atg$c4)$utr3_via_non_ptc_stop_nt,
  as.data.table(ref_atg$c4)$comparator_isoform_id
)

compute_secA_feat <- function(comp_ids, ra_lookup) {
  own_stop <- own_stop_tx_lookup$own_stop_tx_pos[match(comp_ids, own_stop_tx_lookup$isoform_id)]
  own_atg  <- own_atg_tx_lookup[comp_ids]
  tx_len   <- tx_len_lookup[comp_ids]
  data.table(
    comparator_isoform_id = comp_ids,
    cds_nt   = pmax(0L, own_stop - own_atg + 1L),
    utr3_nt  = pmax(0L, tx_len - own_stop),
    utr3_via_non_ptc_stop_nt = ra_lookup[comp_ids]
  )
}

feat_ptc_pos <- compute_secA_feat(ptc_pos_ids, ra_c2_lookup)
feat_ptc_neg <- compute_secA_feat(ptc_neg_ids, ra_c2_lookup)
feat_ctrl    <- compute_secA_feat(gencode_all3_c4$comparator_isoform_id, ra_c4_lookup)
# At the n=130 all-3-ENST + coding-CDS scope every comparator's own GENCODE
# stop is non-PTC by classification (PTC+ pairs were segregated via §2a's
# has_ptc flag), so for PTC- and Control the bias-corrected 3'UTR length IS
# the translation-based measure. Overwrite unconditionally (matches the
# figure-side CDSand3UTR_GENCODEonly/data_export.R convention).
feat_ptc_neg[, utr3_via_non_ptc_stop_nt := utr3_nt]
feat_ctrl[,    utr3_via_non_ptc_stop_nt := utr3_nt]

cds_lookup <- setNames(
  c(feat_ptc_pos$cds_nt, feat_ptc_neg$cds_nt, feat_ctrl$cds_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)
utr3_lookup <- setNames(
  c(feat_ptc_pos$utr3_nt, feat_ptc_neg$utr3_nt, feat_ctrl$utr3_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)
utr3_npstop_lookup <- setNames(
  c(feat_ptc_pos$utr3_via_non_ptc_stop_nt,
    feat_ptc_neg$utr3_via_non_ptc_stop_nt,
    feat_ctrl$utr3_via_non_ptc_stop_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)

secA_cds_desc  <- group_descriptives(sec_A_groups, cds_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_cds_pw    <- pairwise_stats(sec_A_groups, cds_lookup)
secA_utr3_desc <- group_descriptives(sec_A_groups, utr3_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr3_pw   <- pairwise_stats(sec_A_groups, utr3_lookup)
secA_utr3np_desc <- group_descriptives(sec_A_groups, utr3_npstop_lookup,
                                        c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr3np_pw   <- pairwise_stats(sec_A_groups, utr3_npstop_lookup)

secA_cds_pw$wilcox_p_fmt    <- vapply(secA_cds_pw$wilcox_p,    format_p, character(1))
secA_utr3_pw$wilcox_p_fmt   <- vapply(secA_utr3_pw$wilcox_p,   format_p, character(1))
secA_utr3np_pw$wilcox_p_fmt <- vapply(secA_utr3np_pw$wilcox_p, format_p, character(1))
```

### CDS length

```{r sec2d-cds, echo=FALSE}
# Claim 4.6.4's third assertion — "no significant difference in CDS length". A NULL result, so
# what has to be recorded is the p-value of each comparison involving NMD+/PTC-, not a median:
# the claim is that neither reaches significance, and one of the two being null is not the
# claim. Supporting numbers, since the paper prints none.
for (i in seq_len(nrow(secA_cds_pw))) {
  gx <- secA_cds_pw$group_x[i]; gy <- secA_cds_pw$group_y[i]
  if (!("NMD+/PTC-" %in% c(gx, gy))) next
  claim_emit("4.6.4", sprintf("Wilcoxon p for CDS length, %s vs %s", gx, gy),
             secA_cds_pw$wilcox_p[i], n = nrow(secA_cds_desc),
             population = SEC_A_FEATURE_POP)
}

knitr::kable(secA_cds_desc, caption = "CDS length (nt)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_cds_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "CDS length — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### 3'UTR length (translation-based)

```{r sec2d-utr3, echo=FALSE}
knitr::kable(secA_utr3_desc, caption = "3'UTR length, translation-based (own stop → tx end)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_utr3_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "3'UTR translation-based — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### 3'UTR length (bias-corrected, walks past PTCs)

```{r sec2d-utr3np, echo=FALSE}
knitr::kable(secA_utr3np_desc,
             caption = "3'UTR length, bias-corrected (non-PTC-stop based)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_utr3np_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "3'UTR bias-corrected — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

<!-- MOVED 2026-08-11. This export chunk stood at line 1903, AFTER sec2d-supp below, and that
     ordering made the report unrenderable from a clean state. sec2d-supp embeds SF36.png;
     figure_sf36.py reads SF33/data/features_1166_subset_long.tsv; SF33's data_export.R does not
     compute that file; and THIS chunk is its only writer. The report therefore had to pass
     sec2d-supp to produce what sec2d-supp requires. Demonstrated in cleanroom_2026-08-11: an
     8-pass render-build loop reached 13 of 14 required figures and stalled here for four passes.
     The chunk is include=FALSE, so moving it emits nothing new and changes no rendered output;
     its inputs refaug_matched_c2/c4 are created at lines 666-668, far above either position. -->
```{r sec3d-refaug-cds-utr3-export, include=FALSE}
nmd_C <- refaug_matched_c2[, .(comparator_isoform_id, group = group3,
                                cds_nt = comp_orf_length_from_ref_atg,
                                utr3_to_tx_end_nt, utr3_via_non_ptc_stop_nt)]
ctrl_C <- refaug_matched_c4[, .(comparator_isoform_id, group = "Control",
                                 cds_nt = comp_orf_length_from_ref_atg,
                                 utr3_to_tx_end_nt, utr3_via_non_ptc_stop_nt)]
all_feat_C <- rbind(nmd_C, ctrl_C, fill = TRUE)
all_feat_C[, group := factor(group, levels = c("NMD+/PTC+", "NMD+/PTC-", "Control"))]

SF33_DATA <- file.path(.nmd_root, "figures/supplemental/SF33_CdsAnd3UTR_GENCODE/data")
dir.create(SF33_DATA, recursive = TRUE, showWarnings = FALSE)

data.table::fwrite(
  all_feat_C[, .(comparator_isoform_id, group, cds_nt,
                 utr3_to_tx_end_nt, utr3_via_non_ptc_stop_nt)],
  file.path(SF33_DATA, "features_1166_subset_long.tsv"), sep = "\t")

emit_per_measure_C <- function(measure_col, slug) {
  groups <- c("NMD+/PTC+", "NMD+/PTC-", "Control")
  desc <- data.table::rbindlist(lapply(groups, function(g) {
    vals <- all_feat_C[group == g, get(measure_col)]; vals <- vals[!is.na(vals)]
    data.table::data.table(group = g, n = length(vals), median = median(vals),
                           q25 = quantile(vals, 0.25), q75 = quantile(vals, 0.75))
  }))
  data.table::fwrite(desc, file.path(SF33_DATA, sprintf("%s_descriptives.tsv", slug)), sep = "\t")
  pairs <- list(c("NMD+/PTC-", "NMD+/PTC+"), c("NMD+/PTC-", "Control"),
                c("NMD+/PTC+", "Control"))
  pw <- data.table::rbindlist(lapply(pairs, function(p) {
    a <- all_feat_C[group == p[1], get(measure_col)]; a <- a[!is.na(a)]
    b <- all_feat_C[group == p[2], get(measure_col)]; b <- b[!is.na(b)]
    w <- suppressWarnings(wilcox.test(a, b, conf.int = TRUE))
    data.table::data.table(
      group_x = p[1], group_y = p[2], n_x = length(a), n_y = length(b),
      median_x = median(a), median_y = median(b),
      hl_shift = as.numeric(w$estimate),
      hl_ci_lo = as.numeric(w$conf.int[1]), hl_ci_hi = as.numeric(w$conf.int[2]),
      cliffs_delta = (sum(outer(a, b, ">")) - sum(outer(a, b, "<"))) / (length(a) * length(b)),
      wilcox_p = w$p.value)
  }))
  data.table::fwrite(pw, file.path(SF33_DATA, sprintf("%s_pairwise.tsv", slug)), sep = "\t")
  invisible(NULL)
}
emit_per_measure_C("cds_nt",                   "panelD_cds_length_refaug")
emit_per_measure_C("utr3_to_tx_end_nt",        "panelE_utr3_translation_refaug")
emit_per_measure_C("utr3_via_non_ptc_stop_nt", "panelF_utr3_non_ptc_stop_refaug")

secC_utr3_desc   <- fread(file.path(SF33_DATA, "panelE_utr3_translation_refaug_descriptives.tsv"))
secC_utr3np_desc <- fread(file.path(SF33_DATA, "panelF_utr3_non_ptc_stop_refaug_descriptives.tsv"))
secC_utr3_pw     <- fread(file.path(SF33_DATA, "panelE_utr3_translation_refaug_pairwise.tsv"))
secC_utr3np_pw   <- fread(file.path(SF33_DATA, "panelF_utr3_non_ptc_stop_refaug_pairwise.tsv"))
```

```{r sec2d-supp, echo=FALSE, fig.cap="CDSand3UTR_GENCODEonly supplement — 2×3 layout. Row 1 (Section A, n=130): A CDS length, B 3'UTR translation-based, C 3'UTR non-PTC-stop (bias-corrected). Row 2 (Section C, n=819, ref-AUG-projected): D CDS length, E 3'UTR translation-based, F 3'UTR non-PTC-stop. The bias-correction pattern (PTC+ inflation under translation-based collapses under non-PTC-stop measure) replicates across both scopes.", out.width="100%"}
inc_fig(
  c(file.path(FIG_ROOT, "figures/supplemental/SF33_CdsAnd3UTR_GENCODE/figure_sf33.png"),
    file.path(FIG_ROOT, "figures/supplemental/SF36_CdsAnd3UTR_refAUG/figure_sf36.png"))
)
```

# §3 — Larger-scope replication at ref-AUG-traceable (n=819)

Section C extends Section A by relaxing the all-3-ENST + coding-CDS
constraints and instead routing through the ref-AUG-projected ORF
classification. This roughly 6× larger cohort lets us check that Section A's
PTC findings replicate at the larger scope without relying on TD2's CDS calls
on novel comparators.

**Scope construction** (replicated from
[`figures/multipanel/figure4_ptcneg_and_model/data_export.R`](../../../../figures/multipanel/figure4_ptcneg_and_model/data_export.R) §"Section C scope"):

<!-- The computation chunk that was here is now hoisted above the §1 flowchart, so the
     flowchart can label itself from computed values instead of hardcoded literals.
     It is include=FALSE and emits nothing, so nothing moved visibly. 2026-07-26 -->

Filter cascade from pop_BC down to the n=819 Section C scope:

```{r sec3-scope-table, echo=FALSE}
knitr::kable(sec3_cascade,
             caption = sprintf("Section C scope cascade — pop_BC → ENST-ref → ref-AUG-traceable → re-intersect (n=%s / %s)",
                               format(nrow(refaug_matched_c2), big.mark=","),
                               format(nrow(refaug_matched_c4), big.mark=",")),
             format.args = list(big.mark = ","),
             col.names = c("Filter step", "c2 rows", "c4 rows"))
```

`refaug_matched_c2` = `r format(nrow(refaug_matched_c2), big.mark=",")` NMD
pairs; `refaug_matched_c4` = `r format(nrow(refaug_matched_c4), big.mark=",")`
Control pairs. Both arms share the same
`r format(nrow(shared_keys), big.mark=",")` `(gene, reference)` pairs.

**Mechanism classification** of the NMD arm (`mechanism_class_4()` on the
ref-AUG-projected ORF category + ORF-length comparison):

```{r sec3-class-table, echo=FALSE}
knitr::kable(sec3_class,
             caption = "Section C NMD-arm 3-group classification",
             col.names = c("Group", "n"),
             format.args = list(big.mark = ","))

# ── Claim 4.7.4 — PTC+ rate at the reference AUG-traceable scope ──────────────────────
n_ptc_pos_C <- sec3_class$n[sec3_class$group == "NMD+/PTC+"]
n_secC      <- nrow(refaug_matched_c2)
claim_emit("4.7.4", "NMD isoforms with an identified PTC (count)", n_ptc_pos_C,
           published = 768, n = n_secC, population = SEC_C_POPULATION)
claim_emit("4.7.4", "NMD isoforms in the reference AUG-traceable set (denominator)", n_secC,
           published = 833, n = n_secC, population = SEC_C_POPULATION)
claim_emit("4.7.4", "share of NMD isoforms with an identified PTC (%)",
           100 * n_ptc_pos_C / n_secC, published = 92, n = n_secC,
           population = SEC_C_POPULATION)
```

PTC+ rate at this scope: `r format(sec3_class$n[sec3_class$group == "NMD+/PTC+"], big.mark=",")` /
`r format(nrow(refaug_matched_c2), big.mark=",")` =
`r round(100 * sec3_class$n[sec3_class$group == "NMD+/PTC+"] / nrow(refaug_matched_c2), 1)`%.
This is higher than Section A's PTC+ rate because the classifier definitions
differ — see §5 for reconciliation.

### Section C CDS / 3'UTR feature export (SF33 / SF36)

Migrated here from `figures/supplemental/SF33_CdsAnd3UTR_GENCODE/data_export.R`
on 2026-07-28 (S13: no result may originate in `figures/`). That script did not
merely export — it **rebuilt the Section C cohort itself**, repeating the
`sec3-scope` construction and the `mechanism_class → group3` classification, so
one population was defined in two places and only one of them could be checked
by re-rendering this report (S4). Claim 4.7.5's operative half — that the 3'UTR
effect is limited to the ref-AUG-traceable set — was produced entirely outside
the report layer.

**Proven output-preserving before the move.** Both constructions were run against
the same tree and compared: the same 833/833 cohort with an identical key set, an
identical 1,666-row feature table, and identical descriptives and pairwise tables
for all three measures. The `n = 1,166` in the old script's header was a stale
legacy-tree vintage, not a different scope.


```{r sec3d-utr3-tables, echo=FALSE}
knitr::kable(secC_utr3_desc, caption = "Section C 3'UTR length, translation-based (ref-AUG-projected stop → tx end)")
knitr::kable(secC_utr3np_desc, caption = "Section C 3'UTR length, bias-corrected (non-PTC-stop based)")
```

Claim 4.7.5 asserts the 3'UTR effect is *limited to* the ref-AUG-traceable set.
Read against both scopes, that holds under the **bias-corrected** measure and
fails under the **translation-based** one, so the sentence needs its measure named.

```{r sec3d-utr3-4-7-5, echo=FALSE}
knitr::kable(
  rbind(secC_utr3_pw[, .(measure = "translation-based", group_x, group_y,
                         median_x, median_y, wilcox_p)],
        secC_utr3np_pw[, .(measure = "bias-corrected", group_x, group_y,
                           median_x, median_y, wilcox_p)]),
  caption = "Section C 3'UTR pairwise tests — the comparison claim 4.7.5 rests on")

# ── Claim 4.7.5 — "limited to the Reference AUG-traceable set only" ────────────────────
# A CONTRAST ACROSS TWO SCOPES, so BOTH have to be emitted or the word "limited" is unchecked:
# the claim fails if the effect is also present in the GENCODE reference set. It is also
# measure-dependent and the sentence names no measure, which is the defect on record -- so both
# the translation-based and the bias-corrected test are emitted, for both scopes, four rows.
# All supporting: the paper prints no number here.
# BASE-R SUBSETTING, DELIBERATELY: Section A's pairwise table is a data.frame (from
# pairwise_stats) and Section C's is a data.table (from fread), so data.table's
# unquoted-column syntax works on one and raises "object 'group_x' not found" on the other.
# Measured -- it halted the knit at this chunk after 56 emissions had already been written.
utr3_p <- function(pw) {
  i <- which(pw$group_x == "NMD+/PTC+" & pw$group_y == "Control")
  if (!length(i)) i <- which(pw$group_x == "Control" & pw$group_y == "NMD+/PTC+")
  if (!length(i)) NA_real_ else as.numeric(pw$wilcox_p[i[1]])
}
MEASURE_DEF <- c(
  "translation-based" = paste("3'UTR measured from the isoform's own stop codon to the",
                              "transcript end, so a PTC inflates it by construction"),
  "bias-corrected"    = paste("3'UTR measured from the first NON-PTC stop codon, walking past",
                              "a PTC, which removes that inflation"))
for (spec in list(
      list(pw = secA_utr3_pw,   scope = "GENCODE reference set",       m = "translation-based",
           pop = SEC_A_POPULATION),
      list(pw = secA_utr3np_pw, scope = "GENCODE reference set",       m = "bias-corrected",
           pop = SEC_A_POPULATION),
      list(pw = secC_utr3_pw,   scope = "reference AUG-traceable set", m = "translation-based",
           pop = SEC_C_POPULATION),
      list(pw = secC_utr3np_pw, scope = "reference AUG-traceable set", m = "bias-corrected",
           pop = SEC_C_POPULATION))) {
  claim_emit("4.7.5",
             sprintf("Wilcoxon p, 3'UTR length PTC+ vs Control, %s measure, %s",
                     spec$m, spec$scope),
             utr3_p(spec$pw),
             population = paste0(spec$pop, ". ", MEASURE_DEF[[spec$m]]))
}
```

## §3a — 5'UTR length + longest 5'UTR ORF at Section C

At the n=819 scope, most NMD comparators are novel (TD2-anchored CDS not
reliable for 5'UTR measurement) — so we use `utr5_features_refaug`, which is
the 5'UTR delimited by the ref-AUG-projected start. Three groups (from §3
classification): NMD+/PTC+, NMD+/PTC− (ORF match + ORF diff merged), Control.

```{r sec3a-secC-groups, include=FALSE}
sec_C_groups <- list(
  "NMD+/PTC+" = refaug_matched_c2[group3 == "NMD+/PTC+", comparator_isoform_id],
  "NMD+/PTC-" = refaug_matched_c2[group3 == "NMD+/PTC-", comparator_isoform_id],
  "Control"   = refaug_matched_c4$comparator_isoform_id
)

u5_refaug_dt <- as.data.table(utr5_refaug$isoform_features)
u5_len_ra <- setNames(as.numeric(u5_refaug_dt$utr5_length),    u5_refaug_dt$isoform_id)
u5_orf_ra <- setNames(as.numeric(u5_refaug_dt$longest_orf_nt), u5_refaug_dt$isoform_id)

secC_utr5_desc <- group_descriptives(sec_C_groups, u5_len_ra,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secC_utr5_pw   <- pairwise_stats(sec_C_groups, u5_len_ra)
secC_orf_desc  <- group_descriptives(sec_C_groups, u5_orf_ra,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secC_orf_pw    <- pairwise_stats(sec_C_groups, u5_orf_ra)
secC_utr5_pw$wilcox_p_fmt <- vapply(secC_utr5_pw$wilcox_p, format_p, character(1))
secC_orf_pw$wilcox_p_fmt  <- vapply(secC_orf_pw$wilcox_p,  format_p, character(1))
```

### Panel C — 5'UTR length (ref-AUG-anchored)

```{r sec3a-panelC-tables, echo=FALSE}
knitr::kable(secC_utr5_desc, caption = "5'UTR length (nt) — Section C descriptives",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secC_utr5_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "5'UTR length — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### Panel D — Longest 5'UTR ORF (ref-AUG-anchored)

```{r sec3a-panelD-tables, echo=FALSE}
knitr::kable(secC_orf_desc, caption = "Longest 5'UTR ORF (nt) — Section C descriptives",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secC_orf_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "Longest 5'UTR ORF — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

```{r sec3a-fig4CD, echo=FALSE, fig.cap=c("Figure 4 Panel C — 5'UTR length at n=819 scope", "Figure 4 Panel D — Longest 5'UTR ORF at n=819 scope"), out.width="90%"}
inc_fig(c(
  file.path(FIG_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelC_5utr_length_refaug.png"),
  file.path(FIG_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelD_longest_5utr_orf_refaug.png")
))
```

## §3b — PTC rate at Section C

Already computed inline in the §3 scope construction:
`r format(sec3_class$n[sec3_class$group == "NMD+/PTC+"], big.mark=",")` PTC+ /
`r format(nrow(refaug_matched_c2), big.mark=",")` =
**`r round(100 * sec3_class$n[sec3_class$group == "NMD+/PTC+"] / nrow(refaug_matched_c2), 1)`%** at this scope (vs 37.9% at the n=130
all-3-ENST scope). The classifier definitions differ —
see §5 for reconciliation.

# §4 — TD2 bias evidence (TD2BiasEvidence supplement)

Scope (broad): `refaug_matched_c2 ∩ TD2 CDS available ∩ ref-AUG ORF computable
∩ ref AUG exonic` → n=819 broad.
Scope (occult-PTC): broad `∩ effectively_ptc ∩ original_ptc == FALSE` → n=348.

**Scope construction.** Broad (n=819) = `refaug_matched_c2` ∩ TD2 CDS
available ∩ ref-AUG ORF computable ∩ ref AUG exonic. Occult-PTC (n=348) =
broad ∩ `category == "effectively_ptc"` ∩ `original_ptc == FALSE` — pairs
where ref-AUG tracing revealed a PTC that TD2 missed. This is the population
in which TD2's PTC-avoidance bias is most pronounced.

```{r sec4-scope, include=FALSE}
c2_full_dt <- as.data.table(refaug_matched_c2)
comp_cds   <- cds[match(c2_full_dt$comparator_isoform_id, isoform_id)]
c2_full_dt[, strand_use      := comp_cds$strand]
c2_full_dt[, td2_atg_genomic := ifelse(comp_cds$strand == "+",
                                       comp_cds$cds_start, comp_cds$cds_stop)]
c2_full_dt[, td2_orf_nt      := comp_cds$orf_length * 3L]
c2_full_dt[, ref_atg_orf_nt  := comp_orf_length_from_ref_atg]

broad <- c2_full_dt[!is.na(td2_orf_nt) & !is.na(ref_atg_orf_nt) &
                     ref_atg_exonic_in_comp == TRUE]
```

`broad` = `r format(nrow(broad), big.mark=",")` pairs (matches the
n=819 TD2BiasEvidence supplement scope).

```{r sec4-compute, include=FALSE}
# CACHING REMOVED 2026-07-30 (Pete's call, W134). This chunk was cache=TRUE with
# cache.extra=tools::md5sum(FASTA_PATH). Three reasons it had to go, in order of severity:
#
# 1. The cache directory was TRACKED IN GIT, so it shipped. .gitignore had no cache rule, and a
#    git add on the directory swept it in at 6332ffd. D1 forecloses intermediate data in the
#    citable repo, and a knitr cache is intermediate data by any reading.
# 2. cache.extra covered ONLY the FASTA. This chunk also reads the isopair intermediates, so the
#    key could not notice them changing -- which is exactly the case a deposit-native rebuild
#    creates. Anyone who did hit the cache would see section 4 "reproduce" regardless of what
#    their own intermediates contained.
# 3. tools::md5sum() returns a PATH-NAMED vector, so the key silently included the deposit's
#    ABSOLUTE PATH. That makes the cache non-portable on principle: it can only ever hit on the
#    machine that wrote it, so it could never do its job in a repo meant for readers.
#
# The saving was never measured. The whole report renders in 42 s uncached (stage 3d, C74), so
# there was nothing to protect.
genomic_to_tx <- function(gp, es, ee, s) {
  if (s == "-") { es <- rev(es); ee <- rev(ee) }
  cum <- 0L
  for (j in seq_along(es)) {
    if (s == "+" && gp >= es[j] && gp <= ee[j]) return(cum + (gp - es[j]) + 1L)
    if (s == "-" && gp >= es[j] && gp <= ee[j]) return(cum + (ee[j] - gp) + 1L)
    cum <- cum + (ee[j] - es[j] + 1L)
  }
  NA_integer_
}

bs <- structures[match(broad$comparator_isoform_id, isoform_id)]
ref_tx <- integer(nrow(broad)); td2_tx <- integer(nrow(broad))
for (i in seq_len(nrow(broad))) {
  es <- bs$exon_starts[[i]]; ee <- bs$exon_ends[[i]]
  s  <- broad$strand_use[i]
  if (length(es) == 0L || is.na(broad$ref_atg_genomic[i]) ||
      is.na(broad$td2_atg_genomic[i])) {
    ref_tx[i] <- NA_integer_; td2_tx[i] <- NA_integer_; next
  }
  ref_tx[i] <- genomic_to_tx(broad$ref_atg_genomic[i], es, ee, s)
  td2_tx[i] <- genomic_to_tx(broad$td2_atg_genomic[i], es, ee, s)
}
broad[, ref_atg_tx_pos := ref_tx]
broad[, td2_atg_tx_pos := td2_tx]
broad[, distance_nt   := td2_atg_tx_pos - ref_atg_tx_pos]
broad_clean <- broad[!is.na(ref_atg_tx_pos) & !is.na(td2_atg_tx_pos)]

# Kozak PWM scoring — load sequences for broad_clean comparators from the
# nmd_lungcells corrected FASTA via awk (avoid full Biostrings dependency).
load_sequences <- function(isoform_ids, fasta_path) {
  ids <- unique(isoform_ids)
  id_file <- tempfile(fileext = ".txt")
  writeLines(ids, id_file); on.exit(unlink(id_file), add = TRUE)
  awk_cmd <- sprintf(
    "awk 'BEGIN{while((getline id < \"%s\") > 0) want[id]=1} /^>/{if(seq && found) print hdr \"\\t\" seq; hdr=substr($1,2); found=(hdr in want); seq=\"\"; next} found{seq=seq$0} END{if(seq && found) print hdr \"\\t\" seq}' '%s'",
    id_file, fasta_path)
  raw <- system(awk_cmd, intern = TRUE)
  split_raw <- strsplit(raw, "\t", fixed = TRUE)
  setNames(toupper(vapply(split_raw, `[`, character(1), 2)),
                   vapply(split_raw, `[`, character(1), 1))
}
seqs <- load_sequences(broad_clean$comparator_isoform_id, FASTA_PATH)
pwm_seqs <- character(nrow(broad_clean))
for (k in seq_len(nrow(broad_clean))) {
  cid <- broad_clean$comparator_isoform_id[k]
  if (cid %in% names(seqs)) pwm_seqs[k] <- seqs[cid]
}
v_ref <- !is.na(broad_clean$ref_atg_tx_pos) & pwm_seqs != ""
v_td2 <- !is.na(broad_clean$td2_atg_tx_pos) & pwm_seqs != ""
ref_pwm <- rep(NA_real_, nrow(broad_clean)); td2_pwm <- rep(NA_real_, nrow(broad_clean))
if (sum(v_ref) > 0) {
  r <- Isopair::scoreKozakPWM(pwm_seqs[v_ref], broad_clean$ref_atg_tx_pos[v_ref])
  ref_pwm[v_ref] <- r$score
}
if (sum(v_td2) > 0) {
  r <- Isopair::scoreKozakPWM(pwm_seqs[v_td2], broad_clean$td2_atg_tx_pos[v_td2])
  td2_pwm[v_td2] <- r$score
}
broad_clean[, ref_kozak := ref_pwm]
broad_clean[, td2_kozak := td2_pwm]

# Broad-scope summary stats — ratio convention TD2/ref-AUG (matches the
# TD2BiasEvidence supplement). For PTC+ pairs TD2 picks a downstream ATG
# that extends past the PTC, so TD2 ORF > ref-AUG ORF on average.
broad_orf_ratio <- median(broad_clean$td2_orf_nt) / median(broad_clean$ref_atg_orf_nt)
broad_koz_valid  <- !is.na(broad_clean$ref_kozak) & !is.na(broad_clean$td2_kozak)
broad_n_kozak    <- sum(broad_koz_valid)
broad_n_ref_gt   <- sum(broad_clean$ref_kozak[broad_koz_valid] >
                         broad_clean$td2_kozak[broad_koz_valid])
broad_n_ref_eq   <- sum(broad_clean$ref_kozak[broad_koz_valid] ==
                         broad_clean$td2_kozak[broad_koz_valid])
broad_n_nontied  <- broad_n_kozak - broad_n_ref_eq
broad_koz_w      <- wilcox.test(broad_clean$ref_kozak[broad_koz_valid],
                                 broad_clean$td2_kozak[broad_koz_valid],
                                 paired = TRUE)
broad_pos_eq    <- sum(broad_clean$distance_nt == 0L)
broad_pos_down  <- sum(broad_clean$distance_nt > 0L)
broad_pos_up    <- sum(broad_clean$distance_nt < 0L)
broad_pos_pct_down <- 100 * broad_pos_down / nrow(broad_clean)

# Occult-PTC subset
occ <- broad_clean[category == "effectively_ptc" & original_ptc == FALSE]
occ_orf_ratio <- median(occ$td2_orf_nt) / median(occ$ref_atg_orf_nt)
occ_koz_valid <- !is.na(occ$ref_kozak) & !is.na(occ$td2_kozak)
occ_n_kozak   <- sum(occ_koz_valid)
occ_n_ref_gt  <- sum(occ$ref_kozak[occ_koz_valid] > occ$td2_kozak[occ_koz_valid])
occ_n_ref_eq  <- sum(occ$ref_kozak[occ_koz_valid] == occ$td2_kozak[occ_koz_valid])
occ_n_nontied <- occ_n_kozak - occ_n_ref_eq
occ_koz_w     <- wilcox.test(occ$ref_kozak[occ_koz_valid],
                              occ$td2_kozak[occ_koz_valid], paired = TRUE)
occ_pos_down  <- sum(occ$distance_nt > 0L)
occ_pos_pct_down <- 100 * occ_pos_down / nrow(occ)
```

## §4a — TD2 vs ref-AUG ORF length

| Scope | n | TD2 median ORF (nt) | ref-AUG median ORF (nt) | TD2/ref ratio |
|---|---|---|---|---|
| Broad | `r format(nrow(broad_clean), big.mark=",")` | `r format(median(broad_clean$td2_orf_nt), big.mark=",")` | `r format(median(broad_clean$ref_atg_orf_nt), big.mark=",")` | **`r sprintf("%.2f×", broad_orf_ratio)`** |
| Occult-PTC | `r format(nrow(occ), big.mark=",")` | `r format(median(occ$td2_orf_nt), big.mark=",")` | `r format(median(occ$ref_atg_orf_nt), big.mark=",")` | **`r sprintf("%.2f×", occ_orf_ratio)`** |

## §4b — Paired Kozak PWM at ref AUG vs TD2 ATG

The "ref > TD2 (% of all)" column counts pairs where the ref-AUG Kozak score
exceeds TD2's — pooled over all pairs including ties (broad scope has many
pairs where TD2 == ref AUG, distance 0). The "% of non-tied" column drops
ties so the directional preference is unobscured. The paired-Wilcoxon p-value
is computed on the raw score vector, ignoring ties as `wilcox.test` does.

| Scope | n | ref > TD2 | ties | ref > TD2 (% of all) | ref > TD2 (% of non-tied) | Paired-Wilcoxon p |
|---|---|---|---|---|---|---|
| Broad | `r format(broad_n_kozak, big.mark=",")` | `r format(broad_n_ref_gt, big.mark=",")` | `r format(broad_n_ref_eq, big.mark=",")` | `r sprintf("%.1f%%", 100 * broad_n_ref_gt / broad_n_kozak)` | `r sprintf("%.1f%%", 100 * broad_n_ref_gt / broad_n_nontied)` | `r format_p(broad_koz_w$p.value)` |
| Occult-PTC | `r format(occ_n_kozak, big.mark=",")` | `r format(occ_n_ref_gt, big.mark=",")` | `r format(occ_n_ref_eq, big.mark=",")` | `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_kozak)` | `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_nontied)` | `r format_p(occ_koz_w$p.value)` |

## §4c — TD2 ATG position relative to ref AUG

| Scope | n | TD2 == ref | TD2 downstream | TD2 upstream | % downstream |
|---|---|---|---|---|---|
| Broad | `r format(nrow(broad_clean), big.mark=",")` | `r format(broad_pos_eq, big.mark=",")` | `r format(broad_pos_down, big.mark=",")` | `r format(broad_pos_up, big.mark=",")` | **`r sprintf("%.1f%%", broad_pos_pct_down)`** |
| Occult-PTC | `r format(nrow(occ), big.mark=",")` | `r format(sum(occ$distance_nt == 0L), big.mark=",")` | `r format(occ_pos_down, big.mark=",")` | `r format(sum(occ$distance_nt < 0L), big.mark=",")` | **`r sprintf("%.1f%%", occ_pos_pct_down)`** |

In the occult-PTC scope, **TD2 selects an ATG downstream of the ref-AUG in
`r sprintf("%.1f%%", occ_pos_pct_down)` of pairs**, with a Kozak score lower
than the ref-AUG in `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_kozak)`
of pairs (paired-Wilcoxon `r format_p(occ_koz_w$p.value)`). The TD2 ORF is
`r sprintf("%.2f×", occ_orf_ratio)` LONGER than the ref-AUG-projected ORF
(ratio of medians) — TD2 picks a downstream ATG that extends past the PTC into an
in-frame coding region, producing a longer ORF that lacks the EJC-downstream
stop that defines a PTC. Together these three signals describe TD2's
PTC-avoidance bias: TD2 systematically routes around occult PTCs by picking a
downstream ATG, even at the cost of a lower Kozak score.

```{r sec4-supp, echo=FALSE, fig.cap="TD2BiasEvidence supplement — 2×3 layout: broad scope (row 1, n=819) + occult-PTC scope (row 2, n=348)", out.width="100%"}
inc_fig(
  c(file.path(FIG_ROOT, "figures/supplemental/SF34_TD2Bias_broad/figure_sf34.png"),
    file.path(FIG_ROOT, "figures/supplemental/SF35_TD2Bias_occult/figure_sf35.png"))
)
```

# §5 — Cumulative accounting [verification chokepoint]

Recompute from §2 / §3 result objects so the totals in this report can be
audited end-to-end without trusting upstream caches.

```{r sec5-accounting, include=FALSE}
secA_ptc_pos  <- length(ptc_pos_ids)
secA_ptc_neg  <- length(ptc_neg_ids)
secA_total    <- nrow(gencode_all3_c2)
stopifnot(
  "§2a NMD denominator drift"   = n_avail_nmd  == secA_total,
  "§2a Control denominator drift" = n_avail_ctrl == nrow(gencode_all3_c4),
  "§2a PTC+ + PTC- != Section A total" = secA_ptc_pos + secA_ptc_neg == secA_total,
  "§2b mechanism table sum != unique direct attributions" =
    sum(mech_tab$n) == n_unique_direct
)

get_n <- function(df, g) {
  v <- df$n[df$group == g]
  if (length(v) == 0L) 0L else v
}
secC_ptc_pos  <- get_n(sec3_class, "NMD+/PTC+")
secC_ptc_neg  <- get_n(sec3_class, "NMD+/PTC-")
secC_ref_abs  <- get_n(sec3_class, "Ref AUG absent")
secC_total    <- nrow(refaug_matched_c2)
stopifnot(
  "§3 class sum != Section C total" =
    secC_ptc_pos + secC_ptc_neg + secC_ref_abs == secC_total,
  "§3a Control denominator != Section C total" =
    length(sec_C_groups$Control) == secC_total
)

acct <- data.frame(
  Scope = c("Section A (n=130)", "Section C (n=819)"),
  PTC_pos = c(secA_ptc_pos, secC_ptc_pos),
  PTC_neg = c(secA_ptc_neg, secC_ptc_neg),
  Ref_AUG_absent = c(NA_integer_, secC_ref_abs),
  Total = c(secA_total, secC_total),
  PTC_pos_rate_pct = c(round(100 * secA_ptc_pos / secA_total, 1),
                       round(100 * secC_ptc_pos / secC_total, 1))
)
```

```{r sec5-acct-table, echo=FALSE}
knitr::kable(acct,
             caption = "Cumulative accounting at both canonical scopes",
             col.names = c("Scope", "PTC+", "PTC−", "Ref AUG absent",
                           "Total", "PTC+ rate (%)"))
```

- **Section A (n=`r secA_total`):** `r secA_ptc_pos` PTC+ and `r secA_ptc_neg` PTC− sum to `r secA_total`. PTC+ rate **`r round(100 * secA_ptc_pos / secA_total, 1)`%**.
- **Section C (n=`r format(secC_total, big.mark=",")`):** `r format(secC_ptc_pos, big.mark=",")` PTC+, `r secC_ptc_neg` PTC−, and `r secC_ref_abs` Ref-AUG-absent sum to `r format(secC_total, big.mark=",")`. PTC+ rate **`r round(100 * secC_ptc_pos / secC_total, 1)`%**.

**Why the two scopes give different PTC+ rates** is a methodological — not
biological — divergence:

| Difference | Section A | Section C |
|---|---|---|
| Cohort filter | all-3-ENST + coding-CDS (most stringent) | ENST-reference only (broader) |
| PTC classifier | own-GENCODE-stop + 50-nt rule | ref-AUG-projected `category == "effectively_ptc"` |
| Comparator CDS source | curated GENCODE | TD2 (heavily debiased via ref-AUG projection) |
| Selection-on-PTC | none — every coding-CDS comparator measured | TR-category filter requires ref-AUG-traceable ORF — pairs where ref-AUG points at "no_downstream_ejc" or "truncated_no_ejc" classify as PTC− regardless of comparator's own stop |

The Section C 90% PTC+ rate is enriched by the TR-category filter, which
removes pairs where ref-AUG tracing finds no PTC and no stop near the EJC.
The Section A 38% PTC+ rate has no such pre-filter: it's the raw PTC rate
across all GENCODE-only NMD pairs. The two rates are not directly
comparable. The point of measuring both is to show that the §2-§4 5'UTR /
ORF / Kozak signals replicate at both scopes.

# §6 — What is intentionally NOT in this Rmd

These analyses remain in `05_final_report_mashr.Rmd` (legacy-bannered) or
`06_orf_analysis_mashr.Rmd` (legacy-bannered) — they are retained for
reproducibility of the prior submission and for sensitivity analyses:

- Reclassification two-pass (TD2 → ref-AUG handoff at the old scope) — legacy
  `05_final_report_mashr.Rmd`.
- Per-cell-type PTC stratification at the old scope — legacy
  `05_final_report_mashr.Rmd`.
- "85% combined" framework — legacy `05_final_report_mashr.Rmd`.
- n=1,904 / n=22,335 GENCODE-only 3'UTR outside-set — legacy
  `05_final_report_mashr.Rmd`.
- ORF / deep-learning model analyses — `06_orf_analysis_mashr.Rmd` (in this
  repo, legacy). The canonical home of the manuscript deep-learning model is
  the **linked repo `peter4244/NMD_orf_model_v5_4ct`**: `03_train.py`,
  `config.yaml`, `model.py`, `METHODS.md`, plus DeepSHAP / KernelSHAP scripts.
  The "ORF identification is the bottleneck" headline result lives there.

Cross-cell-type consistency checks are out of scope for this Rmd; per-CT
mashr CSVs in `shortread_dge/mashr/` and `isocall_dge/mashr/` are the source
of truth for CT-stratified numbers.
