---
title: "Short-Read Gene Vs Long-Read Isoform: NMD Response Comparison"
author: "Leshem, Yul"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
.nmd_root <- local({ d <- getwd(); while (!file.exists(file.path(d, "config", "paths.yml"))) {
  p <- dirname(d); if (identical(p, d)) stop("repo root not found"); d <- p }; d })
source(file.path(.nmd_root, "R", "load_config.R")); .P <- nmd_paths(.nmd_root)
source(file.path(.nmd_root, "R", "claim_emit.R"))  # Phase 2 claim-value emission
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dev = "png")
options(bitmapType = "cairo")

```

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/comparison_analysis.Rmd`

# Overview

Two-pipeline comparison of NMD susceptible gene calls across the four kept
cell types (AT2, LAE, FB, MV), restricted to short-read gene-level DGE and
long-read isoform-level DIE. The long-read gene-level DGE pipeline is
intentionally excluded here.

For long-read isoforms we need a single per-gene logFC summary. We use a
significance-weighted mean of isoform-level posterior means, with weights
`-log10(lfsr)` clipped at the analysis threshold so that only confidently
called isoforms contribute. A gene is considered NMD susceptible in a cell
type if **any** of its isoforms is significant with positive logFC.

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(patchwork)
  library(scales)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice
first  <- dplyr::first
```

# Configuration

```{r config}
OUTPUT_DIR <- file.path(.P$OUT, "comparison")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Source directories for the two mashr result sets
SR_DGE_DIR <- file.path(.P$OUT, "mashr_gene")
DIE_DIR    <- file.path(.P$OUT, "mashr_isoform")
SR_PATTERN  <- "^nmd_mashr_dge_.*_2026\\.3\\.10\\.csv$"
DIE_PATTERN <- "^nmd_mashr_die_.*_2026\\.3\\.10\\.csv$"

# Four cell types — DD_ALI and DO_ALI are cut from the paper.
CT_KEEP  <- c("AT2", "LAE", "FB", "MV")
CT_ORDER <- CT_KEEP

# Cell-type code (lowercase, from filename) -> display name.
# Pipeline writes nmd_mashr_*_at2/lae/fb/mv, so the keys are at2/lae/fb/mv.
CELLTYPE_MAP <- c("at2" = "AT2", "lae" = "LAE", "fb" = "FB", "mv" = "MV")

LFSR_THRESH <- 0.05
LOGFC_THRESH <- 0

# ── Claim 2.8.3 — the Figure 1C definition of NMD susceptibility ──────────────────────
# The two published numbers here are THRESHOLDS, not results: "posterior mean logFC > 0 and
# lfsr < 0.05". Emitting them against themselves would be a check that cannot fail, so they are
# recorded as SUPPORTING values naming the constants the whole section depends on. What can
# actually be verified is that the definition is APPLIED consistently, and that is asserted
# where the feature tables are built rather than claimed here.
claim_emit("2.8.3", "lfsr threshold defining NMD susceptibility", LFSR_THRESH,
           population = "the constant applied to every NMD susceptibility call in this report")
claim_emit("2.8.3", "posterior mean logFC threshold defining NMD susceptibility", LOGFC_THRESH,
           population = "the constant applied to every NMD susceptibility call in this report")

```

# Load mashr Results

```{r load-results}
# Helper: read one per-cell-type result CSV and tag with display CT name
read_results <- function(path, ct_extractor, id_candidates) {
  ct_code <- ct_extractor(basename(path))
  ct_name <- CELLTYPE_MAP[ct_code]
  if (is.na(ct_name) || !ct_name %in% CT_KEEP) return(NULL)

  df <- fread(path)
  id_col <- intersect(names(df), id_candidates)
  if (length(id_col) == 0) id_col <- names(df)[1]

  # RETAIN gene_id WHEN THE SOURCE CARRIES IT (added 2026-07-27, §2 tree gate).
  #
  # This used to select(ct, .id_col, logFC, adj.P.Val) unconditionally, discarding a
  # `gene_id` column that the long-read die CSVs DO carry (alongside hgnc_symbol). The
  # consequence was that 2.1.4's "7,105 unique genes" and 2.1.5's 2,326 / 4,779 could not be
  # emitted by this report at all, and were recomputed by hand instead.
  #
  # The obvious workaround -- map isoforms to genes through `iso_to_gene` -- is WRONG here and
  # was measured to be so: gmap is Ensembl-derived, most NMD isoforms are novel, and only
  # 4,540 of 34,387 (13.2%) join. That route gives 2,595 genes against the published 7,105.
  # The die output's own SQANTI-derived gene_id is the right source, which is also what the
  # W41 finding concluded independently.
  #
  # Guarded on presence and on not colliding with .id_col: for the SR side the id column IS
  # gene_id, so an unconditional keep would duplicate it.
  keep <- c("ct", ".id_col", "logFC", "adj.P.Val")
  out <- df %>%
    rename(.id_col = !!id_col[1]) %>%
    mutate(ct = ct_name)
  if ("gene_id" %in% names(out) && id_col[1] != "gene_id") keep <- c(keep, "gene_id")
  out %>% select(all_of(keep))
}

# ---- SR gene-level DGE ----
sr_files <- list.files(SR_DGE_DIR, pattern = SR_PATTERN, full.names = TRUE)
cat("FILTER: SR files matching pattern:", length(sr_files), "\n"); print(basename(sr_files))
sr_dge_all <- lapply(
  sr_files, read_results,
  ct_extractor  = function(fn) sub("^nmd_mashr_dge_(.+)_2026.*", "\\1", fn),
  id_candidates = c("gene_id", "gene", "V1")
) %>%
  bind_rows() %>%
  rename(gene_id = .id_col) %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id))

# ---- LR isoform-level DIE ----
die_files <- list.files(DIE_DIR, pattern = DIE_PATTERN, full.names = TRUE)
cat("FILTER: DIE files matching pattern:", length(die_files), "\n"); print(basename(die_files))
lr_die_all <- lapply(
  die_files, read_results,
  ct_extractor  = function(fn) sub("^nmd_mashr_die_(.+)_2026.*", "\\1", fn),
  id_candidates = c("txid", "isoform", "V1", "gene")
) %>%
  bind_rows() %>%
  rename(txid = .id_col)

stopifnot("No SR rows loaded — check CELLTYPE_MAP vs filenames"  = nrow(sr_dge_all) > 0)
stopifnot("No LR rows loaded — check CELLTYPE_MAP vs filenames"  = nrow(lr_die_all) > 0)

cat("SR DGE rows:", nrow(sr_dge_all), "\n")
cat("LR DIE rows:", nrow(lr_die_all), "\n")
cat("\nPer-CT row counts:\n")
cat("  SR DGE: "); print(table(sr_dge_all$ct))
cat("  LR DIE: "); print(table(lr_die_all$ct))
```

# Build Transcript → Gene Mapping

The LR isoform results are keyed by txid. For both the logFC aggregation
and the gene-level sharing analysis, we need a txid → gene_id link. Source
is the project gmap RDS, with a fallback to the long-read DGEList.

```{r load-iso-to-gene}
GMAP_PATH        <- file.path(.P$ANNOT, "gmap_txlevel_ENSGv115_2025.08.12.rds")
DGELIST_ISO_PATH <- file.path(.P$OUT, "dge_isoform_longread_2026.3.3.rds")

if (file.exists(GMAP_PATH)) {
  gmap <- readRDS(GMAP_PATH)
  tx_col_candidates <- c("ensembl_transcript_id_version",
                         "ensembl_transcript_id",
                         "transcript_id_version",
                         "transcript_id", "txid")
  tx_col <- intersect(tx_col_candidates, colnames(gmap))[1]
  if (is.na(tx_col)) stop("No transcript-ID column found in gmap")

  iso_to_gene <- unique(data.frame(
    txid        = gmap[[tx_col]],
    gene_id     = gmap$ensembl_gene_id_version,
    hgnc_symbol = gmap$hgnc_symbol,
    stringsAsFactors = FALSE
  ))
  cat("iso_to_gene from gmap:", nrow(iso_to_gene), "rows\n")

} else if (file.exists(DGELIST_ISO_PATH)) {
  dge_iso <- readRDS(DGELIST_ISO_PATH)
  iso_to_gene <- unique(dge_iso$genes[, c("txid", "gene_id", "hgnc_symbol")])
  cat("iso_to_gene from long-read DGEList:", nrow(iso_to_gene), "rows\n")

} else {
  stop("Neither gmap RDS nor long-read DGEList found — cannot build iso_to_gene")
}

iso_to_gene <- iso_to_gene %>%
  filter(!is.na(gene_id), gene_id != "") %>%
  mutate(gene_id_clean = sub("\\..*", "", gene_id)) %>%
  distinct(txid, .keep_all = TRUE)

cat("After cleanup:", nrow(iso_to_gene), "rows\n")

```

# Aggregate LR Isoform logFC To Gene Level

For each gene in each cell type, compute a significance-weighted mean
isoform logFC:

$$
\bar{\beta}_{g,c} = \frac{\sum_{i \in g} w_{i,c} \, \beta_{i,c}}{\sum_{i \in g} w_{i,c}}, \quad
w_{i,c} = \max\!\left(-\log_{10}(\text{lfsr}_{i,c}),\, \epsilon\right)
$$

The floor $\epsilon$ on the weight prevents non-significant isoforms from
contributing zero weight (which would crash the divide). It's set to a
small positive value so high-significance isoforms still dominate.

```{r aggregate-lr-iso-to-gene}
EPS <- 1e-3  # weight floor

n_before <- nrow(lr_die_all)
lr_iso_with_gene <- lr_die_all %>%
  inner_join(iso_to_gene %>% select(txid, gene_id_clean),
             by = "txid") %>%
  mutate(weight = pmax(-log10(adj.P.Val), EPS))
cat(sprintf("FILTER: isoforms joined to a gene_id.\n  Rows before: %d / after join: %d (dropped %d unmapped).\n",
            n_before, nrow(lr_iso_with_gene), n_before - nrow(lr_iso_with_gene)))

lr_iso_gene <- lr_iso_with_gene %>%
  group_by(ct, gene_id_clean) %>%
  summarise(logFC      = sum(weight * logFC) / sum(weight),
            min_lfsr   = min(adj.P.Val, na.rm = TRUE),
            n_isoforms = n(),
            .groups = "drop")

cat("Per-cell-type gene counts after aggregation:\n")
print(table(lr_iso_gene$ct))
```

# Combined Long-Format Table For Plotting

```{r build-long-format}
sr_long <- sr_dge_all %>%
  mutate(analysis = "Short-Read Gene") %>%
  select(analysis, ct, gene_id_clean, logFC, lfsr = adj.P.Val)

lr_long <- lr_iso_gene %>%
  mutate(analysis = "Long-Read Isoform") %>%
  select(analysis, ct, gene_id_clean, logFC, lfsr = min_lfsr)

combined <- bind_rows(sr_long, lr_long) %>%
  mutate(analysis = factor(analysis,
                           levels = c("Short-Read Gene", "Long-Read Isoform")),
         ct       = factor(ct, levels = CT_ORDER),
         is_sig   = lfsr < LFSR_THRESH,
         is_nmd   = is_sig & logFC > LOGFC_THRESH)

cat("Combined table:", nrow(combined), "rows\n")
cat("\nNMD susceptible counts (lfsr <", LFSR_THRESH, "& logFC >", LOGFC_THRESH, "):\n")
combined %>%
  group_by(analysis, ct) %>%
  summarise(n_total = n(),
            n_nmd   = sum(is_nmd, na.rm = TRUE),
            .groups = "drop") %>%
  print()

```

# logFC Distribution Of NMD Susceptible Genes

Boxplots of posterior-mean logFC restricted to NMD susceptible genes
(lfsr < 0.05 & logFC > 0), faceted by cell type, dodged by analysis.

```{r logfc-distribution, fig.width=10, fig.height=6}
nmd_only <- combined %>% filter(is_nmd)

# Median logFC per (analysis, ct) for the dashed reference line
median_logfc <- nmd_only %>%
  group_by(analysis, ct) %>%
  summarise(median_logFC = median(logFC), .groups = "drop")
overall_median <- median(nmd_only$logFC)

p_logfc <- ggplot(nmd_only, aes(x = ct, y = logFC, fill = analysis)) +
  geom_boxplot(alpha = 0.85, outlier.size = 0.4, outlier.alpha = 0.4,
               position = position_dodge(width = 0.75), width = 0.7) +
  geom_hline(yintercept = overall_median, linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  scale_fill_manual(values = c("Short-Read Gene"   = "#7B8DC9",
                               "Long-Read Isoform" = "#7BB07B")) +
  labs(
    title    = "logFC Distribution Of NMD Susceptible Genes",
    subtitle = "lfsr < 0.05 & logFC > 0 | dashed line: overall median",
    x        = "Cell Type",
    y        = "Posterior Mean logFC (Smg1i Vs DMSO)",
    fill     = "Analysis"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(color = "grey40"),
        legend.position = "top",
        panel.grid.minor = element_blank())

print(p_logfc)

ggsave(file.path(OUTPUT_DIR, "logfc_distribution_sr_vs_lriso.png"),
       p_logfc, width = 10, height = 6, dpi = 300, bg = "white")
```

```{r logfc-summary-stats}
# ── Claim 2.1.6 — relational, and it states no number ─────────────────────────────────
# "Mean posterior logFC substantially larger at isoform than gene level." All SUPPORTING: the
# paper prints no value, so the assertion is the ORDERING, and it must hold in EVERY cell type
# rather than on average -- emitting the eight means lets a reader check that directly.
.mlogfc <- tapply(nmd_only$logFC, list(nmd_only$analysis, nmd_only$ct), mean)
POP_216 <- paste("mean mashr posterior logFC of NMD susceptible features within each cell type,",
                 "computed at each platform's OWN resolution: genes for short-read, isoforms for",
                 "long-read. Not a paired comparison -- the two sets contain different entities")
for (.a in rownames(.mlogfc)) for (.c in colnames(.mlogfc)) {
  if (!is.na(.mlogfc[.a, .c]))
    claim_emit("2.1.6", sprintf("mean posterior logFC, %s, %s", .a, .c), .mlogfc[.a, .c],
               n = sum(nmd_only$analysis == .a & nmd_only$ct == .c), population = POP_216)
}

cat("Summary statistics (NMD susceptible genes only):\n")
nmd_only %>%
  group_by(analysis, ct) %>%
  summarise(n      = n(),
            min    = round(min(logFC), 3),
            Q1     = round(quantile(logFC, 0.25), 3),
            median = round(median(logFC), 3),
            mean   = round(mean(logFC), 3),
            Q3     = round(quantile(logFC, 0.75), 3),
            max    = round(max(logFC), 3),
            .groups = "drop") %>%
  print()

```

# Shared Vs Cell-Type-Specific NMD Features

For each pipeline, count features (genes for SR, isoforms for LR) at their
native resolution — no cross-pipeline aggregation. A feature is NMD susceptible
in cell type X if its own lfsr < 0.05 and logFC > 0 in X. Sharing tier =
number of cell types where the same feature ID crosses both thresholds.

```{r build-sharing-table}
# ---- SR side: features = genes ----
sr_features <- sr_dge_all %>%
  mutate(ct     = factor(ct, levels = CT_ORDER),
         is_sig = adj.P.Val < LFSR_THRESH,
         is_nmd = is_sig & logFC > LOGFC_THRESH) %>%
  filter(is_nmd) %>%
  transmute(analysis  = "Short-Read Gene",
            ct,
            feature_id = gene_id_clean)

# ---- LR side: features = isoforms (no gene aggregation) ----
lr_features <- lr_die_all %>%
  mutate(ct     = factor(ct, levels = CT_ORDER),
         is_sig = adj.P.Val < LFSR_THRESH,
         is_nmd = is_sig & logFC > LOGFC_THRESH) %>%
  filter(is_nmd) %>%
  transmute(analysis  = "Long-Read Isoform",
            ct,
            feature_id = txid)

features_long <- bind_rows(sr_features, lr_features) %>%
  mutate(analysis = factor(analysis,
                           levels = c("Short-Read Gene", "Long-Read Isoform")))

# Per-feature sharing count (within an analysis, # CTs the feature appears in)
sharing_counts <- features_long %>%
  group_by(analysis, feature_id) %>%
  summarise(n_cts = n_distinct(ct), .groups = "drop")

# Join back so each (analysis, ct, feature_id) row gets its sharing bin
sharing_long <- features_long %>%
  inner_join(sharing_counts, by = c("analysis", "feature_id")) %>%
  mutate(sharing_bin = factor(as.character(n_cts),
                              levels = as.character(seq_len(length(CT_KEEP)))))

# Per-CT, per-tier counts
sharing_summary <- sharing_long %>%
  group_by(analysis, ct, sharing_bin) %>%
  summarise(n_features = n(), .groups = "drop") %>%
  group_by(analysis, ct) %>%
  mutate(n_total      = sum(n_features),
         n_specific   = ifelse(sharing_bin == "1", n_features, 0),
         pct_specific = round(100 * sum(n_specific) / n_total, 1)) %>%
  ungroup()

totals <- sharing_summary %>%
  distinct(analysis, ct, n_total, pct_specific)

# ── Claim 2.3.2 — per-cell-type specificity at isoform level ──────────────────────────
# DENOMINATOR IS PER CELL TYPE: of the isoforms NMD susceptible IN cell type X, the share that
# are NMD susceptible ONLY in X. Not a share of the 34,387 union.
.pc <- dplyr::distinct(dplyr::filter(sharing_summary, analysis == "Long-Read Isoform"),
                       ct, n_total, pct_specific)
POP_232 <- paste("of the isoforms NMD susceptible in a given cell type, the share NMD",
                 "susceptible in that cell type ALONE; denominator is that cell type's own",
                 "NMD susceptible set, isoform resolution, no gene aggregation")
for (.i in seq_len(nrow(.pc))) {
  .ctn <- as.character(.pc$ct[.i])
  claim_emit("2.3.2", sprintf("cell type-specific share, %s (%%)", .ctn),
             .pc$pct_specific[.i],
             published = if (.ctn == "LAE") "23.7" else NULL,
             n = .pc$n_total[.i], population = POP_232)
}
.others <- .pc$pct_specific[as.character(.pc$ct) != "LAE"]
claim_emit("2.3.2", "lowest cell type-specific share outside LAE (%)", min(.others),
           published = "0.8", n = sum(.pc$n_total), population = POP_232)
claim_emit("2.3.2", "highest cell type-specific share outside LAE (%)", max(.others),
           published = "3.4", n = sum(.pc$n_total), population = POP_232)

cat("Sharing tier definitions: sharing_bin = number of cell types a feature is",
    "NMD susceptible in (1 = cell-type-specific, 4 = all).\n")
cat("Per-CT totals and % CT-specific:\n")
print(totals)

cat("\nUnique feature counts per analysis (each feature counted once):\n")
features_long %>%
  group_by(analysis) %>%
  summarise(n_unique_features = n_distinct(feature_id), .groups = "drop") %>%
  print()
```

```{r sharing-plot, fig.width=11, fig.height=6}
# Color gradient: warm for unique (1 CT), deep for shared in all 4
fill_palette <- c("1" = "#D6604D",  # 1 CT — unique (warmest)
                  "2" = "#92C5DE",  # 2 CTs
                  "3" = "#4393C3",  # 3 CTs
                  "4" = "#2166AC")  # all 4 CTs (deepest)

p_sharing <- ggplot(sharing_summary,
                    aes(x = ct, y = n_features, fill = sharing_bin)) +
  geom_col(alpha = 0.9, width = 0.7) +
  geom_text(data = totals,
            aes(x = ct, y = n_total,
                label = paste0(comma(n_total), "\n(", pct_specific, "% specific)"),
                fill = NULL),
            vjust = -0.2, size = 3.2, lineheight = 0.9) +
  facet_wrap(~ analysis, ncol = 2, scales = "free_y") +
  scale_fill_manual(values = fill_palette,
                    name = "Shared In # Cell Types",
                    labels = c("1" = "1 (unique)",
                               "2" = "2",
                               "3" = "3",
                               "4" = "4 (all)")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.18))) +
  labs(title    = "Shared Vs Cell-Type-Specific NMD Features Per Analysis",
       subtitle = "lfsr < 0.05 & logFC > 0 | SR features = genes, LR features = isoforms (native resolution)",
       x = "Cell Type", y = "Number Of NMD Susceptible Features") +
  theme_minimal(base_size = 13) +
  theme(plot.title         = element_text(face = "bold"),
        plot.subtitle      = element_text(color = "grey40"),
        strip.text         = element_text(face = "bold", size = 12),
        legend.position    = "top",
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank())

print(p_sharing)

ggsave(file.path(OUTPUT_DIR, "sharing_sr_genes_vs_lr_isoforms.png"),
       p_sharing, width = 11, height = 6, dpi = 300, bg = "white")

```

```{r sharing-stats}
cat("Sharing-tier feature counts per analysis (each feature counted once per CT it appears in):\n")
sharing_summary %>%
  group_by(analysis, sharing_bin) %>%
  summarise(n_feat_x_ct = sum(n_features), .groups = "drop") %>%
  pivot_wider(names_from = sharing_bin, values_from = n_feat_x_ct,
              names_prefix = "tier_") %>%
  print()

cat("\nUnique-feature counts per tier (each feature_id counted once):\n")
sharing_counts %>%
  count(analysis, n_cts, name = "n_features") %>%
  pivot_wider(names_from = n_cts, values_from = n_features,
              names_prefix = "tier_") %>%
  print()
```

# Save Results

## Claims 2.1.4 and 2.1.5 — cross-platform NMD sharing

Every value in §2's two sharing sentences is emitted here. Added 2026-07-27 to close the §2
tree gate: the numbers were **derivable** from this report's objects but no shipped chunk
emitted them, so they had been recomputed by hand — which a reader of the deposit cannot do.

The gene mapping comes from the die output's own SQANTI-derived `gene_id`, retained at load
(see `read_results`). It deliberately does **not** route through `iso_to_gene`: gmap is
Ensembl-derived and most NMD isoforms are novel, so only 4,540 of 34,387 (13.2%) join, giving
2,595 genes against the published 7,105.

```{r claims-2-1-4-and-2-1-5}
lr_sc_214  <- dplyr::filter(sharing_counts, analysis == "Long-Read Isoform")
lr_lng_214 <- dplyr::filter(features_long,  analysis == "Long-Read Isoform")
n_iso_214  <- dplyr::n_distinct(lr_lng_214$feature_id)
core_214   <- sum(lr_sc_214$n_cts == length(CT_KEEP))
spec_214   <- sum(lr_sc_214$n_cts == 1)

cat(sprintf("2.1.4  NMD susceptible isoforms      : %s\n", format(n_iso_214, big.mark = ",")))
cat(sprintf("2.1.4  core across all %d cell types  : %s (%.1f%%)\n",
            length(CT_KEEP), format(core_214, big.mark = ","), 100 * core_214 / n_iso_214))
cat(sprintf("2.1.4  cell-type specific (1 CT)     : %s (%.1f%%)\n",
            format(spec_214, big.mark = ","), 100 * spec_214 / n_iso_214))
spec_ct_214 <- dplyr::filter(lr_lng_214,
                             feature_id %in% lr_sc_214$feature_id[lr_sc_214$n_cts == 1])
cat("2.1.4  cell-type-specific isoforms by CT (%):\n")
print(round(100 * prop.table(table(droplevels(spec_ct_214$ct))), 1))

nmd_214    <- dplyr::filter(lr_die_all, adj.P.Val < LFSR_THRESH, logFC > LOGFC_THRESH)
lr_genes   <- unique(sub("\\..*", "", nmd_214$gene_id))
sr_genes   <- unique(dplyr::filter(features_long, analysis == "Short-Read Gene")$feature_id)
both_genes <- intersect(sr_genes, lr_genes)
cat(sprintf("\n2.1.4  unique genes those isoforms map to : %s\n", format(length(lr_genes), big.mark = ",")))
cat(sprintf("2.1.5  unique short-read NMD genes        : %s\n", format(length(sr_genes), big.mark = ",")))
cat(sprintf("2.1.5  ...of which have an NMD isoform    : %s (%.0f%%)\n",
            format(length(both_genes), big.mark = ","), 100 * length(both_genes) / length(sr_genes)))
cat(sprintf("2.1.5  long-read-only NMD genes           : %s\n",
            format(length(setdiff(lr_genes, sr_genes)), big.mark = ",")))

# ── Claims 2.1.4 and 2.1.5 ────────────────────────────────────────────────────────────
# TWO DIFFERENT ENTITIES IN ONE PARAGRAPH: 2.1.4 counts ISOFORMS (and the genes they map to),
# 2.1.5 counts GENES on each platform. The gene mapping uses the DIE output's own SQANTI-derived
# gene_id, version-stripped -- deliberately NOT the Ensembl gmap, because most NMD isoforms are
# novel and only 4,540 of 34,387 would join, which would give 2,595 genes instead of 7,105.
POP_ISO_214 <- paste("long-read isoforms NMD susceptible (lfsr < 0.05 and posterior mean logFC",
                     "> 0) in at least one of the four primary cell types; sharing tier is the",
                     "number of cell types the same isoform crosses both thresholds in")
POP_GENE_215 <- paste("genes NMD susceptible on each platform at its OWN native resolution --",
                      "short-read gene level vs long-read isoform level rolled up to gene by the",
                      "SQANTI-derived gene_id -- with no cross-platform aggregation")
claim_emit("2.1.4", "NMD susceptible isoforms across all cell types", n_iso_214,
           published = "34,387", n = n_iso_214, population = POP_ISO_214)
claim_emit("2.1.4", "genes those NMD susceptible isoforms map to", length(lr_genes),
           published = "7,105", n = n_iso_214, population = POP_GENE_215)
claim_emit("2.1.4", "core NMD susceptible isoforms shared across all cell types", core_214,
           published = "19,803", n = n_iso_214, population = POP_ISO_214)
claim_emit("2.1.4", "core share of NMD susceptible isoforms (%)", 100 * core_214 / n_iso_214,
           published = "57.6", n = n_iso_214, population = POP_ISO_214)
claim_emit("2.1.4", "cell type-specific NMD susceptible isoforms", spec_214,
           published = "9,154", n = n_iso_214, population = POP_ISO_214)
claim_emit("2.1.4", "cell type-specific share of NMD susceptible isoforms (%)",
           100 * spec_214 / n_iso_214, published = "26.6", n = n_iso_214,
           population = POP_ISO_214)
claim_emit("2.1.4", "LAE share of the cell type-specific isoforms (%)",
           100 * prop.table(table(droplevels(spec_ct_214$ct)))[["LAE"]],
           published = "83.4", n = spec_214,
           population = paste("of the", spec_214, "isoforms NMD susceptible in exactly ONE cell",
                              "type, the share whose single cell type is LAE"))
claim_emit("2.1.5", "short-read NMD susceptible genes", length(sr_genes),
           published = "3,736", n = length(sr_genes), population = POP_GENE_215)
claim_emit("2.1.5", "short-read NMD genes that also have an NMD susceptible isoform",
           length(both_genes), published = "2,326", n = length(sr_genes),
           population = POP_GENE_215)
claim_emit("2.1.5", "share of short-read NMD genes with an NMD susceptible isoform (%)",
           100 * length(both_genes) / length(sr_genes), published = "62",
           n = length(sr_genes), population = POP_GENE_215)
claim_emit("2.1.5", "NMD susceptible genes detected only by long-read isoform analysis",
           length(setdiff(lr_genes, sr_genes)), published = "4,779", n = length(lr_genes),
           population = POP_GENE_215)
```

```{r save-results}
# logFC distributions of NMD susceptible genes (gene-level, both pipelines)
fwrite(nmd_only,
       file.path(OUTPUT_DIR, "logfc_nmd_responsive_sr_vs_lriso.csv"))

# Feature-resolution sharing breakdown (SR genes vs LR isoforms)
fwrite(sharing_summary,
       file.path(OUTPUT_DIR, "sharing_summary_sr_genes_vs_lr_isoforms.csv"))

# Per-feature sharing counts (one row per analysis × feature_id)
fwrite(sharing_counts,
       file.path(OUTPUT_DIR, "sharing_counts_per_feature.csv"))

# LR isoform-to-gene aggregation (kept as audit table for the logFC panel)
fwrite(lr_iso_gene,
       file.path(OUTPUT_DIR, "lr_iso_aggregated_to_gene.csv"))

cat("All saved to:", OUTPUT_DIR, "\n")
print(list.files(OUTPUT_DIR, pattern = "sr_genes_vs|sr_vs_lriso|lr_iso"))

```

# Session Info

```{r session-info}
sessionInfo()
```