---
title: "Short-Read Gene-Level DGE: limma-voom + mashr"
author: "Leshem, Yul"
date: "2026-03-10"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
    number_sections: true
---

Code location: `/udd/reyle/nmd_lungcells_2026/code/final/NMD_shortread_dge_fullmodel_2026.5.5.Rmd`

# Overview

This report produces gene-level differential expression results for the NMD-inhibitor
(Smg1i) vs DMSO contrast across four primary human lung cell types (AT2, LAE, FB, MV),
using short-read Salmon-quantified gene counts from the nf-core/rnaseq pipeline.

```{r setup, include=FALSE}
.nmd_root <- local({ d <- getwd(); while (!file.exists(file.path(d, "config", "paths.yml"))) {
  p <- dirname(d); if (identical(p, d)) stop("repo root not found"); d <- p }; d })
source(file.path(.nmd_root, "R", "load_config.R")); .P <- nmd_paths(.nmd_root)
source(file.path(.nmd_root, "R", "claim_emit.R"))  # Phase 2 claim-value emission
# .P$OUT is written to directly below and nothing else creates it (clean-room rerun,
# 2026-07-28). paths.yml promises "created on demand"; nmd_ensure_out is that promise.
nmd_ensure_out(.P)
knitr::opts_chunk$set(echo = TRUE)
options(bitmapType = "cairo")
```

# Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(edgeR)
  library(limma)
  library(DT)
  library(EnhancedVolcano)
  library(BiocParallel)
  library(htmltools)
  library(biomaRt)
  library(dplyr)
  library(ggrepel)
  library(gridExtra)
  library(parallel)
  library(topGO)
  library(org.Hs.eg.db)
})

# Bioconductor namespace masking guard
select <- dplyr::select
filter <- dplyr::filter
rename <- dplyr::rename
count  <- dplyr::count
slice  <- dplyr::slice

# Helper: volcano plot for a single topTable-style data frame
plot_volcano <- function(df, design_name, coef_name) {
  df <- df %>%
    mutate(
      negLogP = -log10(P.Value),
      sig = ifelse(adj.P.Val < 0.05, "Significant", "Not Sig"),
      label = ifelse(
        is.na(external_gene_name) | external_gene_name == "",
        ensembl_gene_id_version,
        external_gene_name
      )
    )
  top_hits <- df %>%
    filter(sig == "Significant") %>%
    arrange(P.Value) %>%
    slice_head(n = 10)

  ggplot(df, aes(x = logFC, y = negLogP)) +
    geom_point(aes(color = sig), alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Sig" = "gray")) +
    geom_text_repel(data = top_hits, aes(label = label), size = 3,
                    max.overlaps = Inf, box.padding = 0.3,
                    point.padding = 0.3, segment.color = "grey50") +
    labs(
      title = paste0("Volcano Plot — ", design_name, "\nCoefficient: ", coef_name),
      x = "log2 Fold Change",
      y = expression(-log[10]("P-value"))
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "bottom",
          plot.title = element_text(hjust = 0.5))
}
```

# Step 1 — Load & Harmonize Data

```{r load-data}
annot_dir        <- .P$ANNOT   # INPUT (deposited maps) -- named out_dir until 2026-07-25
OUT_DIR_FIGURES  <- file.path(.P$FIGURES, "sr")
DATE_STAMP       <- "4.28.26"
dir.create(OUT_DIR_FIGURES, recursive = TRUE, showWarnings = FALSE)

gmap <- readRDS(file.path(annot_dir, "gmap_ENSGv115_2025.08.12.rds"))

raw_r_file <- file.path(.P$DEPOSIT, "salmon_gene_counts_4ct.csv")
raw_r <- fread(raw_r_file)
# Count the ID columns actually present rather than assuming two: the deposit ships only
# gene_id, so a hardcoded -2L under-reports the sample count by one (25 for the real 26).
cat(sprintf("Loading raw count data. Dimensions: %d genes x %d sample columns.\n",
            nrow(raw_r),
            ncol(raw_r) - sum(c("gene_id", "gene_name") %in% names(raw_r))))

names(raw_r) <- gsub("\\.", "_", names(raw_r))
sample_cols       <- names(raw_r)[!names(raw_r) %in% c("gene_id", "gene_name")]
sample_cols_clean <- gsub("_clean", "", sample_cols)
sample_cols_clean <- gsub("_(x|y)$", "", sample_cols_clean)
sample_cols_clean <- gsub("^AT2", "AT", sample_cols_clean)
sample_cols_clean <- gsub("^DO",  "DO_ALI", sample_cols_clean)
names(raw_r)[!names(raw_r) %in% c("gene_id", "gene_name")] <- sample_cols_clean
cat("Cleaned count-matrix column names (stripped _clean/_x/_y, AT2->AT, DO->DO_ALI). Examples:\n  ",
    paste(head(sample_cols_clean, 5), collapse = ", "), "\n")

if (any(duplicated(names(raw_r)))) {
  dup_names <- unique(names(raw_r)[duplicated(names(raw_r))])
  n_before  <- ncol(raw_r)
  raw_r <- raw_r[, !duplicated(names(raw_r)), with = FALSE]
  cat(sprintf("FILTER: dropped %d duplicate-named column(s), keeping the FIRST of each (%s).\n  Columns before: %d / Columns now: %d. Verify these are true duplicates, not distinct reseq runs.\n",
              n_before - ncol(raw_r), paste(dup_names, collapse = ", "), n_before, ncol(raw_r)))
} else {
  cat("No duplicate column names found.\n")
}

# ---- phenotypes ----
# 4-CT-native: the deposited sheet is already the 26 manuscript samples with published
# cell-type names. The ALI split is NOT in the source sheets -- it was inferred from a
# hard-coded donor list (001V, 027U, 029T) and applied at deposit build time, so it is
# asserted here rather than re-derived. See deposit/build_4ct_pheno.R.
phe <- fread(file.path(.P$DEPOSIT, "pheno_4ct.csv"), data.table = FALSE)
phe$ct   <- phe$cell_type
phe$name <- phe$sample_name
stopifnot(
  "pheno is not the 26 manuscript samples" = nrow(phe) == 26L,
  "pheno carries a cell type outside the manuscript scope" =
    setequal(unique(phe$cell_type), c("AT2", "LAE", "FB", "MV")),
  "design is not balanced 13 DMSO / 13 SMG1i" = all(table(phe$treatment) == 13L)
)
cat(sprintf("Loaded deposited phenotype table: %d samples (%s), %s.\n", nrow(phe),
            paste(sort(unique(phe$cell_type)), collapse = ", "),
            paste(names(table(phe$treatment)), table(phe$treatment), sep = "=", collapse = " / ")))


# ---- analysis labels only: DD -> LAE, AT -> AT2 (sample NAMES keep DD/AT prefix to match counts) ----
n_lae <- sum(phe$ct == "DD"); n_at2 <- sum(phe$ct == "AT")
phe$ct[phe$ct == "DD"] <- "LAE"
phe$ct[phe$ct == "AT"] <- "AT2"
cat(sprintf("Relabeled cell types for analysis: DD -> LAE (%d samples), AT -> AT2 (%d samples).\n  Sample names keep their DD/AT prefixes to match count columns.\n",
            n_lae, n_at2))

# ---- match phenotype to count matrix ----
keep_samples <- phe$name
present <- keep_samples %in% names(raw_r)
if (!all(present)) {
  cat(sprintf("FILTER: %d phenotype sample(s) not found in count matrix, dropped: %s.\n",
              sum(!present), paste(keep_samples[!present], collapse = ", ")))
}
# `gene_name` is OPTIONAL. The deposited salmon_gene_counts_4ct.csv ships gene_id + the 26
# sample columns and no gene_name, so naming it unconditionally here stopped this file dead
# against the deposit. Nothing downstream needs it: every gene symbol in this document comes
# from gmap (`external_gene_name` / `hgnc_symbol`, attached to dge$genes below), and the very
# next chunk already anticipates its absence -- `ncol(raw_r) - ("gene_name" %in% ...)`. This
# line was the only place that did not. Selecting what is present keeps a legacy file with
# the column working and lets the deposit through unchanged (deposit less, not more).
id_cols <- intersect(c("gene_id", "gene_name"), names(raw_r))
raw_r <- raw_r[, c(id_cols, keep_samples[present]), with = FALSE]
phe   <- phe[present, ]
cat(sprintf("Matched phenotype to counts: %d of %d samples retained.\n", sum(present), length(present)))

phe$ct        <- relevel(factor(phe$ct),        ref = "LAE")
phe$treatment <- relevel(factor(phe$treatment), ref = "DMSO")
cat("New sample names (cell type + donor ID + treatment). Examples:\n  ",
    paste(head(phe$name, 5), collapse = ", "), "\n")
cat(sprintf("Final design: %d samples; cell-type reference = %s, treatment reference = %s.\n",
            nrow(phe), levels(phe$ct)[1], levels(phe$treatment)[1]))
cat("Samples per cell type x treatment:\n"); print(table(phe$ct, phe$treatment))
```

# Step 2 — Format Gene Data

```{r format-genes}
raw_r           <- as.data.frame(raw_r)
rownames(raw_r) <- raw_r$gene_id
raw_r$gene_id   <- NULL
n_samp <- ncol(raw_r) - ("gene_name" %in% colnames(raw_r))   # gene_name is an annotation col, not a sample
cat(sprintf("Formatted gene matrix (gene IDs as row names). Dimensions: %d genes x %d samples.\n",
            nrow(raw_r), n_samp))
```

# Step 3 — Build DGEList, Filter, Normalize

```{r dgelist}
common_sid <- intersect(colnames(raw_r), phe$name)
cat(sprintf("Aligning counts and phenotype: %d samples shared.\n", length(common_sid)))
raw_r <- raw_r[, common_sid]
phe   <- phe[match(common_sid, phe$name), ]

gmap <- gmap[match(rownames(raw_r), gmap$ensembl_gene_id_version), ]
stopifnot(identical(gmap$ensembl_gene_id_version, rownames(raw_r)))

dge <- DGEList(
  counts  = raw_r,
  samples = as.data.frame(phe),
  genes   = gmap[, c("ensembl_gene_id_version", "hgnc_symbol", "external_gene_name")]
)
# Outputs go to .P$OUT, never an input root. Writing regenerated objects back over the
# retained reference copies destroys the only comparison available for verification --
# it happened on 2026-07-25 and cost the published isoform DGEList and part of the
# isoform mashr set. See MORNING_NOTES.md.
saveRDS(dge, file = file.path(.P$OUT, "dge_shortread_gene_2026.3.2.rds"))
cat(sprintf("Built DGEList: %d genes x %d samples (before expression filter).\n", nrow(dge), ncol(dge)))

design_all <- model.matrix(~ ct + treatment + ct:treatment, data = dge$samples)

n_before <- nrow(dge)
keep <- filterByExpr(dge, design = design_all, min.count = 5, min.total.count = 10, min.prop = 0)
dge  <- dge[keep, , keep.lib.sizes = FALSE]

# ── Claim 2.1.2 — the gene universe ───────────────────────────────────────────────────
# "Analysis of 25,955 genes PER CELL TYPE" is ONE universe, identical for all four contrasts,
# not four different gene sets. Note there is NO DMSO-only subsetting in this report: DMSO is
# the reference LEVEL and the Smg1i samples contribute to both the filter and the fit. That
# differs from Isoform_Landscape.Rmd, which does restrict to DMSO -- the two reports are not
# interchangeable about population even where they quote similar-looking numbers.
GENE_UNIVERSE_POP <- paste("short-read Salmon gene rows surviving filterByExpr (min.count = 5,",
                           "min.total.count = 10, min.prop = 0) under the ~ct + treatment +",
                           "ct:treatment design, over all 26 samples in BOTH treatment arms;",
                           "the four primary cell types only, ALI types absent from the deposit")
claim_emit("2.1.2", "genes analysed per cell type", nrow(dge),
           published = "25,955", n = nrow(dge), population = GENE_UNIVERSE_POP)
cat(sprintf("FILTER: removed low-expression genes (filterByExpr; min.count=5, min.total.count=10).\n  Genes before: %d / Genes now: %d (%.1f%% kept).\n",
            n_before, nrow(dge), 100 * nrow(dge) / n_before))

dge <- calcNormFactors(dge, method = "TMM")
cat(sprintf("Applied TMM normalization. Norm factors: range %.3f-%.3f, median %.3f.\n",
            min(dge$samples$norm.factors), max(dge$samples$norm.factors), median(dge$samples$norm.factors)))

saveRDS(dge, file = file.path(.P$OUT, "dge_shortread_gene_filtered_2026.3.2.rds"))
cat(sprintf("Final filtered + normalized DGEList: %d samples, %d genes.\n", ncol(dge), nrow(dge)))
```

## Diagnostic PCA

```{r pca}
v <- voom(dge, design = design_all, plot = TRUE)
cat(sprintf("Applied voom (log2-CPM, %d genes x %d samples). Running PCA for QC.\n", nrow(v$E), ncol(v$E)))

pca     <- prcomp(t(v$E), scale. = TRUE)
pca_df  <- as.data.frame(pca$x[, 1:2])
var_exp <- round(100 * (pca$sdev^2 / sum(pca$sdev^2))[1:2], 1)

pca_df$treatment <- dge$samples$treatment
pca_df$ct        <- dge$samples$ct
pca_df$sample_name <- dge$samples$sample_name   # was sample_id, a column that does not exist

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = treatment, shape = ct)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("DMSO" = "orange", "Smg1i" = "purple")) +
  labs(
    title = "Pre-Adjustment PCA",
    x = paste0("PC1 (", var_exp[1], "%)"),
    y = paste0("PC2 (", var_exp[2], "%)")
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave(file.path(OUT_DIR_FIGURES, paste0("pca_plot_sr_", DATE_STAMP, ".png")),
       p_pca, width = 10, height = 6, dpi = 300, bg = "white")

print(p_pca)
```

# Step 4 — Fit Model

```{r limma-fit}
# BLOCK ON `donor`, not `sample_id`. There is no sample_id column: pheno_4ct.csv ships
# sample_name / donor / treatment / cell_type / bam, and the DGEList adds ct / name. So
# `dge$samples$sample_id` was NULL, block=NULL, and duplicateCorrelation died inside
# unwrapdups ("dim(M) <- c(spacing, ndups, ngroups, nslides)") -- this chunk could never
# have run against the deposit. `donor` is the Methods' own blocking term ("a within-donor
# blocking term") and has the right structure here: 7 donors, each paired DMSO/Smg1i.
corfit <- duplicateCorrelation(v, design_all, block = dge$samples$donor)
cat(sprintf("Estimated within-donor correlation using duplicateCorrelation (blocked on donor): consensus = %.4f.\n",
            corfit$consensus.correlation))

fit <- lmFit(v, design_all,
             block       = dge$samples$donor,          # same blocking term as corfit above
             correlation = corfit$consensus.correlation)
fit <- eBayes(fit)

plotSA(fit, main = "Residual SD vs Average Expression")

colnames(fit$design)       <- make.names(colnames(fit$design))
colnames(fit$coefficients) <- colnames(fit$design)

contrast_matrix <- makeContrasts(
  Smg1i_in_LAE = treatmentSmg1i,
  Smg1i_in_AT2 = treatmentSmg1i + ctAT2.treatmentSmg1i,
  Smg1i_in_FB  = treatmentSmg1i + ctFB.treatmentSmg1i,
  Smg1i_in_MV  = treatmentSmg1i + ctMV.treatmentSmg1i,
  levels = fit$design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
cat("Fitted limma-voom model and defined per-cell-type contrasts: ",
    paste(colnames(contrast_matrix), collapse = ", "), ".\n", sep = "")
```

# Step 5 — Per-Cell-Type Results: Tables and Volcano Plots

```{r limma-results-extract}
contrast_names <- c("Smg1i_in_AT2", "Smg1i_in_LAE", "Smg1i_in_FB", "Smg1i_in_MV")
ct_labels      <- c("AT2", "LAE", "FB", "MV")

limma_tt <- lapply(contrast_names, function(cn) {
  topTable(fit2, coef = cn, number = Inf, sort.by = "none")
})
names(limma_tt) <- ct_labels

cat(sprintf("Extracted per cell-type results (%d genes each). Significant genes (FDR < 0.05):\n",
            nrow(limma_tt[[1]])))
for (ct in ct_labels) {
  n_sig <- sum(limma_tt[[ct]]$adj.P.Val < 0.05, na.rm = TRUE)
  n_up  <- sum(limma_tt[[ct]]$adj.P.Val < 0.05 & limma_tt[[ct]]$logFC > 0, na.rm = TRUE)
  cat(sprintf("  %s: %d total, %d up\n", ct, n_sig, n_up))
}

LIMMA_OUT_DIR <- file.path(.P$OUT, "limma")
dir.create(LIMMA_OUT_DIR, recursive = TRUE, showWarnings = FALSE)

for (ct in ct_labels) {
  tt_save <- limma_tt[[ct]][order(limma_tt[[ct]]$adj.P.Val), ]
  fwrite(tt_save,
         file = file.path(LIMMA_OUT_DIR,
                          paste0("nmd_limma_dge_", tolower(ct), "_", DATE_STAMP, ".csv")))
}
cat(sprintf("Wrote %d per-cell-type limma result tables to %s.\n", length(ct_labels), LIMMA_OUT_DIR))
```

```{r limma-volcano-2x2, fig.width=10, fig.height=10}
volcano_df <- bind_rows(lapply(ct_labels, function(ct) {
  tt <- limma_tt[[ct]]
  if (!"ensembl_gene_id_version" %in% colnames(tt)) tt$ensembl_gene_id_version <- rownames(tt)
  if (!"external_gene_name" %in% colnames(tt)) {
    tt$external_gene_name <- if ("hgnc_symbol" %in% colnames(tt)) tt$hgnc_symbol else NA_character_
  }
  tt$cell_type <- ct
  tt[, c("ensembl_gene_id_version", "external_gene_name",
         "logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B", "cell_type")]
}))

volcano_df <- volcano_df %>%
  mutate(
    negLogP = -log10(P.Value),
    sig     = ifelse(adj.P.Val < 0.05, "Significant", "Not Sig"),
    label   = ifelse(is.na(external_gene_name) | external_gene_name == "",
                     ensembl_gene_id_version, external_gene_name)
  )

top_labels <- volcano_df %>%
  filter(sig == "Significant") %>%
  group_by(cell_type) %>%
  arrange(P.Value, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

cat(sprintf("Plotting volcanoes: %d total points, %d significant, labeling top 10 per cell type.\n",
            nrow(volcano_df), sum(volcano_df$sig == "Significant")))

volcano_df$cell_type <- factor(volcano_df$cell_type, levels = c("AT2", "LAE", "FB", "MV"))
top_labels$cell_type <- factor(top_labels$cell_type, levels = c("AT2", "LAE", "FB", "MV"))

p_limma_volcano <- ggplot(volcano_df, aes(x = logFC, y = negLogP)) +
  geom_point(aes(color = sig), alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c("Significant" = "red", "Not Sig" = "gray70")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.3) +
  geom_text_repel(data = top_labels, aes(label = label),
                  size = 2.8, max.overlaps = Inf, box.padding = 0.3,
                  segment.color = "grey50") +
  facet_wrap(~ cell_type, ncol = 2) +
  labs(
    title    = "Volcano Plots — Smg1i Vs DMSO",
    subtitle = "Per-cell-type contrasts; red = adj.P.Val < 0.05, top 10 NMD susceptible labeled per panel",
    x = "log2 Fold Change (Smg1i / DMSO)",
    y = expression(-log[10]("P-value"))
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        strip.text      = element_text(face = "bold", size = 13),
        plot.title      = element_text(face = "bold", hjust = 0.5),
        plot.subtitle   = element_text(color = "grey40", hjust = 0.5))

ggsave(file.path(OUT_DIR_FIGURES, paste0("volcano_2x2_sr_", DATE_STAMP, ".png")),
       p_limma_volcano, width = 10, height = 10, dpi = 300, bg = "white")

print(p_limma_volcano)
```

```{r limma-top-tables}
# Collect widgets in a tagList; return as the chunk's last value so knitr
# registers JS deps and renders all four tables in the HTML output.
cat("Building top 50 by FDR tables per cell type.\n")
limma_dt_list <- list()

for (ct in ct_labels) {
  tt <- limma_tt[[ct]]
  if (!"ensembl_gene_id_version" %in% colnames(tt)) tt$ensembl_gene_id_version <- rownames(tt)
  if (!"external_gene_name" %in% colnames(tt)) {
    tt$external_gene_name <- if ("hgnc_symbol" %in% colnames(tt)) tt$hgnc_symbol else NA_character_
  }

  tt_top <- tt %>%
    dplyr::arrange(adj.P.Val) %>%
    dplyr::select(ensembl_gene_id_version, external_gene_name,
                  logFC, AveExpr, t, P.Value, adj.P.Val, B) %>%
    head(50)

  limma_dt_list[[ct]] <- htmltools::tagList(
    htmltools::h3(paste0(ct, " — Top 50 By FDR")),
    datatable(
      tt_top,
      options   = list(pageLength = 10, scrollX = TRUE),
      caption   = paste0("Top 50 — ", ct),
      elementId = paste0("limma_dt_", ct)
    ) %>%
      formatRound(columns  = c("logFC", "AveExpr", "t", "B"), digits = 3) %>%
      formatSignif(columns = c("P.Value", "adj.P.Val"),       digits = 3)
  )
}

htmltools::tagList(limma_dt_list)
```

# Step 6 — Adaptive Shrinkage Across Cell Types

## 6.1 Setup

```{r mashr-setup}
if (!require("ashr",     quietly = TRUE)) install.packages("ashr")
if (!require("mashr",    quietly = TRUE)) BiocManager::install("mashr")
if (!require("reshape2", quietly = TRUE)) install.packages("reshape2")

library(mashr)
library(ashr)
library(reshape2)
```

## 6.2 Extract Effect Sizes and Standard Errors

```{r mashr-extract}
bhat <- fit2$coefficients
shat <- fit2$coefficients / fit2$t
n_before <- nrow(shat)

finite_rows <- apply(shat, 1, function(x) all(is.finite(x) & x > 0))
bhat <- bhat[finite_rows, ]
shat <- shat[finite_rows, ]
cat(sprintf("FILTER: kept genes with a valid (finite, positive) SE in all %d cell types.\n  Genes before: %d / Genes now: %d (dropped %d).\n",
            ncol(shat), n_before, nrow(bhat), n_before - sum(finite_rows)))

stopifnot("bhat and shat dims must match" = identical(dim(bhat), dim(shat)))
cat(sprintf("Effect-size and SE matrices ready for mashr: %d genes x %d cell types.\n",
            nrow(bhat), ncol(bhat)))
```

## 6.3 Estimate Null Correlation

```{r mashr-correlation}
mash_data_init <- mash_set_data(Bhat = bhat, Shat = shat)
m_1by1 <- mash_1by1(mash_data_init)
strong <- get_significant_results(m_1by1, thresh = 0.05)
cat(sprintf("Found %d strong signals (gene-by-gene ash, lfsr < 0.05) to learn data-driven covariance.\n",
            length(strong)))

set.seed(42)
n_random   <- min(20000, nrow(bhat))
random_idx <- sample(seq_len(nrow(bhat)), n_random)
cat(sprintf("Sampling %d random genes (seed 42) to estimate the cross-cell-type null correlation.\n", n_random))
mash_data_random <- mash_set_data(Bhat = bhat[random_idx, ],
                                   Shat = shat[random_idx, ])
Vhat <- estimate_null_correlation_simple(mash_data_random)
cat("Estimated null correlation (Vhat):\n"); print(round(Vhat, 3))

mash_data <- mash_set_data(Bhat = bhat, Shat = shat, V = Vhat)
```

## 6.4 Build Covariance Library

```{r mashr-covariances}
if (length(strong) >= 5) {
  U_pca <- cov_pca(mash_data, npc = min(5, ncol(bhat) - 1), subset = strong)
} else {
  U_pca <- list()
}
U_c   <- cov_canonical(mash_data)
U_all <- c(U_pca, U_c)
cat(sprintf("Built covariance library: %d data-driven (PCA) + %d canonical = %d total.\n",
            length(U_pca), length(U_c), length(U_all)))
```

## 6.5 Fit Model

```{r mashr-fit}
t_start <- Sys.time()
m       <- mash(mash_data, Ulist = U_all, outputlevel = 2)
cat(sprintf("Fitted mashr model in %.1f min.\n",
            as.numeric(difftime(Sys.time(), t_start, units = "mins"))))

pi_vals <- get_estimated_pi(m)
cat("Estimated mixture weights (components > 1%):\n")
print(round(sort(pi_vals[pi_vals > 0.01], decreasing = TRUE), 3))
```

## 6.6 Posterior Means and lfsr

```{r mashr-results}
pm   <- get_pm(m)
lfsr <- get_lfsr(m)
cat(sprintf("Posterior means and lfsr computed for %d genes x %d cell types.\n", nrow(pm), ncol(pm)))

cat("--- Significant Genes (lfsr < 0.05) ---\n")
n_sig_mash <- apply(lfsr, 2, function(x) sum(x < 0.05))
# 2.1.2's significant range. SIGN-AGNOSTIC: lfsr < 0.05 alone, so genes with a NEGATIVE
# posterior mean are included -- which is exactly why the 49-67% clause has a denominator this
# size. NOT the unshrunken limma counts (far smaller) and NOT the 7,349 union across cell types.
SIG_POP <- paste("genes with mashr lfsr < 0.05 in a given cell type, counted per cell type over",
                 "the filtered gene universe; no requirement on the SIGN of the effect")
claim_emit("2.1.2", "fewest significant genes in a cell type", min(n_sig_mash),
           published = "3,122", n = nrow(dge), population = SIG_POP)
claim_emit("2.1.2", "most significant genes in a cell type", max(n_sig_mash),
           published = "6,753", n = nrow(dge), population = SIG_POP)
print(n_sig_mash)

cat("\n--- Shrunken (mashr) vs Unshrunken (limma) significant counts ---\n")
for (j in seq_len(ncol(bhat))) {
  ct_name   <- colnames(bhat)[j]
  limma_res <- topTable(fit2, coef = ct_name, number = Inf, sort.by = "none")
  n_limma   <- sum(limma_res$adj.P.Val < 0.05)
  n_mash    <- n_sig_mash[ct_name]
  cat(sprintf("  %-20s: unshrunken = %5d, shrunken = %5d (ratio: %.2f)\n",
              ct_name, n_limma, n_mash, n_mash / max(n_limma, 1)))
}
```

## 6.6b Claim 2.1.2 — genes analysed, significant, and NMD susceptible

Every value in §2's sentence *"analysis of 25,955 genes per cell type identified
3,122–6,753 significant genes, of which 49–67% were NMD susceptible"* is emitted here.

Added 2026-07-27 (W49-adjacent, §2 tree gate). The numbers were always **derivable** from
this report's outputs, but no shipped chunk emitted them — they had been recomputed by hand
to match, which is not something a reader of the deposit can do. `computes_claim` was
`derivable` for exactly this reason, and a `derivable` claim still emits a gap edge: for
step 2 the question is whether a READER can rebuild the number from the shipped code, and
"I recomputed it in a scratch script" is not that.

NMD susceptible is `pm > 0 & lfsr < 0.05`, the same definition used at the sharing plot and
in the per-cell-type export below — stated here rather than re-derived, so the three cannot
drift apart.

```{r claim-2-1-2}
nmd_mask_212 <- (pm > 0) & (lfsr < 0.05)
sig_212 <- colSums(lfsr < 0.05)
nmd_212 <- colSums(nmd_mask_212)
pct_212 <- 100 * nmd_212 / sig_212

# THE DENOMINATOR IS THE PER-CELL-TYPE SIGNIFICANT SET, not the gene universe: sig_212 is
# colSums(lfsr < 0.05), so the quantity is "of the genes significant in this cell type, the
# share whose effect goes UP under SMG1i". The endpoints come from OPPOSITE ends of the
# significance ranking -- LAE has the most significant genes and the lowest up-share -- so the
# range hides that LAE is the outlier in both directions.
NMD_SHARE_POP <- paste("of the genes significant (mashr lfsr < 0.05) in a given cell type, the",
                       "share with posterior mean logFC > 0, i.e. up under SMG1i; denominator",
                       "is that cell type's significant set, NOT the filtered gene universe")
claim_emit("2.1.2", "lowest NMD susceptible share of significant genes (%)", min(pct_212),
           published = "49", n = sum(sig_212), population = NMD_SHARE_POP)
claim_emit("2.1.2", "highest NMD susceptible share of significant genes (%)", max(pct_212),
           published = "67", n = sum(sig_212), population = NMD_SHARE_POP)

cat(sprintf("Genes analysed per cell type: %s\n", format(nrow(pm), big.mark = ",")))
print(data.frame(cell_type = colnames(pm), significant = as.integer(sig_212),
                 nmd_susceptible = as.integer(nmd_212), pct_nmd = round(pct_212, 1),
                 row.names = NULL))
cat(sprintf("\nSignificant genes range : %s - %s\n",
            format(min(sig_212), big.mark = ","), format(max(sig_212), big.mark = ",")))
cat(sprintf("NMD susceptible range   : %.0f%% - %.0f%%\n", min(pct_212), max(pct_212)))
```

## 6.7 Effect Sharing Across Cell Types

```{r mashr-sharing}
n_sig_total <- length(get_significant_results(m, thresh = 0.05))
cat(sprintf("Total genes significant in >=1 cell type (lfsr < 0.05): %d.\n", n_sig_total))

if (n_sig_total >= 2) {
  sharing      <- get_pairwise_sharing(m, factor = 0.5, lfsr_thresh = 0.05)
  sharing_sign <- get_pairwise_sharing(m, factor = 0,   lfsr_thresh = 0.05)
  cat("Computed pairwise effect sharing (same sign & within 2-fold magnitude):\n")
  print(round(sharing, 2))

  # ── Claim 2.3.3 — pairwise effect sharing ─────────────────────────────────────────
  # THE METRIC IS mashr SHARING BY MAGNITUDE (factor = 0.5), not Jaccard and not sign-only.
  # For each pair the denominator is the UNION of the two cell types' lfsr < 0.05 sets and the
  # numerator is the features whose posterior-mean ratio lies in (0.5, 2) -- same sign, within
  # two-fold. THE DENOMINATOR THEREFORE DIFFERS FOR EVERY PAIR; there is no single n behind
  # this range, which is why n is omitted rather than guessed. `sharing_sign` (factor = 0) is a
  # different and larger quantity and is NOT what the paper reports; Jaccard over the same sets
  # gives yet a third answer that matches neither published range.
  # THE DIAGONAL IS EXCLUDED -- it is exactly 1 by construction, and including it would put
  # 100% at the top of every range for a reason unrelated to the data.
  .lae  <- grep("LAE", rownames(sharing), value = TRUE)[1]
  .off  <- 100 * sharing[.lae, colnames(sharing) != .lae]
  .trio <- setdiff(rownames(sharing), .lae)
  .ts   <- 100 * sharing[.trio, .trio][upper.tri(sharing[.trio, .trio])]
  SHARE_POP_G <- paste("pairwise mashr sharing by magnitude (factor = 0.5) between cell types at",
                       "GENE level; per-pair denominator is the union of the two cell types'",
                       "lfsr < 0.05 sets, so it differs for every pair")
  claim_emit("2.3.3", "lowest gene-level sharing for a LAE-involving pair (%)", min(.off),
             published = "51", population = SHARE_POP_G)
  claim_emit("2.3.3", "highest gene-level sharing for a LAE-involving pair (%)", max(.off),
             published = "70", population = SHARE_POP_G)
  claim_emit("2.3.3", "lowest gene-level sharing within the AT2/FB/MV trio (%)", min(.ts),
             published = "83", population = SHARE_POP_G)
  claim_emit("2.3.3", "highest gene-level sharing within the AT2/FB/MV trio (%)", max(.ts),
             published = "90", population = SHARE_POP_G)
} else {
  sharing      <- NULL
  sharing_sign <- NULL
  cat("Too few significant genes for sharing analysis.\n")
}
```

## 6.8 Visualizations

```{r mashr-sharing-plot, fig.width=8, fig.height=7}
if (!is.null(sharing)) {
  ct_order <- c("LAE", "AT2", "FB", "MV")

  sharing_clean <- sharing
  rownames(sharing_clean) <- gsub("Smg1i_in_", "", rownames(sharing_clean))
  colnames(sharing_clean) <- gsub("Smg1i_in_", "", colnames(sharing_clean))

  sharing_df <- melt(sharing_clean)
  colnames(sharing_df) <- c("CellType1", "CellType2", "Sharing")
  sharing_df$CellType1 <- factor(sharing_df$CellType1, levels = ct_order)
  sharing_df$CellType2 <- factor(sharing_df$CellType2, levels = ct_order)

  p_share <- ggplot(sharing_df, aes(x = CellType1, y = CellType2, fill = Sharing)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = round(Sharing, 2)),
              size = 4, fontface = "bold",
              color = ifelse(sharing_df$Sharing > 0.6, "white", "grey20")) +
    scale_fill_gradient(low = "white", high = "#185FA5",
                        limits = c(0, 1), name = "Sharing") +
    labs(title    = "Pairwise Effect Sharing",
         subtitle = "Same sign & within 2-fold magnitude (lfsr < 0.05)",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 13) +
    theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
          axis.text.y   = element_text(face = "bold"),
          panel.grid    = element_blank(),
          plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(color = "grey40", hjust = 0.5))

  ggsave(file.path(OUT_DIR_FIGURES, paste0("sharing_heatmap_sr_", DATE_STAMP, ".png")),
         p_share, width = 8, height = 7, dpi = 300, bg = "white")
  print(p_share)
}
```

```{r logfc-cor-heatmap, fig.width=8, fig.height=7}
CT_ORDER <- c("LAE", "AT2", "FB", "MV")

nmd_mask <- (pm > 0) & (lfsr < 0.05)
nmd_any  <- rowSums(nmd_mask) >= 1
cat(sprintf("FILTER: subsetting to NMD susceptible genes (posterior mean > 0 & lfsr < 0.05 in >=1 cell type).\n  Genes before: %d / NMD susceptible: %d.\n",
            nrow(pm), sum(nmd_any)))

stopifnot("Need at least 2 NMD susceptible genes" = sum(nmd_any) >= 2)

drop_col <- grep("LAE_ALI", colnames(pm), value = TRUE)
pm_nmd   <- pm[nmd_any, !colnames(pm) %in% drop_col]

cor_pearson <- cor(pm_nmd, method = "pearson")
rownames(cor_pearson) <- gsub("Smg1i_in_", "", rownames(cor_pearson))
colnames(cor_pearson) <- gsub("Smg1i_in_", "", colnames(cor_pearson))

cor_df <- melt(cor_pearson)
colnames(cor_df) <- c("CellType1", "CellType2", "Correlation")
cor_df$CellType1 <- factor(cor_df$CellType1, levels = CT_ORDER)
cor_df$CellType2 <- factor(cor_df$CellType2, levels = CT_ORDER)

p_cor <- ggplot(cor_df, aes(x = CellType1, y = CellType2, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(Correlation, 2)),
            size = 4, fontface = "bold",
            color = ifelse(cor_df$Correlation > 0.6, "white", "grey20")) +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#185FA5",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(title    = "logFC Correlation: NMD-Responsive Genes",
       subtitle = "Pearson r of posterior mean logFC across cell type pairs",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x   = element_text(angle = 45, hjust = 1, face = "bold"),
        axis.text.y   = element_text(face = "bold"),
        panel.grid    = element_blank(),
        plot.title    = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(color = "grey40", hjust = 0.5))

ggsave(file.path(OUT_DIR_FIGURES, paste0("logfc_cor_heatmap_sr_", DATE_STAMP, ".png")),
       p_cor, width = 8, height = 7, dpi = 300, bg = "white")

print(p_cor)
```

## 6.9 Save Results

```{r mashr-save}
OUT_DIR_MASH    <- file.path(.P$OUT, "mashr_gene")   # never .P$MASHR_SR -- that is an input root
DATE_STAMP_MASH <- "2026.3.10"
dir.create(OUT_DIR_MASH, recursive = TRUE, showWarnings = FALSE)

saveRDS(m, file = file.path(OUT_DIR_MASH, paste0("mashr_model_", DATE_STAMP_MASH, ".rds")))

gene_anno <- dge$genes[, c("ensembl_gene_id_version", "external_gene_name")]
gene_anno <- gene_anno[!duplicated(gene_anno$ensembl_gene_id_version), ]

pm_df <- as.data.frame(pm)
pm_df$gene_id <- rownames(pm)
pm_df <- merge(pm_df, gene_anno,
               by.x = "gene_id", by.y = "ensembl_gene_id_version", all.x = TRUE)
pm_df <- pm_df[, c("gene_id", "external_gene_name", colnames(pm))]
pm_df <- pm_df[order(pm_df$gene_id), ]
fwrite(pm_df, file.path(OUT_DIR_MASH, paste0("mashr_posterior_means_", DATE_STAMP_MASH, ".csv")))

lfsr_df <- as.data.frame(lfsr)
lfsr_df$gene_id <- rownames(lfsr)
lfsr_df <- merge(lfsr_df, gene_anno,
                 by.x = "gene_id", by.y = "ensembl_gene_id_version", all.x = TRUE)
lfsr_df <- lfsr_df[, c("gene_id", "external_gene_name", colnames(lfsr))]
lfsr_df <- lfsr_df[order(lfsr_df$gene_id), ]
fwrite(lfsr_df, file.path(OUT_DIR_MASH, paste0("mashr_lfsr_", DATE_STAMP_MASH, ".csv")))

if (!is.null(sharing)) {
  fwrite(as.data.frame(sharing),
         file.path(OUT_DIR_MASH, paste0("mashr_sharing_2fold_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
  fwrite(as.data.frame(sharing_sign),
         file.path(OUT_DIR_MASH, paste0("mashr_sharing_sign_", DATE_STAMP_MASH, ".csv")),
         row.names = TRUE)
}
fwrite(as.data.frame(Vhat),
       file.path(OUT_DIR_MASH, paste0("mashr_null_correlation_", DATE_STAMP_MASH, ".csv")),
       row.names = TRUE)

for (j in seq_len(ncol(pm))) {
  ct_name <- colnames(pm)[j]
  res <- data.frame(
    ensembl_gene_id_version = rownames(pm),
    logFC                   = pm[, j],
    adj.P.Val               = lfsr[, j],
    nmd_responsive          = pm[, j] > 0 & lfsr[, j] < 0.05,
    stringsAsFactors        = FALSE
  )
  res <- merge(res, gene_anno,
               by.x = "ensembl_gene_id_version",
               by.y = "ensembl_gene_id_version", all.x = TRUE)
  colnames(res)[colnames(res) == "external_gene_name"] <- "hgnc_symbol"
  res <- res[order(res$adj.P.Val), ]

  ct_code <- tolower(gsub("Smg1i_in_", "", ct_name))
  ct_code <- gsub("_", "", ct_code)

  fwrite(res, file.path(OUT_DIR_MASH,
                        paste0("nmd_mashr_dge_", ct_code, "_", DATE_STAMP_MASH, ".csv")))
}

cat(sprintf("Saved mashr outputs to %s: model RDS, posterior means, lfsr, sharing, null correlation,\n  and %d per-cell-type CSVs.\n",
            OUT_DIR_MASH, ncol(pm)))
```

# Session Info

```{r session-info}
sessionInfo()
```
