---
title: "NMD predictor head-to-head comparison"
subtitle: "Our deep-learning model vs NMDetective-B vs the NMDEP rule-tree baseline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 9
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(data.table)
library(knitr)
library(DT)

.nmd_root <- local({ d <- getwd(); while (!file.exists(file.path(d, "config", "paths.yml"))) {
  p <- dirname(d); if (identical(p, d)) stop("repo root not found"); d <- p }; d })
source(file.path(.nmd_root, "R", "load_config.R")); .P <- nmd_paths(.nmd_root)
HERE     <- file.path(.P$OUT, "predictor_comparison")
# 2026-07-26: the chain writes to .P$OUT/predictor_comparison, not the source dir.
SCORES   <- file.path(.P$OUT, "predictor_comparison", "per_isoform_scores_2026.7.11.tsv")
METRICS  <- file.path(.P$OUT, "predictor_comparison", "metrics_summary_2026.7.11.tsv")
FIG_PNG  <- "../../figures/SupplementalFigures/SF43_ModelComparison/figure_s_model_comparison.png"

scores  <- fread(SCORES)
metrics <- fread(METRICS)

# Head-to-head intersection restricted to the model's held-out test split
# (chr1/3/5/7; the model never saw these isoforms at training time).
hh <- scores[!is.na(nmdetective_b_score) &
             !is.na(nmdep_rule_score) &
             !is.na(our_model_prob) &
             h5_split == "test"]

fmt <- function(x) formatC(x, format = "d", big.mark = ",")
fmt2 <- function(x) formatC(x, format = "f", digits = 2)
render_dt <- function(df, caption = NULL, pageLength = 10) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE))
}

N_ISO    <- nrow(hh)
N_GENES  <- uniqueN(hh$gene_id)
N_PTCP   <- nrow(hh[subclass == "NMD+/PTC+"])
N_PTCN   <- nrow(hh[subclass == "NMD+/PTC-"])
N_CTRL   <- nrow(hh[subclass == "Control"])
```

## 1. Purpose & approach

This report benchmarks our deep-learning NMD predictor against two
published variant-level NMD prediction models — **NMDetective-B**
(Lindeboom *et al.* 2019) and the **NMDEP rule-tree baseline**
(Saadat & Fellay 2025) — on the same long-read-observed isoforms,
using a common continuous gold standard.

Three design choices distinguish this from the original NMDetective /
NMDEP performance evaluations:

1.  **Common gold standard.** All three models are scored against the
    **mashr posterior-mean log₂ fold-change of each isoform under SMG1
    inhibition** (4-CT cross-cell-type estimate), not against the
    TCGA-derived NMD-efficacy estimates the rule-based models were
    originally trained on. This is a *cross-dataset generalisation*
    test: the rule-based models are applied to a domain different
    from their training domain.

2.  **Features computed from our isoforms, not by canonical-transcript
    lookup.** For every isoform, the four NMDetective-B features
    (in-last-exon, distance-to-start, exon-length-at-stop, within-50-nt
    of last EJ) are computed directly from the long-read-observed
    transcript structure. This eliminates the canonical-transcript
    limitation of variant-level models.

3.  **Test-set evaluation.** Because our model has training-set
    isoforms in the broader cohort, all reported metrics use only the
    **chr1 / chr3 / chr5 / chr7 held-out test split** — isoforms the
    deep-learning model never saw at training time. The rule-based
    models also see these isoforms for the first time, so the
    comparison is symmetric.

Full-rank NMDetective-A (the random-forest variant) and the full NMDEP
deep-learning model were deferred from this comparison because their
trained weights / inference code are not publicly available
(see §6).

## 2. Cohort

```{r cohort-table, echo=FALSE}
tbl <- data.table(
  Subclass = c("NMD+/PTC+", "NMD+/PTC-", "Control", "**Total**"),
  Isoforms = c(N_PTCP, N_PTCN, N_CTRL, N_ISO),
  Genes    = c(
    uniqueN(hh[subclass == "NMD+/PTC+", gene_id]),
    uniqueN(hh[subclass == "NMD+/PTC-", gene_id]),
    uniqueN(hh[subclass == "Control",   gene_id]),
    N_GENES
  )
)
kable(tbl, format.args = list(big.mark = ","),
      caption = sprintf("Test-set head-to-head cohort: %s isoforms from %s unique genes.",
                        fmt(N_ISO), fmt(N_GENES)))
```

The Control comparators are gene-matched to a subset of the
NMD-susceptible comparators, which is why the unique-gene count
(`r fmt(N_GENES)`) is below the per-subclass sum.

## 3. Per-isoform feature computation

For each cohort isoform we computed the four NMDetective-B input
features from its long-read-observed transcript structure:

| Feature                       | Definition                                                                                                  |
|-------------------------------|-------------------------------------------------------------------------------------------------------------|
| **`in_last_exon`**            | Is the comparator stop codon in the last exon of the transcript?                                            |
| **`distance_to_start_nt`**    | Transcript-coordinate distance from the start codon to the stop codon (nt).                                 |
| **`exon_length_at_stop_nt`**  | Length of the exon containing the stop codon (nt).                                                          |
| **`within_50nt_of_last_ej`**  | Is the stop codon within 50 nt upstream of the last exon-exon junction (i.e., *not* a PTC by the 50-nt rule)? |

The stop codon used per subclass:

-   **NMD+/PTC+**: the **reference-AUG-projected** stop codon
    (`comp_stop_tx_pos` from `Isopair::traceReferenceAtg()`), avoiding
    the SQANTI/TD2 PTC-avoidance bias (see CLAUDE.md).
-   **NMD+/PTC−**: same reference-AUG-projected stop.
-   **Control**: the natural stop from the comparator's own
    GENCODE-annotated CDS, projected into transcript coordinates.

## 4. Scoring

Three models scored on each isoform:

### 4.1 NMDetective-B

The published 4-rule decision tree (Lindeboom *et al.* 2019, Fig. 1c),
applied to our features with the original thresholds:

-   in last exon → **escape** (leaf score 0.00)
-   distance to start \< **150** nt → escape (0.12)
-   exon length at stop \> **407** nt → escape (0.41)
-   within **50** nt of last EJ → escape (0.20)
-   otherwise → **trigger** NMD (0.65)

### 4.2 NMDEP rule-tree baseline

Same 4-rule structure with the NMDEP-re-optimised thresholds from
Table 3 of Saadat & Fellay 2025: **49** / **120** / **355** nt. Leaf
scores inherited from NMDetective-B for comparability of magnitudes.
Agreement with NMDetective-B's binary calls is 98.9% — the small
threshold deltas only shift a handful of borderline cases.

### 4.3 Our model

Per-isoform predicted NMD probability from the trained CNN +
attention model
(`peter4244/NMD_orf_model_v5_4ct`,
`best_model_atg1000_stop1000_seed42.pt`), restricted to the chr1/3/5/7
held-out test split.

## 5. Metrics

```{r metrics-table, echo=FALSE}
nm <- function(x) c(nmdetective_b = "NMDetective-B",
                    nmdep_rule    = "NMDEP rule baseline",
                    our_model     = "Our model")[x]
sub_label <- function(s) c("head-to-head:test"            = "Pooled",
                            "head-to-head:test:NMD+/PTC+" = "NMD+/PTC+",
                            "head-to-head:test:NMD+/PTC-" = "NMD+/PTC-",
                            "head-to-head:test:Control"   = "Control")[s]

m <- metrics[stratum %in% names(sub_label())]
m[, Subclass := factor(sub_label(stratum),
                        levels = c("Pooled","NMD+/PTC+","NMD+/PTC-","Control"))]
m[, Model    := factor(nm(model),
                        levels = c("NMDetective-B","NMDEP rule baseline","Our model"))]
setorder(m, Subclass, Model)

out <- m[, .(Subclass, Model, n,
              Spearman = formatC(spearman, digits = 2, format = "f"),
              Pearson  = formatC(pearson,  digits = 2, format = "f"),
              `R²`     = formatC(r2,       digits = 2, format = "f"),
              MAE      = formatC(mae,      digits = 2, format = "f"),
              RMSE     = formatC(rmse,     digits = 2, format = "f"))]
kable(out, caption = "Test-set metrics (n = 415 isoforms; chr1/3/5/7 held out).")
```

### 5.1 Pooled

All three models reach a similar pooled performance on the test set:
**Spearman ρ `r fmt2(m[Subclass == "Pooled" & Model == "NMDetective-B", spearman])`**
(NMDetective-B), **`r fmt2(m[Subclass == "Pooled" & Model == "NMDEP rule baseline", spearman])`**
(NMDEP rule baseline), **`r fmt2(m[Subclass == "Pooled" & Model == "Our model", spearman])`**
(our model). This pooled near-equality reflects the easy separation
of NMD-susceptible isoforms from non-NMD Controls, which all three
models capture.

### 5.2 Stratified by PTC subclass

Within a single PTC subclass — where the easy between-class
separation is removed — the deep-learning model substantially
outperforms both rule-based predictors:

-   **NMD+/PTC+** (n = `r N_PTCP`): our model **Spearman `r fmt2(m[Subclass == "NMD+/PTC+" & Model == "Our model", spearman])`**,
    vs. `r fmt2(m[Subclass == "NMD+/PTC+" & Model == "NMDetective-B", spearman])`
    (NMDetective-B) and `r fmt2(m[Subclass == "NMD+/PTC+" & Model == "NMDEP rule baseline", spearman])`
    (NMDEP rule baseline).

-   **NMD+/PTC−** (n = `r N_PTCN`): our model **Spearman `r fmt2(m[Subclass == "NMD+/PTC-" & Model == "Our model", spearman])`**,
    while **both rule-based models are negatively correlated** with
    the gold standard (`r fmt2(m[Subclass == "NMD+/PTC-" & Model == "NMDetective-B", spearman])`).
    This is by construction: every PTC− isoform is a real NMD
    substrate (median gold log-FC ≈ 2.3) that lacks a premature stop
    detectable by the 50-nt-rule cascade, so the rule-based models
    assign them all low scores. The deep-learning model recovers
    these substrates because it does not depend on a discrete PTC
    call.

-   **Control** (n = `r N_CTRL`): our model **Spearman `r fmt2(m[Subclass == "Control" & Model == "Our model", spearman])`**,
    vs. essentially zero for both rule-based models
    (`r fmt2(m[Subclass == "Control" & Model == "NMDetective-B", spearman])`
    and `r fmt2(m[Subclass == "Control" & Model == "NMDEP rule baseline", spearman])`).
    The rule-based models cluster all Control scores near zero
    (matching the Controls' true log-FC ≈ 0 by coincidence, not by
    discrimination), which yields a small MAE but near-zero rank
    correlation.

## 6. Figure

```{r figure, echo=FALSE, out.width="100%", fig.cap="Supplemental Figure S_x — see figures/SupplementalFigures/SF43_ModelComparison/figure_s_model_comparison_legend.md for the published legend."}
include_graphics(FIG_PNG)
```

## 7. Take-aways

1.  **Pooled performance is uninformative for this comparison.** The
    pooled Spearman of ≈0.76–0.77 across all three models would
    suggest equivalent predictive power; the within-subclass
    breakdown shows this is driven entirely by the (easy) Controls-vs-NMD-substrate
    separation.

2.  **Rule-based predictors fail on PTC-negative NMD substrates.**
    Both rule cascades classify every NMD+/PTC− isoform as "escape NMD"
    despite their being real NMD substrates by the mashr gold
    standard. The resulting Spearman ≈ −0.14 is small in magnitude
    only because the rule scores cluster on a few discrete leaves; the
    sign is the meaningful result.

3.  **The deep-learning model retains within-subclass discrimination
    across all three subclasses.** This pattern is consistent with
    the model learning continuous sequence and structural signal
    beyond the categorical 50-nt rule.

4.  **The NMDetective-B vs NMDEP rule-tree threshold difference is
    immaterial** at the scale of this cohort (98.9% binary-call
    agreement). The two rule-based predictors are effectively a
    single baseline.

## 8. Deferred models

Two model variants were not scored in this comparison because their
trained weights / inference code are not publicly available:

-   **Full-rank NMDetective-A** (Lindeboom 2019) — the random-forest
    variant. The associated figshare deposit
    ([10.6084/m9.figshare.7803398](https://doi.org/10.6084/m9.figshare.7803398))
    contains only precomputed per-nucleotide NMD scores as GTF files
    for canonical UCSC knownGene transcripts (hg19 / hg38 / mm9 / mm10),
    with no model object or inference code released.

-   **Full NMDEP MLP** (Saadat & Fellay 2025) — a 60-feature deep
    model that augments the rule tree with Orthrus pre-trained mRNA
    sequence embeddings and three internal XGBoost auxiliary
    regressors (half-life / ribosome loading / localization). No
    public code, weights, or precomputed predictions located at the
    time of writing
    ([arXiv:2502.14547](https://arxiv.org/abs/2502.14547)), and the
    published input pipeline is variant-centric (VEP-annotated
    stop-gain variants on a single canonical transcript per gene)
    rather than designed for arbitrary long-read isoform sequences.

## 9. Reproducibility

All files live under `code/nmd_predictor_comparison/`:

| File                                       | Produced by                          |
|--------------------------------------------|--------------------------------------|
| `METHODS.md`                               | manual (pre-registration)            |
| `README.md`                                | manual                               |
| `predictions_atg1000_stop1000_seed42_all.tsv` | deposited; see `explorer_run/` for how it was produced |
| `isoforms_2026.7.11.tsv`                   | `01_extract_our_isoforms.R`          |
| `nmdetective_b_scores_2026.7.11.tsv`       | `02_score_nmdetective_b.R`           |
| `nmdep_rule_baseline_scores_2026.7.11.tsv` | `03_score_nmdep_rule_baseline.R`     |
| `per_isoform_scores_2026.7.11.tsv`         | `04_compute_metrics.R`               |
| `metrics_summary_2026.7.11.tsv`            | `04_compute_metrics.R`               |
| this report (`.Rmd` and rendered `.html`)  | `Rscript -e 'rmarkdown::render(...)'` |

Figure + legend:
`figures/SupplementalFigures/SF43_ModelComparison/figure_s_model_comparison.{py,pdf,png,_legend.md}`.

```{r session-info, echo=FALSE}
sessionInfo()
```
